### 2024-11-25 index.qmd
Paper, converted, with quarto template to start with. Orig sdss template as well.

---
title: "Enhancing Data Visualization: Insights from Graphical Perception and Comparative Analysis of Charts"
format:
  pdf:
    keep-tex: true
    documentclass: article
    papersize: letter
    fontsize: 10pt
    template-partials:
      - title.tex
      - before-bib.tex
    include-in-header:
      - text: |
          \usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
          \usepackage{url}
          \usepackage{hyperref}
          \usepackage{latexsym}
          \usepackage{amsmath, amsthm, amsfonts}
          \usepackage{algorithm, algorithmic}
          \usepackage[dvipsnames]{xcolor} % colors
          \newcommand{\mt}[1]{{\textcolor{blue}{#1}}}
          \newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
    classoption: [twocolumn]
    mainfont: Times New Roman
    colorlinks: true
author:
  - name: Maksuda Atkar Toma
    affiliation:
      department: Statistics Department
      name: University of Nebraska, Lincoln
    email: mtoma2@huskers.unl.edu
    corresponding: true
  - name: Susan Vanderplas
    affiliation:
      department: Statistics Department
      name: University of Nebraska, Lincoln
    email: svanderplas2@unl.edu
    orcid: 0000-0002-3803-0972
bibliographystyle: acl
bibliography: refs.bib
filters:
   - latex-environment
commands: [mt,svp]
---

# Historical Background

Following William Playfair's introduction of the pie [@playfair1801statistical] and bar chart [@playfairCommercialPoliticalAtlas1801], there were several discussions on its use and assessment.
Von Huhn concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications.
Bar charts are often better for the majority of practical purposes, although pie charts may occasionally match or surpass them in certain situations, according to Frederick E. Croxton and Roy E. Stryker, who made this statement later in 1927.
This study will attempt to determine if pie charts may be completely disregarded in this new era or whether there are still some fields in which they can be more beneficial than bar charts.

In the era of 80’s Cleveland and McGill's work on graphical perception provides a foundational framework for understanding how humans decode visual information in graphs. In their 1984 study, they ranked perceptual tasks, identifying position along a common scale as the most accurate, while area-based judgments were deemed the least reliable. They advocated for replacing less effective visualization types, such as pie charts, with alternatives like dot charts and framed rectangle charts to enhance perceptual accuracy. Expanding on this, their 1985 paper applied these principles to scientific data visualization, introducing tasks like density and color saturation while highlighting innovations such as scatterplot smoothing and logarithmic scaling to uncover hidden data patterns. The 1986 study experimentally validated their rankings, demonstrating that position and length consistently outperformed angle, slope, and area in accuracy, with distance between objects significantly influencing judgment. Across their research, they emphasized the importance of prioritizing high-accuracy encodings, particularly for tasks requiring quantitative precision. Their recommendations included refining traditional graph designs to align with perceptual principles and improve interpretability

# Purpose and Motivation

The ability to visualize quantitative data effectively is essential in decision-making, scientific research, and education. Graphical tools must convey data with precision while engaging users to ensure optimal understanding. Yet, disagreements persist about which chart types best balance clarity, accessibility, and accuracy. Motivated by the need for improved visual communication, this research synthesizes findings from decades of studies in graphical perception, including experimental evaluations and historical insights. It aims to explore the relative merits of widely used visualizations, including bar charts, pie charts, and hybrid formats like "grables," which combine numerical tables with graphical clarity. The goal is to establish task-specific design principles, empowering practitioners and educators to build more effective visual tools.

We will try to find out the importance of understanding the historical context of graphical methods, highlighting how their development reflects broader trends in statistical communication. We want to explore how modern visualization software (e.g., Tableau, Power BI) aligns with Cleveland and McGill’s perceptual task hierarchy. Assess whether newer tools mitigate the limitations of suboptimal encodings like area and angle.  Another idea could be to investigate whether dynamic visualizations (e.g., interactive scatterplots, time-lapse animations) improve perceptual accuracy for traditionally low-ranking encodings such as area and slope.

# Problem

While charts like bar and pie graphs are staples of data communication, their utility varies widely depending on the context. Pie charts often appeal due to their simplicity and visual aesthetics but falter in scenarios requiring detailed quantitative comparisons due to their reliance on angular and area-based judgments, which humans inherently struggle to process accurately. Bar charts, on the other hand, excel in tasks requiring precision, as they leverage a common positional scale. However, they lack the visual immediacy that non-technical audiences may seek. Meanwhile, emerging formats like grables combine the precision of tables with graphical representation but are underexplored in practice. This research addresses these gaps, seeking to clarify how these visualizations perform across different cognitive and perceptual tasks. In this modern era where AI has become very available to make such kinds of charts within a minute, how do humans’ perspectives change in different fields?

# Methods

This study draws upon seminal experiments, including Cleveland and McGill’s hierarchy of graphical encodings and Eells' foundational comparisons of pie and bar charts. A detailed review of tasks—value retrieval, trend analysis, and comparisons—provides insights into the cognitive load and accuracy associated with each visualization type. Historical analysis traces the evolution of statistical graphics, from William Playfair’s pioneering work in bar and pie charts to modern hybrid visualizations. Experimental findings are organized into a taxonomy ranking encodings like position, length, area, and color in terms of effectiveness and alignment with specific tasks.

# Conclusion

This research highlights the critical importance of tailoring visualizations to specific tasks and user contexts. Bar charts, with their emphasis on positional accuracy, remain the gold standard for detailed analyses, while pie charts can serve as accessible tools for summarizing information. Grables represent a promising innovation, combining clarity with precision for complex data sets. By integrating historical perspectives with modern insights from graphical perception studies, this research provides a comprehensive roadmap for designing effective visual tools. Practitioners, educators, and researchers are encouraged to adopt these principles, ensuring their visualizations not only communicate data effectively but also foster deeper engagement and understanding.

### 2024-11-25 sdss-template.tex
Paper, converted, with quarto template to start with. Orig sdss template as well.

% File SDSS2020_SampleExtendedAbstract.tex
\documentclass[10pt]{article}
\usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
\usepackage{url}
\usepackage{latexsym}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage{algorithm, algorithmic}
\usepackage{graphicx}

\title{Symposium on Data Science and Statistics (SDSS 2023) \\
Submission and Formatting Instructions for Extended Abstracts}

\author{
  Anonymous 1 \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  {\tt email@domain} \\\And
 Anonymous 2 \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  {\tt email@domain} \\\And
  Anonymous 3 \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  {\tt email@domain} \\}

\date{}

\begin{document}
\maketitle
\begin{abstract}
This document provides a basic template for the 2-page extended abstract
and some submission guidelines. The authors' names and affiliations should not be included. \footnote{Footnotes are also in smaller font.
See Table~\ref{fontsizes} for a list of font sizes.}
Including an abstract is optional, as the document itself is an extended abstract.
If included, an abstract should be a single paragraph.
\end{abstract}

{\bf Keywords:} Optional, Yet Potentially, Helpful

\section{Introduction}

We have based this SDSS template almost entirely on the template for
the annual conference of the Association for Computational Linguistics (ACL), whose
template includes the following paragraph of credits:
\begin{quotation}
\small % Okay to use \small inside quotation environment ONLY
This document has been adapted from the instructions for earlier ACL
proceedings, including those for ACL-2012 by Maggie Li and Michael
White, those from ACL-2010 by Jing-Shing Chang and Philipp Koehn,
those for ACL-2008 by Johanna D. Moore, Simone Teufel, James Allan,
and Sadaoki Furui, those for ACL-2005 by Hwee Tou Ng and Kemal
Oflazer, those for ACL-2002 by Eugene Charniak and Dekang Lin, and
earlier ACL and EACL formats. Those versions were written by several
people, including John Chen, Henry S. Thompson and Donald
Walker. Additional elements were taken from the formatting
instructions of the {\em International Joint Conference on Artificial
  Intelligence}.
\end{quotation}

\section{Methods}

In preparing the extended abstract,
we strongly encourage using the same four section headings seen here, namely,
Introduction, Methods, Data (or Results), and Discussion (or Conclusions).
Use of subsections is at the authors' discretion; see Section~\ref{info-on-sections}.

\subsection{Font Sizes and Layout}
Extended abstracts should be no more than two pages, including references,
in 10-point Times Roman font and two-column format, and must be submitted as pdf documents.
The \LaTeX\ template automatically uses the {\tt newtx} package if available
or {\tt times} if not to select Times Roman.
Please avoid changing the font size except, if desired, in the case of an extended
quotation as demonstrated above.

Both \LaTeX\ and
MS Word templates may be downloaded by visiting the conference website:
\begin{center}
\textbf{\texttt{github.com/SDSS2020/SDSS-templates}}
\end{center}
Margins should not be altered from those used in the template files.

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rl|}
\hline \bf Type of Text & \bf Font Size & \bf Style \\ \hline
paper title & 15 pt & bold \\
author names & 12 pt & bold \\
author affiliation & 12 pt & \\
the word ``Abstract'' & 12 pt & bold \\
section titles & 12 pt & bold \\
document text & 10 pt  &\\
captions & 10 pt & \\
abstract text & 10 pt & \\
bibliography & 10 pt & \\
footnotes & 9 pt & \\
\hline
\end{tabular}
\end{center}
\caption{\label{fontsizes} Sizes and styles of fonts used.}
\end{table}

\subsection{Sections}
\label{info-on-sections}

{\bf Headings}: While it is permissible to use bold print to introduce a paragraph
as seen here, please use the \verb+\section+,
\verb+\subsection+, and \verb+\subsubsection+
commands as in this \LaTeX\ document.  If you are using MS Word, please duplicate
the numbering and font-size conventions in this document.

\subsubsection{Numbering Sections}
As shown here, all sections, sub-sections, and sub-subsections are numbered.

\subsection{Citations}

In \LaTeX, please use the
{\tt sdss2020} bibliography style, as demonstrated by this example document.
MS Word users should duplicate the citation and reference section format seen here.

Citations within the text appear in parentheses
as~\cite{Nussbaum2018} or, if the author's name appears in the text
itself, as Nussbaum~\shortcite{Nussbaum2018}.  Treat double authors as
in~\cite{Wasserstein2016}, but write as in~\cite{Brunner} when more than two
authors are involved. Collapse multiple citations as
in~\cite{LaVange2019,Wasserstein2016}.  As alluded to above, full citations should not
be used as sentence constituents.  For instance, instead of
\begin{quote}
  ``\cite{Arbuthnot1710} found the first $p$-value ...''
\end{quote}
use
\begin{quote}
``Arbuthnot~\shortcite{Arbuthnot1710} found the first $p$-value ...''.
\end{quote}
Please make references complete, presentable, and
consistent. When using Bib\TeX, protect capital letters of names and
abbreviations in titles, as in \{B\}ayesian or \{EM\} algorithm,
in your .bib file.  For an example, notice that the ``FDA'' in
the title of \cite{LaVange2019} is typed as \verb+{FDA}+ in the .bib file
or else it would be incorrectly shown as ``fda'' in the references section.

\subsection{Footnotes}

{\bf Footnotes}: In \LaTeX, use the \verb+\footnote+
command\footnote{This is how a footnote should appear.}. If using
MS Word, please separate footnotes
from the text by a line\footnote{Note the line
separating the footnotes from the text.}.

\section{Data/Results}

We encourage the selective use of tables and/or figures as appropriate.

\subsection{Graphics and Tables}

Place figures and tables in the
paper near where they are first discussed, rather than at the end, if
possible.  It is acceptable that wide illustrations
may run across both columns.  The use of color
is allowed in extended abstract graphics.

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{bayes}}
\caption{Ironically, Thomas Bayes was never asked to provide additional
simulation studies in the Results sections of his papers.}
\label{bayespic}
\end{center}
\end{figure}

\subsection{Captions}
Provide a caption below every figure and table; number each one
sequentially in the form shown in Figure~\ref{bayespic} and
Table~\ref{fontsizes}.

\section{Discussion/Conclusions}

\subsection{Appendices and Acknowledgements}

Please avoid the use of appendices and acknowledgements
as separate sections.  Any pertinent material should be included in the
body of the extended abstract rather than in an appendix.  If
acknowledgements are necessary, then they should be made using footnotes.

\subsection{Computer Code}

If you wish to provide links to accompanying code, this is allowable because
SDSS encourages best practices for reproducible research.  However, all
conclusions and exposition relevant to the refereeing process should appear in
the body of the extended abstract; furthermore, we cannot guarantee that
code will be checked or even consulted during the refereeing process.

\subsection{Refereeing Decisions}
The review is \textbf{anonymous}. Accepted papers must be presented in person at SDSS in St. Louis
from May 23 to 26, 2023.
Accepted papers' authors will be notified in plenty of time to make registration and travel
arrangements and for papers not accepted to submit a lightning abstract to the
conference if they so choose.

\bibliographystyle{sdss2020} % Please do not change the bibliography style
\bibliography{SampleReferencesForExtendedAbstract}

\end{document}

### 2024-11-25 index.qmd
references and new lines

@@
 Following William Playfair's introduction of the pie [@playfair1801statistical] and bar chart [@playfairCommercialPoliticalAtlas1801], there were several discussions on its use and assessment.
-Von Huhn concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications.
-Bar charts are often better for the majority of practical purposes, although pie charts may occasionally match or surpass them in certain situations, according to Frederick E.
-Croxton and Roy E.
-Stryker, who made this statement later in 1927.
+@vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications.
+Bar charts are often better for the majority of practical purposes, although pie charts may occasionally match or surpass them in certain situations [@croxtonBarChartsCircle1927].
 This study will attempt to determine if pie charts may be completely disregarded in this new era or whether there are still some fields in which they can be more beneficial than bar charts.
@@
 Practitioners, educators, and researchers are encouraged to adopt these principles, ensuring their visualizations not only communicate data effectively but also foster deeper engagement and understanding.
+
+# References

### 2024-11-25 index.qmd
title change

@@
 ---
-title: "Enhancing Data Visualization: Insights from Graphical Perception and Comparative Analysis of Charts"
+title: "100 years of Pies vs. Bars"
 format:

### 2024-12-01 index.qmd
starting edits

@@
 
-Following William Playfair's introduction of the pie [@playfair1801statistical] and bar chart [@playfairCommercialPoliticalAtlas1801], there were several discussions on its use and assessment.
-@vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications.
-Bar charts are often better for the majority of practical purposes, although pie charts may occasionally match or surpass them in certain situations [@croxtonBarChartsCircle1927].
-This study will attempt to determine if pie charts may be completely disregarded in this new era or whether there are still some fields in which they can be more beneficial than bar charts.
+William Playfair introduced the pie and bar charts in the early 1800s[, and by the turn of the next century, strong opinions had been publicized in textbooks and manuals, and statisticians began conducting experimental studies comparing the two.]{.svp}
+[These studies have contradictory results: ]{.svp}
+@vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications[, while @croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
+[Much later, @clevelandGraphicalPerceptionTheory1984 conducted experiments on simple graphical elements, stating that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
+Their hierarchy of feature comprehension is often assumed to be experimentally derived, but the experiments presented in the papers [@clevelandGraphicalPerceptionTheory1984;@clevelandGraphicalPerceptionGraphical1985;@clevelandExperimentGraphicalPerception1986] only covered a few elements of this hierarchy.]{.svp}
 
-In the era of 80’s Cleveland and McGill's work on graphical perception provides a foundational framework for understanding how humans decode visual information in graphs. 
-In their 1984 study, they ranked perceptual tasks, identifying position along a common scale as the most accurate, while area-based judgments were deemed the least reliable. 
-They advocated for replacing less effective visualization types, such as pie charts, with alternatives like dot charts and framed rectangle charts to enhance perceptual accuracy. 
-Expanding on this, their 1985 paper applied these principles to scientific data visualization, introducing tasks like density and color saturation while highlighting innovations such as scatterplot smoothing and logarithmic scaling to uncover hidden data patterns.
-The 1986 study experimentally validated their rankings, demonstrating that position and length consistently outperformed angle, slope, and area in accuracy, with distance between objects significantly influencing judgment.
-Across their research, they emphasized the importance of prioritizing high-accuracy encodings, particularly for tasks requiring quantitative precision. 
-Their recommendations included refining traditional graph designs to align with perceptual principles and improve interpretability
+[Here, we re-examine the historical literature surrounding the use of pie and bar charts, examining recommendations about the use of pie and bar charts, tracing these guidelines back to empirical studies, and evaluating whether the recommendations are well-supported by the studies.]{.svp}
+[We begin with an initial set of four studies from late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.]{.svp}
+[We trace these studies forward in time, examining the claims made about the studies as well as the recommendations made in citing papers.]{.svp}
+[In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results.
+Asking for absolute estimates of differences (A - B) likely gives the advantage to bar charts, while asking for part-to-whole estimates (A/(A+B)) would be expected to give the advantage to pie charts; asking for A/B ratio comparisons may produce yet a different result.]{.svp}
+[While these studies are the earliest systematic studies of statistical graphics we could locate, modern statistical graphics research might be said to date from Cleveland & McGill's work on perception of simple graphical elements; in order to include this work as well, we also look forward and backward from these papers, examining both cited papers and papers describing their work in subsequent research.]{.svp}
 

### 2024-12-01 index.qmd

@@
 
-The ability to visualize quantitative data effectively is essential in decision-making, scientific research, and education. 
-Graphical tools must convey data with precision while engaging users to ensure optimal understanding. 
+Visualizing quantitative data effectively is essential in decision-making, scientific research, and education.
+Graphical tools must convey data precisely while engaging users to ensure optimal understanding.
 Yet, disagreements persist about which chart types best balance clarity, accessibility, and accuracy.
 Motivated by the need for improved visual communication, this research synthesizes findings from decades of studies in graphical perception, including experimental evaluations and historical insights.
-It aims to explore the relative merits of widely used visualizations, including bar charts, pie charts, and hybrid formats like "grables," which combine numerical tables with graphical clarity. 
-The goal is to establish task-specific design principles, empowering practitioners and educators to build more effective visual tools.
+It aims to explore the relative merits of widely used visualizations, including bar charts, pie charts, and hybrid formats like "grables," [@hinkGrablesEnableExtraction1998] which combine numerical tables with graphical clarity.
+The goal is to establish task-specific design principles and empower practitioners and educators to build more effective visual tools.
+(**Check this paragraph, took help from AI**)
 
@@
 In this modern era where AI has become very available to make such kinds of charts within a minute, how do humans’ perspectives change in different fields?
+[**Paraphrased this section**]
 
@@
 
-# References
+This study emphasises the relevance of designing visualizations for specific tasks and user settings.
+Bar charts, which emphasize positional accuracy, remain the gold standard for extensive analysis, although pie charts can be useful tools for summarizing information.
+Grables are a potential innovation that combines clarity and precision in big data sets.
+This study provides a complete roadmap for developing effective visual aids by combining historical viewpoints with contemporary insights from graphical perception studies.
+Along with that, this article investigates the relevance of human visual perception in current tools such as Tableau and PowerBI [.mt **added this line**].
+Practitioners, educators, and researchers are advised to follow these guidelines to ensure that their visualizations effectively communicate data while also fostering deeper engagement and understanding.

### 2024-12-02 index.qmd
Summary added

--- 2024-11-25 index.qmd
@@
 commands: [mt,svp]
+editor: 
+  markdown: 
+    wrap: sentence
 ---
@@
 
-Following William Playfair's introduction of the pie [@playfair1801statistical] and bar chart [@playfairCommercialPoliticalAtlas1801], there were several discussions on its use and assessment.
-@vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications.
-Bar charts are often better for the majority of practical purposes, although pie charts may occasionally match or surpass them in certain situations [@croxtonBarChartsCircle1927].
-This study will attempt to determine if pie charts may be completely disregarded in this new era or whether there are still some fields in which they can be more beneficial than bar charts.
+William Playfair introduced the pie and bar charts in the early 1800s[, and by the turn of the next century, strong opinions had been publicized in textbooks and manuals, and statisticians began conducting experimental studies comparing the two.]{.svp}
+[These studies have contradictory results:]{.svp} @vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications[, while @croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
+[Much later, @clevelandGraphicalPerceptionTheory1984 conducted experiments on simple graphical elements, stating that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
+Their hierarchy of feature comprehension is often assumed to be experimentally derived, but the experiments presented in the papers [@clevelandGraphicalPerceptionTheory1984; @clevelandGraphicalPerceptionGraphical1985; @clevelandExperimentGraphicalPerception1986] only covered a few elements of this hierarchy.\]{.svp}
 
-In the era of 80’s Cleveland and McGill's work on graphical perception provides a foundational framework for understanding how humans decode visual information in graphs. 
-In their 1984 study, they ranked perceptual tasks, identifying position along a common scale as the most accurate, while area-based judgments were deemed the least reliable. 
-They advocated for replacing less effective visualization types, such as pie charts, with alternatives like dot charts and framed rectangle charts to enhance perceptual accuracy. 
-Expanding on this, their 1985 paper applied these principles to scientific data visualization, introducing tasks like density and color saturation while highlighting innovations such as scatterplot smoothing and logarithmic scaling to uncover hidden data patterns.
-The 1986 study experimentally validated their rankings, demonstrating that position and length consistently outperformed angle, slope, and area in accuracy, with distance between objects significantly influencing judgment.
-Across their research, they emphasized the importance of prioritizing high-accuracy encodings, particularly for tasks requiring quantitative precision. 
-Their recommendations included refining traditional graph designs to align with perceptual principles and improve interpretability
+[Here, we re-examine the historical literature surrounding the use of pie and bar charts, examining recommendations about the use of pie and bar charts, tracing these guidelines back to empirical studies, and evaluating whether the recommendations are well-supported by the studies.]{.svp}
+[We begin with an initial set of four studies from the late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.]{.svp}
+[We trace these studies forward in time, examining the claims made about the studies as well as the recommendations made in citing papers.]{.svp}
+[In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results.
+Asking for absolute estimates of differences (A - B) likely gives the advantage to bar charts, while asking for part-to-whole estimates (A/(A+B)) would be expected to give the advantage to pie charts; asking for A/B ratio comparisons may produce yet a different result.]{.svp}
+[While these studies are the earliest systematic studies of statistical graphics we could locate, modern statistical graphics research might be said to date from Cleveland & McGill's work on the perception of simple graphical elements; to include this work as well, we also look forward and backward from these papers, examining both cited papers and papers describing their work in subsequent research.]{.svp}
 
@@
 
-The ability to visualize quantitative data effectively is essential in decision-making, scientific research, and education. 
-Graphical tools must convey data with precision while engaging users to ensure optimal understanding. 
+Visualizing quantitative data effectively is essential in decision-making, scientific research, and education.
+Graphical tools must convey data precisely while engaging users to ensure optimal understanding.
 Yet, disagreements persist about which chart types best balance clarity, accessibility, and accuracy.
 Motivated by the need for improved visual communication, this research synthesizes findings from decades of studies in graphical perception, including experimental evaluations and historical insights.
-It aims to explore the relative merits of widely used visualizations, including bar charts, pie charts, and hybrid formats like "grables," which combine numerical tables with graphical clarity. 
-The goal is to establish task-specific design principles, empowering practitioners and educators to build more effective visual tools.
+It aims to explore the relative merits of widely used visualizations, including bar charts, pie charts, and hybrid formats like grables, [@hinkGrablesEnableExtraction1998] which combine numerical tables with graphical clarity.
+The goal is to establish task-specific design principles and empower practitioners and educators to build more effective visual tools.
+(**Check this paragraph, took help from AI**)
 
 We will try to find out the importance of understanding the historical context of graphical methods, highlighting how their development reflects broader trends in statistical communication.
-We want to explore how modern visualization software (e.g., Tableau, Power BI) aligns with Cleveland and McGill’s perceptual task hierarchy. 
-Assess whether newer tools mitigate the limitations of suboptimal encodings like area and angle.  
+We want to explore how modern visualization software (e.g., Tableau, Power BI) aligns with Cleveland and McGill's perceptual task hierarchy.
+Assess whether newer tools mitigate the limitations of suboptimal encodings like area and angle.\
 Another idea could be to investigate whether dynamic visualizations (e.g., interactive scatterplots, time-lapse animations) improve perceptual accuracy for traditionally low-ranking encodings such as area and slope.
@@
 This research addresses these gaps, seeking to clarify how these visualizations perform across different cognitive and perceptual tasks.
-In this modern era where AI has become very available to make such kinds of charts within a minute, how do humans’ perspectives change in different fields?
+In this modern era where AI has become very available to make such kinds of charts within a minute, how do humans' perspectives change in different fields?
+[**Paraphrased this section**]{.mt}
 
@@
 
-This study draws upon seminal experiments, including Cleveland and McGill’s hierarchy of graphical encodings and Eells' foundational comparisons of pie and bar charts.
-A detailed review of tasks—value retrieval, trend analysis, and comparisons—provides insights into the cognitive load and accuracy associated with each visualization type.
-Historical analysis traces the evolution of statistical graphics, from William Playfair’s pioneering work in bar and pie charts to modern hybrid visualizations.
+This study draws upon seminal experiments, including Cleveland and McGill's hierarchy of graphical encodings and Eells' foundational comparisons of pie and bar charts.
+A detailed review of tasks---value retrieval, trend analysis, and comparisons---provides insights into the cognitive load and accuracy associated with each visualization type.
+Historical analysis traces the evolution of statistical graphics, from William Playfair's pioneering work in bar and pie charts to modern hybrid visualizations.
 Experimental findings are organized into a taxonomy ranking encodings like position, length, area, and color in terms of effectiveness and alignment with specific tasks.
@@
 
-This research highlights the critical importance of tailoring visualizations to specific tasks and user contexts.
-Bar charts, with their emphasis on positional accuracy, remain the gold standard for detailed analyses, while pie charts can serve as accessible tools for summarizing information.
+This research highlights the importance of tailoring visualizations to specific tasks and user contexts.
+Bar charts, emphasizing positional accuracy, remain the gold standard for detailed analyses, while pie charts can serve as accessible tools for summarizing information.
 Grables represent a promising innovation, combining clarity with precision for complex data sets.
 By integrating historical perspectives with modern insights from graphical perception studies, this research provides a comprehensive roadmap for designing effective visual tools.
-Practitioners, educators, and researchers are encouraged to adopt these principles, ensuring their visualizations not only communicate data effectively but also foster deeper engagement and understanding.
+Practitioners, educators, and researchers are encouraged to adopt these principles, ensuring their visualizations communicate data effectively and foster deeper engagement and understanding.
 
-# References
+This study emphasizes the relevance of designing visualizations for specific tasks and user settings.
+Bar charts, which emphasize positional accuracy, remain the gold standard for extensive analysis, although pie charts can be useful tools for summarizing information.
+Grables are a potential innovation that combines clarity and precision in big data sets.
+This study provides a complete roadmap for developing effective visual aids by combining historical viewpoints with contemporary insights from graphical perception studies.
+
+[This article also investigates the relevance of human visual perception in current tools such as Tableau and PowerBI 
+Practitioners, educators, and researchers are advised to follow these guidelines to ensure that their visualizations effectively communicate data while fostering deeper engagement and understanding.]{.mt}
+
+# Summary from Olivia's note
+
+Eells (1926, 1927) found that pie charts (circles) were more accurate than bar charts when presenting simple distributions, such as 50-50 or 25-75 relationships.
+However, bar charts were superior for more complex distributions and comparative tasks due to their linear scale, which facilitates comparisons across multiple diagrams.
+Accuracy was influenced by the arrangement of segments (e.g., circles with sections centered at specific angles were more straightforward to interpret).
+Vohn (1927) argued that bar charts offer advantages for tasks requiring comparisons across time series or multiple distributions, owing to their standard scale and ease of labeling, while circles are more suited for showing numerous components grouped into categories.
+Frederic and Croxton (1932) compared bars, squares, circles, and cubes for simple comparisons.
+They confirmed that bar charts yielded the most accurate estimates, outperforming circles, squares, and cubes.
+Squares and circles were comparable in accuracy, while cubes consistently performed the worst due to the challenge of representing three-dimensional volume in two-dimensional drawings.
+These findings emphasize that linear encodings (e.g., bars) are more effective than area- or volume-based encodings for precise quantitative judgments.
+Overall, the studies collectively demonstrate that the choice of visualization depends on the task and the data structure.
+Bars are preferable for comparisons and time-series data, while circles may be suitable for single distributions with grouped components.
+Visual aids, such as scales and labels, further enhance accuracy and usability.
+
+# Summary from Mason's note
+
+Studies on graphical perception reveal that position is the most accurate visual encoding method, followed by length and angle, with tasks involving area, volume, curvature, shading, and color saturation needing more reliability.
+Baird (1970) demonstrated that length is perceived more accurately than area and volume, which are often underestimated.
+Baird and Noma (1978) highlighted how visual cues, such as frames, improve accuracy by providing reference structures.
+Experiments comparing position and length judgments found that position tasks were significantly more accurate, with length judgments resulting in 40–250% higher errors.
+Similarly, comparisons of position and angle judgments revealed that pie charts, which rely on angle encodings, are far less effective than bar charts, with position judgments being nearly twice as accurate and accounting for significantly fewer errors.
+Biases and outliers were common, with position accuracy declining as distances increased.
+These findings emphasize the need to avoid traditional visualizations like pie and divided bar charts, recommending alternatives such as dot charts or framed rectangular charts, which leverage position-based encodings to improve accuracy and comprehension.
+These insights underscore the importance of redesigning data visualizations to optimize perceptual accuracy and usability.

### 2024-12-02 index.tex
Summary added

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  10pt,
  letterpaper,
  twocolumn]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
\usepackage{url}
\usepackage{hyperref}
\usepackage{latexsym}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage{algorithm, algorithmic}
\usepackage[dvipsnames]{xcolor} % colors
\newcommand{\mt}[1]{{\textcolor{blue}{#1}}}
\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={100 years of Pies vs.~Bars},
  pdfauthor={Maksuda Atkar Toma; Susan Vanderplas},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{100 years of Pies vs.~Bars}
\author{
Maksuda Atkar Toma\\
Statistics Department, \\University of Nebraska, Lincoln\\
{\tt \href{mailto:mtoma2@huskers.unl.edu}{mtoma2@huskers.unl.edu}}\\
\\\And
Susan Vanderplas\\
Statistics Department, \\University of Nebraska, Lincoln\\
{\tt \href{mailto:svanderplas2@unl.edu}{svanderplas2@unl.edu}}\\
}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, breakable, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

\hypertarget{historical-background}{%
\section{Historical Background}\label{historical-background}}

William Playfair introduced the pie and bar charts in the early
1800s\svp{, and by the turn of the next century, strong opinions had
been publicized in textbooks and manuals, and statisticians began
conducting experimental studies comparing the two.} \svp{These studies
have contradictory results:} Huhn (1927) concludes that bar charts are
much more useful and efficient in the majority of situations, even
though pie charts have specific applications\svp{, while Croxton and
Stryker (1927) found that pie charts may match or surpass them in
certain situations.} \svp{Much later, Cleveland and McGill (1984)
conducted experiments on simple graphical elements, stating that aligned
length judgments are more accurate than judgments based on area or
angle, with the implication that bar charts are more accurately
perceived than pie charts.} Their hierarchy of feature comprehension is
often assumed to be experimentally derived, but the experiments
presented in the papers (Cleveland and McGill 1984, 1985, 1986) only
covered a few elements of this hierarchy.{]}\{.svp\}

\svp{Here, we re-examine the historical literature surrounding the use
of pie and bar charts, examining recommendations about the use of pie
and bar charts, tracing these guidelines back to empirical studies, and
evaluating whether the recommendations are well-supported by the
studies.} \svp{We begin with an initial set of four studies from the
late 1920s and early 1930s: Eells (1926), Croxton and Stryker (1927),
Huhn (1927), and Croxton and Stein (1932).} \svp{We trace these studies
forward in time, examining the claims made about the studies as well as
the recommendations made in citing papers.} \svp{In addition, we lay out
the different experimental designs used in each experimental study, with
the goal of assessing how the design of each study might contribute to
the ultimate results. Asking for absolute estimates of differences (A -
B) likely gives the advantage to bar charts, while asking for
part-to-whole estimates (A/(A+B)) would be expected to give the
advantage to pie charts; asking for A/B ratio comparisons may produce
yet a different result.} \svp{While these studies are the earliest
systematic studies of statistical graphics we could locate, modern
statistical graphics research might be said to date from Cleveland \&
McGill's work on the perception of simple graphical elements; to include
this work as well, we also look forward and backward from these papers,
examining both cited papers and papers describing their work in
subsequent research.}

\hypertarget{purpose-and-motivation}{%
\section{Purpose and Motivation}\label{purpose-and-motivation}}

Visualizing quantitative data effectively is essential in
decision-making, scientific research, and education. Graphical tools
must convey data precisely while engaging users to ensure optimal
understanding. Yet, disagreements persist about which chart types best
balance clarity, accessibility, and accuracy. Motivated by the need for
improved visual communication, this research synthesizes findings from
decades of studies in graphical perception, including experimental
evaluations and historical insights. It aims to explore the relative
merits of widely used visualizations, including bar charts, pie charts,
and hybrid formats like grables, (Hink, Eustace, and S. Wogalter 1998)
which combine numerical tables with graphical clarity. The goal is to
establish task-specific design principles and empower practitioners and
educators to build more effective visual tools. (\textbf{Check this
paragraph, took help from AI})

We will try to find out the importance of understanding the historical
context of graphical methods, highlighting how their development
reflects broader trends in statistical communication. We want to explore
how modern visualization software (e.g., Tableau, Power BI) aligns with
Cleveland and McGill's perceptual task hierarchy. Assess whether newer
tools mitigate the limitations of suboptimal encodings like area and
angle.\\
Another idea could be to\,investigate whether dynamic visualizations
(e.g., interactive scatterplots, time-lapse animations) improve
perceptual accuracy for traditionally low-ranking encodings such as area
and slope.

\hypertarget{problem}{%
\section{Problem}\label{problem}}

While charts like bar and pie graphs are staples of data communication,
their utility varies widely depending on the context. Pie charts often
appeal due to their simplicity and visual aesthetics but falter in
scenarios requiring detailed quantitative comparisons due to their
reliance on angular and area-based judgments, which humans inherently
struggle to process accurately. Bar charts, on the other hand, excel in
tasks requiring precision, as they leverage a common positional scale.
However, they lack the visual immediacy that non-technical audiences may
seek. Meanwhile, emerging formats like grables combine the precision of
tables with graphical representation but are underexplored in practice.
This research addresses these gaps, seeking to clarify how these
visualizations perform across different cognitive and perceptual tasks.
In this modern era where AI has become very available to make such kinds
of charts within a minute, how do humans' perspectives change in
different fields? \mt{\textbf{Paraphrased this section}}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

This study draws upon seminal experiments, including Cleveland and
McGill's hierarchy of graphical encodings and Eells' foundational
comparisons of pie and bar charts. A detailed review of tasks---value
retrieval, trend analysis, and comparisons---provides insights into the
cognitive load and accuracy associated with each visualization type.
Historical analysis traces the evolution of statistical graphics, from
William Playfair's pioneering work in bar and pie charts to modern
hybrid visualizations. Experimental findings are organized into a
taxonomy ranking encodings like position, length, area, and color in
terms of effectiveness and alignment with specific tasks.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This research highlights the importance of tailoring visualizations to
specific tasks and user contexts. Bar charts, emphasizing positional
accuracy, remain the gold standard for detailed analyses, while pie
charts can serve as accessible tools for summarizing information.
Grables represent a promising innovation, combining clarity with
precision for complex data sets. By integrating historical perspectives
with modern insights from graphical perception studies, this research
provides a comprehensive roadmap for designing effective visual tools.
Practitioners, educators, and researchers are encouraged to adopt these
principles, ensuring their visualizations communicate data effectively
and foster deeper engagement and understanding.

This study emphasizes the relevance of designing visualizations for
specific tasks and user settings. Bar charts, which emphasize positional
accuracy, remain the gold standard for extensive analysis, although pie
charts can be useful tools for summarizing information. Grables are a
potential innovation that combines clarity and precision in big data
sets. This study provides a complete roadmap for developing effective
visual aids by combining historical viewpoints with contemporary
insights from graphical perception studies.

\mt{This article also investigates the relevance of human visual
perception in current tools such as Tableau and PowerBI Practitioners,
educators, and researchers are advised to follow these guidelines to
ensure that their visualizations effectively communicate data while
fostering deeper engagement and understanding.}

\hypertarget{summary-from-olivias-note}{%
\section{Summary from Olivia's note}\label{summary-from-olivias-note}}

Eells (1926, 1927) found that pie charts (circles) were more accurate
than bar charts when presenting simple distributions, such as 50-50 or
25-75 relationships. However, bar charts were superior for more complex
distributions and comparative tasks due to their linear scale, which
facilitates comparisons across multiple diagrams. Accuracy was
influenced by the arrangement of segments (e.g., circles with sections
centered at specific angles were more straightforward to interpret).
Vohn (1927) argued that bar charts offer advantages for tasks requiring
comparisons across time series or multiple distributions, owing to their
standard scale and ease of labeling, while circles are more suited for
showing numerous components grouped into categories. Frederic and
Croxton (1932) compared bars, squares, circles, and cubes for simple
comparisons. They confirmed that bar charts yielded the most accurate
estimates, outperforming circles, squares, and cubes. Squares and
circles were comparable in accuracy, while cubes consistently performed
the worst due to the challenge of representing three-dimensional volume
in two-dimensional drawings. These findings emphasize that linear
encodings (e.g., bars) are more effective than area- or volume-based
encodings for precise quantitative judgments. Overall, the studies
collectively demonstrate that the choice of visualization depends on the
task and the data structure. Bars are preferable for comparisons and
time-series data, while circles may be suitable for single distributions
with grouped components. Visual aids, such as scales and labels, further
enhance accuracy and usability.

\hypertarget{summary-from-masons-note}{%
\section{Summary from Mason's note}\label{summary-from-masons-note}}

Studies on graphical perception reveal that position is the most
accurate visual encoding method, followed by length and angle, with
tasks involving area, volume, curvature, shading, and color saturation
needing more reliability. Baird (1970) demonstrated that length is
perceived more accurately than area and volume, which are often
underestimated. Baird and Noma (1978) highlighted how visual cues, such
as frames, improve accuracy by providing reference structures.
Experiments comparing position and length judgments found that position
tasks were significantly more accurate, with length judgments resulting
in 40--250\% higher errors. Similarly, comparisons of position and angle
judgments revealed that pie charts, which rely on angle encodings, are
far less effective than bar charts, with position judgments being nearly
twice as accurate and accounting for significantly fewer errors. Biases
and outliers were common, with position accuracy declining as distances
increased. These findings emphasize the need to avoid traditional
visualizations like pie and divided bar charts, recommending
alternatives such as dot charts or framed rectangular charts, which
leverage position-based encodings to improve accuracy and comprehension.
These insights underscore the importance of redesigning data
visualizations to optimize perceptual accuracy and usability.

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-clevelandGraphicalPerceptionTheory1984}{}}%
Cleveland, William S., and Robert McGill. 1984. {``Graphical Perception:
Theory, Experimentation, and Application to the Development of Graphical
Methods.''} \emph{Journal of the American Statistical Association} 79
(387): 531--54. \url{https://doi.org/10.1080/01621459.1984.10478080}.

\leavevmode\vadjust pre{\hypertarget{ref-clevelandGraphicalPerceptionGraphical1985}{}}%
---------. 1985. {``Graphical Perception and Graphical Methods for
Analyzing Scientific Data.''} \emph{Science} 229 (4716): 828--33.
\url{https://doi.org/10.1126/science.229.4716.828}.

\leavevmode\vadjust pre{\hypertarget{ref-clevelandExperimentGraphicalPerception1986}{}}%
---------. 1986. {``An Experiment in Graphical Perception.''}
\emph{International Journal of Man-Machine Studies} 25 (5): 491--500.
\url{https://doi.org/10.1016/S0020-7373(86)80019-0}.

\leavevmode\vadjust pre{\hypertarget{ref-croxtonGraphicComparisonsBars1932}{}}%
Croxton, Frederick E., and Harold Stein. 1932. {``Graphic Comparisons by
Bars, Squares, Circles, and Cubes.''} \emph{Journal of the American
Statistical Association} 27 (177): 54--60.

\leavevmode\vadjust pre{\hypertarget{ref-croxtonBarChartsCircle1927}{}}%
Croxton, Frederick E., and Roy E. Stryker. 1927. {``Bar Charts Versus
Circle Diagrams.''} \emph{Journal of the American Statistical
Association} 22 (160): 473--82. \url{https://doi.org/10.2307/2276829}.

\leavevmode\vadjust pre{\hypertarget{ref-eellsRelativeMeritsCircles1926}{}}%
Eells, Walter Crosby. 1926. {``The Relative Merits of Circles and Bars
for Representing Component Parts.''} \emph{Journal of the American
Statistical Association} 21 (154): 119--32.
\url{https://doi.org/10.2307/2277140}.

\leavevmode\vadjust pre{\hypertarget{ref-hinkGrablesEnableExtraction1998}{}}%
Hink, Jessica K, Jason K Eustace, and Michael S. Wogalter. 1998. {``Do
Grables Enable the Extraction of Quantitative Information Better Than
Pure Graphs or Tables?''} \emph{International Journal of Industrial
Ergonomics} 22 (6): 439--47.
\url{https://doi.org/10.1016/S0169-8141(97)00017-6}.

\leavevmode\vadjust pre{\hypertarget{ref-vonhuhnFurtherStudiesGraphic1927}{}}%
Huhn, R. von. 1927. {``Further Studies in the Graphic Use of Circles and
Bars: A Discussion of the Eell's Experiment.''} \emph{Journal of the
American Statistical Association} 22 (157): 31--39.
\url{https://doi.org/10.2307/2277346}.

\end{CSLReferences}

\bibliographystyle{sdss2020} % Please do not change the bibliography style

\end{document}

### 2024-12-03 index.qmd
Notes Integrated

@@
 [Much later, @clevelandGraphicalPerceptionTheory1984 conducted experiments on simple graphical elements, stating that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
-Their hierarchy of feature comprehension is often assumed to be experimentally derived, but the experiments presented in the papers [@clevelandGraphicalPerceptionTheory1984; @clevelandGraphicalPerceptionGraphical1985; @clevelandExperimentGraphicalPerception1986] only covered a few elements of this hierarchy.\]{.svp}
+[Their hierarchy of feature comprehension is often assumed to be experimentally derived, but the experiments presented in the papers [@clevelandGraphicalPerceptionTheory1984; @clevelandGraphicalPerceptionGraphical1985; @clevelandExperimentGraphicalPerception1986] only covered a few elements of this hierarchy.]{.svp}
 
@@
 [We trace these studies forward in time, examining the claims made about the studies as well as the recommendations made in citing papers.]{.svp}
-[In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results.
-Asking for absolute estimates of differences (A - B) likely gives the advantage to bar charts, while asking for part-to-whole estimates (A/(A+B)) would be expected to give the advantage to pie charts; asking for A/B ratio comparisons may produce yet a different result.]{.svp}
+[In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results.]{.svp}
+[They demonstrated that the choice of visualization depends on the task and the data structure.
+Bars are preferable for comparisons and time-series data, while circles may be suitable for single distributions with grouped components.]{.mt}
+[Eells (1926, 1927) found that pie charts (circles) were more accurate than bar charts when presenting simple distributions, such as 50-50 or 25-75 relationships.
+However, bar charts were superior for more complex distributions and comparative tasks due to their linear scale, which facilitates comparisons across multiple diagrams.]{.mt}
+[Vohn (1927) argued that bar charts offer advantages for tasks requiring comparisons across time series or multiple distributions, owing to their standard scale and ease of labeling, while circles are more suited for showing numerous components grouped into categories.
+Frederic and Croxton (1932) compared bars, squares, circles, and cubes for simple comparisons.
+They confirmed that bar charts yielded the most accurate estimates, outperforming circles, squares, and cubes.
+Squares and circles were comparable in accuracy, while cubes consistently performed the worst due to the challenge of representing three-dimensional volume in two-dimensional drawings.
+These findings emphasize that linear encodings (e.g., bars) are more effective than area- or volume-based encodings for precise quantitative judgments.]{.mt} [Asking for absolute estimates of differences (A - B) likely gives the advantage to bar charts, while asking for part-to-whole estimates (A/(A+B)) would be expected to give the advantage to pie charts; asking for A/B ratio comparisons may produce yet a different result.]{.svp}
+
+[In late 70's and 80's Studies on graphical perception reveal that position is the most accurate visual encoding method, followed by length and angle, with tasks involving area, volume, curvature, shading, and color saturation needing more reliability.
+Baird (1970) demonstrated that length is perceived more accurately than area and volume, which are often underestimated.
+Baird and Noma (1978) highlighted how visual cues, such as frames, improve accuracy by providing reference structures.
+Experiments comparing position and length judgments found that position tasks were significantly more accurate, with length judgments resulting in 40--250% higher errors.
+Similarly, comparisons of position and angle judgments revealed that pie charts, which rely on angle encodings, are far less effective than bar charts, with position judgments being nearly twice as accurate and accounting for significantly fewer errors.
+Biases and outliers were common, with position accuracy declining as distances increased.
+These findings emphasize the need to avoid traditional visualizations like pie and divided bar charts, recommending alternatives such as dot charts or framed rectangular charts, which leverage position-based encodings to improve accuracy and comprehension.
+These insights underscore the importance of redesigning data visualizations to optimize perceptual accuracy and usability.]{.mt}
+
 [While these studies are the earliest systematic studies of statistical graphics we could locate, modern statistical graphics research might be said to date from Cleveland & McGill's work on the perception of simple graphical elements; to include this work as well, we also look forward and backward from these papers, examining both cited papers and papers describing their work in subsequent research.]{.svp}
@@
 
-[This article also investigates the relevance of human visual perception in current tools such as Tableau and PowerBI 
-Practitioners, educators, and researchers are advised to follow these guidelines to ensure that their visualizations effectively communicate data while fostering deeper engagement and understanding.]{.mt}
-
-# Summary from Olivia's note
-
-Eells (1926, 1927) found that pie charts (circles) were more accurate than bar charts when presenting simple distributions, such as 50-50 or 25-75 relationships.
-However, bar charts were superior for more complex distributions and comparative tasks due to their linear scale, which facilitates comparisons across multiple diagrams.
-Accuracy was influenced by the arrangement of segments (e.g., circles with sections centered at specific angles were more straightforward to interpret).
-Vohn (1927) argued that bar charts offer advantages for tasks requiring comparisons across time series or multiple distributions, owing to their standard scale and ease of labeling, while circles are more suited for showing numerous components grouped into categories.
-Frederic and Croxton (1932) compared bars, squares, circles, and cubes for simple comparisons.
-They confirmed that bar charts yielded the most accurate estimates, outperforming circles, squares, and cubes.
-Squares and circles were comparable in accuracy, while cubes consistently performed the worst due to the challenge of representing three-dimensional volume in two-dimensional drawings.
-These findings emphasize that linear encodings (e.g., bars) are more effective than area- or volume-based encodings for precise quantitative judgments.
-Overall, the studies collectively demonstrate that the choice of visualization depends on the task and the data structure.
-Bars are preferable for comparisons and time-series data, while circles may be suitable for single distributions with grouped components.
-Visual aids, such as scales and labels, further enhance accuracy and usability.
-
-# Summary from Mason's note
-
-Studies on graphical perception reveal that position is the most accurate visual encoding method, followed by length and angle, with tasks involving area, volume, curvature, shading, and color saturation needing more reliability.
-Baird (1970) demonstrated that length is perceived more accurately than area and volume, which are often underestimated.
-Baird and Noma (1978) highlighted how visual cues, such as frames, improve accuracy by providing reference structures.
-Experiments comparing position and length judgments found that position tasks were significantly more accurate, with length judgments resulting in 40–250% higher errors.
-Similarly, comparisons of position and angle judgments revealed that pie charts, which rely on angle encodings, are far less effective than bar charts, with position judgments being nearly twice as accurate and accounting for significantly fewer errors.
-Biases and outliers were common, with position accuracy declining as distances increased.
-These findings emphasize the need to avoid traditional visualizations like pie and divided bar charts, recommending alternatives such as dot charts or framed rectangular charts, which leverage position-based encodings to improve accuracy and comprehension.
-These insights underscore the importance of redesigning data visualizations to optimize perceptual accuracy and usability.
+[This article also investigates the relevance of human visual perception in current tools such as Tableau and PowerBI Practitioners, educators, and researchers are advised to follow these guidelines to ensure that their visualizations effectively communicate data while fostering deeper engagement and understanding.]{.mt}

### 2024-12-03 index.tex
Notes Integrated

@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, breakable, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 angle, with the implication that bar charts are more accurately
-perceived than pie charts.} Their hierarchy of feature comprehension is
-often assumed to be experimentally derived, but the experiments
-presented in the papers (Cleveland and McGill 1984, 1985, 1986) only
-covered a few elements of this hierarchy.{]}\{.svp\}
+perceived than pie charts.} \svp{Their hierarchy of feature
+comprehension is often assumed to be experimentally derived, but the
+experiments presented in the papers (Cleveland and McGill 1984, 1985,
+1986) only covered a few elements of this hierarchy.}
 
@@
 the goal of assessing how the design of each study might contribute to
-the ultimate results. Asking for absolute estimates of differences (A -
-B) likely gives the advantage to bar charts, while asking for
-part-to-whole estimates (A/(A+B)) would be expected to give the
-advantage to pie charts; asking for A/B ratio comparisons may produce
-yet a different result.} \svp{While these studies are the earliest
-systematic studies of statistical graphics we could locate, modern
-statistical graphics research might be said to date from Cleveland \&
-McGill's work on the perception of simple graphical elements; to include
-this work as well, we also look forward and backward from these papers,
-examining both cited papers and papers describing their work in
-subsequent research.}
+the ultimate results.} \mt{They demonstrated that the choice of
+visualization depends on the task and the data structure. Bars are
+preferable for comparisons and time-series data, while circles may be
+suitable for single distributions with grouped components.} \mt{Eells
+(1926, 1927) found that pie charts (circles) were more accurate than bar
+charts when presenting simple distributions, such as 50-50 or 25-75
+relationships. However, bar charts were superior for more complex
+distributions and comparative tasks due to their linear scale, which
+facilitates comparisons across multiple diagrams.} \mt{Vohn (1927)
+argued that bar charts offer advantages for tasks requiring comparisons
+across time series or multiple distributions, owing to their standard
+scale and ease of labeling, while circles are more suited for showing
+numerous components grouped into categories. Frederic and Croxton (1932)
+compared bars, squares, circles, and cubes for simple comparisons. They
+confirmed that bar charts yielded the most accurate estimates,
+outperforming circles, squares, and cubes. Squares and circles were
+comparable in accuracy, while cubes consistently performed the worst due
+to the challenge of representing three-dimensional volume in
+two-dimensional drawings. These findings emphasize that linear encodings
+(e.g., bars) are more effective than area- or volume-based encodings for
+precise quantitative judgments.} \svp{Asking for absolute estimates of
+differences (A - B) likely gives the advantage to bar charts, while
+asking for part-to-whole estimates (A/(A+B)) would be expected to give
+the advantage to pie charts; asking for A/B ratio comparisons may
+produce yet a different result.}
+
+\mt{In late 70's and 80's Studies on graphical perception reveal that
+position is the most accurate visual encoding method, followed by length
+and angle, with tasks involving area, volume, curvature, shading, and
+color saturation needing more reliability. Baird (1970) demonstrated
+that length is perceived more accurately than area and volume, which are
+often underestimated. Baird and Noma (1978) highlighted how visual cues,
+such as frames, improve accuracy by providing reference structures.
+Experiments comparing position and length judgments found that position
+tasks were significantly more accurate, with length judgments resulting
+in 40--250\% higher errors. Similarly, comparisons of position and angle
+judgments revealed that pie charts, which rely on angle encodings, are
+far less effective than bar charts, with position judgments being nearly
+twice as accurate and accounting for significantly fewer errors. Biases
+and outliers were common, with position accuracy declining as distances
+increased. These findings emphasize the need to avoid traditional
+visualizations like pie and divided bar charts, recommending
+alternatives such as dot charts or framed rectangular charts, which
+leverage position-based encodings to improve accuracy and comprehension.
+These insights underscore the importance of redesigning data
+visualizations to optimize perceptual accuracy and usability.}
+
+\svp{While these studies are the earliest systematic studies of
+statistical graphics we could locate, modern statistical graphics
+research might be said to date from Cleveland \& McGill's work on the
+perception of simple graphical elements; to include this work as well,
+we also look forward and backward from these papers, examining both
+cited papers and papers describing their work in subsequent research.}
 
@@
 
-\hypertarget{summary-from-olivias-note}{%
-\section{Summary from Olivia's note}\label{summary-from-olivias-note}}
-
-Eells (1926, 1927) found that pie charts (circles) were more accurate
-than bar charts when presenting simple distributions, such as 50-50 or
-25-75 relationships. However, bar charts were superior for more complex
-distributions and comparative tasks due to their linear scale, which
-facilitates comparisons across multiple diagrams. Accuracy was
-influenced by the arrangement of segments (e.g., circles with sections
-centered at specific angles were more straightforward to interpret).
-Vohn (1927) argued that bar charts offer advantages for tasks requiring
-comparisons across time series or multiple distributions, owing to their
-standard scale and ease of labeling, while circles are more suited for
-showing numerous components grouped into categories. Frederic and
-Croxton (1932) compared bars, squares, circles, and cubes for simple
-comparisons. They confirmed that bar charts yielded the most accurate
-estimates, outperforming circles, squares, and cubes. Squares and
-circles were comparable in accuracy, while cubes consistently performed
-the worst due to the challenge of representing three-dimensional volume
-in two-dimensional drawings. These findings emphasize that linear
-encodings (e.g., bars) are more effective than area- or volume-based
-encodings for precise quantitative judgments. Overall, the studies
-collectively demonstrate that the choice of visualization depends on the
-task and the data structure. Bars are preferable for comparisons and
-time-series data, while circles may be suitable for single distributions
-with grouped components. Visual aids, such as scales and labels, further
-enhance accuracy and usability.
-
-\hypertarget{summary-from-masons-note}{%
-\section{Summary from Mason's note}\label{summary-from-masons-note}}
-
-Studies on graphical perception reveal that position is the most
-accurate visual encoding method, followed by length and angle, with
-tasks involving area, volume, curvature, shading, and color saturation
-needing more reliability. Baird (1970) demonstrated that length is
-perceived more accurately than area and volume, which are often
-underestimated. Baird and Noma (1978) highlighted how visual cues, such
-as frames, improve accuracy by providing reference structures.
-Experiments comparing position and length judgments found that position
-tasks were significantly more accurate, with length judgments resulting
-in 40--250\% higher errors. Similarly, comparisons of position and angle
-judgments revealed that pie charts, which rely on angle encodings, are
-far less effective than bar charts, with position judgments being nearly
-twice as accurate and accounting for significantly fewer errors. Biases
-and outliers were common, with position accuracy declining as distances
-increased. These findings emphasize the need to avoid traditional
-visualizations like pie and divided bar charts, recommending
-alternatives such as dot charts or framed rectangular charts, which
-leverage position-based encodings to improve accuracy and comprehension.
-These insights underscore the importance of redesigning data
-visualizations to optimize perceptual accuracy and usability.
-
 \hypertarget{refs}{}

### 2024-12-05 Lolipop plot.R
Lollipop chart

@@ -1,0 +1,69 @@
+
+#1. Graph Including all years
+library(ggplot2)
+
+# Data: Studies and their respective years
+study_data <- data.frame(
+  Study = c("Von Huhn (1927)", 
+            "Eells (1926)", 
+            "Croxton & Stryker (1932)", 
+            "Croxton & Eells (1927)", 
+            "Cleveland & McGill (1986)", 
+            "Cleveland & McGill (1985)", 
+            "Cleveland & McGill (1984)"),
+  Year = c(1927, 1926, 1932, 1927, 1986, 1985, 1984)
+)
+
+# Reorder the Study variable by Year in descending order
+study_data$Study <- factor(study_data$Study, levels = study_data$Study[order(study_data$Year, decreasing = TRUE)])
+
+# Create the horizontal lollipop chart
+ggplot(study_data, aes(x = Year, y = Study)) +
+  geom_segment(aes(x = min(Year), xend = Year, y = Study, yend = Study), color = "blue", size = 1) +  # Horizontal lines
+  geom_point(aes(x = Year, y = Study), color = "red", size = 4) +  # Red dots for years
+  labs(
+    title = "Initial Study Findings",
+    x = "Year",
+    y = NULL
+  ) +
+  scale_x_continuous(breaks = seq(min(study_data$Year), max(study_data$Year), 5)) +  # X-axis breaks for years
+  theme_minimal() +
+  theme(
+    axis.text.y = element_text(size = 10),  # Improve readability of study labels
+    axis.ticks.y = element_blank(),  # Hide Y-axis ticks
+    axis.title.y = element_blank(),  # Remove Y-axis label
+    axis.title.x = element_text(size = 12, face = "bold"),
+    axis.text.x = element_text(size = 10),  # Adjust X-axis label size
+    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
+  )
+
+
+#2. Graph 
+library(ggplot2)
+
+# Data: Studies and their respective years
+study_data <- data.frame(
+  Study = c( "Croxton & Stryker ", 
+             "Croxton & Eells ", 
+             "Cleveland & McGill "),
+  Year = c( 1927, 1932, 1984)
+)
+
+ggplot(study_data, aes(x = Year, y = Study)) +
+  geom_segment(aes(x = 1925, xend = Year, y = Study, yend = Study), color = "blue", size = 1) +  # Horizontal lines from left
[…]

### 2024-12-05 index.qmd
Some updates.

@@
 
-William Playfair introduced the pie and bar charts in the early 1800s[, and by the turn of the next century, strong opinions had been publicized in textbooks and manuals, and statisticians began conducting experimental studies comparing the two.]{.svp}
-[These studies have contradictory results:]{.svp} @vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications[, while @croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
+William Playfair introduced the pie and bar charts in the early 1800s[, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.]{.svp}
+[For instance, @eellsRelativeMeritsCircles1926 quotes two contemporary resources which discuss pie charts as "not a desirable form of presentation"  [@brintonGraphicMethodsPresenting1914] and more strongly, "an insult to a man's intelligence" [@karsten1923charts].]{.svp}
+[@brintonGraphicMethodsPresenting1914 even identifies that the "sector method" is common, but if horizontal bar methods were used more frequently, they would be preferred for both speed and accuracy.]{.svp}
+[@eellsRelativeMeritsCircles1926 also quotes an excerpt from @secristIntroductionStatisticalMethods1925, an introductory statistics textbook from 1925, which argues that pie charts require more cognitive effort to use accurately than a comparable stacked bar chart, calling pie diagrams "clumsy and defective" [p 184].]
+[These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use [@ichartsmarketingdepartment4ProblemsPie2017;@holtzIssuePieChart2018] and a variety of empirical studies identify problems with circular data representations [@hofmannGraphicalTestsPower2012].]{.svp}
+[@wickhamGraphicalCriticismHistorical2013 previously noted the prescience of design guidelines prior to empirical studies validating their recommendations across different types of visualizations.
+]{.svp}
+
+[Introspection is an incredibly powerful tool when applied to the design of statistical charts, and can lead to design guidelines that precede empirical tests but match their results. ]{.svp} 
+[However, empirical tests are also extremely important: Tufte's data-ink ratio is oft repeated, but empirical studies show that the relationship between minimalism and chart perception is much more complicated [@gillanMinimalismSyntaxGraphs1994a;@tufteVisualDisplayQuantitative2001;@parsonsDataVisualizationPractitioners2020].]{.svp}
+
+Interestingly, the question of pies and bars motivated some of the first empirical evaluations of charts, published in the Journal of the American Statistical Association: @eellsRelativeMeritsCircles1926, @vonhuhnFurtherStudiesGraphic1927, and @croxtonBarChartsCircle1927.
+Experimenters approached the problem head-on, directly asking participants to estimate quantities from charts, and evaluated participants on accuracy, and in some cases, speed. 
+While more sophisticated testing methodologies have been developed over the past century [see][@vanderplasTestingStatisticalCharts2020][for a review], studies are still conducted today using direct estimation approaches [@vanderplasFramedReproducingRevisiting2019].
+
+[The initial studies of the question of pies and bars have contradictory results:]{.svp} @vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications[, while @croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
 [Much later, @clevelandGraphicalPerceptionTheory1984 conducted experiments on simple graphical elements, stating that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
@@
 [In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results.]{.svp}
-[They demonstrated that the choice of visualization depends on the task and the data structure.
-Bars are preferable for comparisons and time-series data, while circles may be suitable for single distributions with grouped components.]{.mt}
+
+# Initial Studies
+<!-- XXX Who XXX?? -->
+<!-- [They demonstrated that the choice of visualization depends on the task and the data structure. -->
+<!-- Bars are preferable for comparisons and time-series data, while circles may be suitable for single distributions with grouped components.]{.mt} -->
+
 [Eells (1926, 1927) found that pie charts (circles) were more accurate than bar charts when presenting simple distributions, such as 50-50 or 25-75 relationships.

### 2024-12-05 index.qmd
rearrange

@@
 commands: [mt,svp]
-editor: 
-  markdown: 
-    wrap: sentence
 ---
@@
 
-[The initial studies of the question of pies and bars have contradictory results:]{.svp} @vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications[, while @croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
+[The initial studies of the question of pies and bars failed to identify a clear "best chart" across all situations, or even to clearly identify in which situations pie charts or bar charts should be preferred.]{.svp}
+[@eellsRelativeMeritsCircles1926 found that pie charts were read as accurately as bar charts and with comparable speed.]{.svp}
+[The simplified nature of the experiment attracted critiques from @vonhuhnFurtherStudiesGraphic1927; attached to these criticisms is a short study previously run by Croxton which examined estimates of the ratio between smaller and larger sectors of two-category pie and bar charts; these contributions will be addressed separately even though they are often cited as a single paper, even within the JASA records.]{.svp}
+@vonhuhnFurtherStudiesGraphic1927's paper provides several examples which highlight concerns with the Eells study, and argues that while pie charts may be better in certain simple situations, bar charts are much more extensible and facilitate more advanced comparisons, but does not include any experimental studies.
+(This is a common issue even today - experimental work is often criticized by those who provide toy examples but no empirical data, as if one person's experience is equivalent to another's experiment.)
+Croxton's early experimental data (attached to @vonhuhnFurtherStudiesGraphic1927) had a larger number of participants than @eellsRelativeMeritsCircles1926, but only tested two-category charts with only two proportions: .25 and .4.
+In addition, Croxton's study elicited estimates from participants as $A/B$ comparisons, where $A + B = 1$, while Eells elicited estimates for $A$ and $B$ separately.
+[It seems that prior to @eellsRelativeMeritsCircles1926, some researchers were empirically testing charts, but these experiments were not necessarily published for broader consumption; when the conversation arose in JASA, Croxton had data from years prior ready to provide additional insight.]{.svp}
+This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.
+
+[Subsequently, @croxtonBarChartsCircle1927 published additional data assessing accuracy of estimates from pie and bar charts, varying number of categories and orientation/alignment of sections.]{.svp}
+
+[@croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
 [Much later, @clevelandGraphicalPerceptionTheory1984 conducted experiments on simple graphical elements, stating that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
@@
 [In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results.]{.svp}
-
-# Initial Studies
-<!-- XXX Who XXX?? -->
-<!-- [They demonstrated that the choice of visualization depends on the task and the data structure. -->
-<!-- Bars are preferable for comparisons and time-series data, while circles may be suitable for single distributions with grouped components.]{.mt} -->
-
-[Eells (1926, 1927) found that pie charts (circles) were more accurate than bar charts when presenting simple distributions, such as 50-50 or 25-75 relationships.
-However, bar charts were superior for more complex distributions and comparative tasks due to their linear scale, which facilitates comparisons across multiple diagrams.]{.mt}
-[Vohn (1927) argued that bar charts offer advantages for tasks requiring comparisons across time series or multiple distributions, owing to their standard scale and ease of labeling, while circles are more suited for showing numerous components grouped into categories.
-Frederic and Croxton (1932) compared bars, squares, circles, and cubes for simple comparisons.
-They confirmed that bar charts yielded the most accurate estimates, outperforming circles, squares, and cubes.
-Squares and circles were comparable in accuracy, while cubes consistently performed the worst due to the challenge of representing three-dimensional volume in two-dimensional drawings.
-These findings emphasize that linear encodings (e.g., bars) are more effective than area- or volume-based encodings for precise quantitative judgments.]{.mt} [Asking for absolute estimates of differences (A - B) likely gives the advantage to bar charts, while asking for part-to-whole estimates (A/(A+B)) would be expected to give the advantage to pie charts; asking for A/B ratio comparisons may produce yet a different result.]{.svp}
-
-[In late 70's and 80's Studies on graphical perception reveal that position is the most accurate visual encoding method, followed by length and angle, with tasks involving area, volume, curvature, shading, and color saturation needing more reliability.
-Baird (1970) demonstrated that length is perceived more accurately than area and volume, which are often underestimated.
-Baird and Noma (1978) highlighted how visual cues, such as frames, improve accuracy by providing reference structures.
-Experiments comparing position and length judgments found that position tasks were significantly more accurate, with length judgments resulting in 40--250% higher errors.
-Similarly, comparisons of position and angle judgments revealed that pie charts, which rely on angle encodings, are far less effective than bar charts, with position judgments being nearly twice as accurate and accounting for significantly fewer errors.
-Biases and outliers were common, with position accuracy declining as distances increased.
-These findings emphasize the need to avoid traditional visualizations like pie and divided bar charts, recommending alternatives such as dot charts or framed rectangular charts, which leverage position-based encodings to improve accuracy and comprehension.
-These insights underscore the importance of redesigning data visualizations to optimize perceptual accuracy and usability.]{.mt}
-
-[While these studies are the earliest systematic studies of statistical graphics we could locate, modern statistical graphics research might be said to date from Cleveland & McGill's work on the perception of simple graphical elements; to include this work as well, we also look forward and backward from these papers, examining both cited papers and papers describing their work in subsequent research.]{.svp}
-
-# Purpose and Motivation
-
-Visualizing quantitative data effectively is essential in decision-making, scientific research, and education.
-Graphical tools must convey data precisely while engaging users to ensure optimal understanding.
-Yet, disagreements persist about which chart types best balance clarity, accessibility, and accuracy.
-Motivated by the need for improved visual communication, this research synthesizes findings from decades of studies in graphical perception, including experimental evaluations and historical insights.
-It aims to explore the relative merits of widely used visualizations, including bar charts, pie charts, and hybrid formats like grables, [@hinkGrablesEnableExtraction1998] which combine numerical tables with graphical clarity.
-The goal is to establish task-specific design principles and empower practitioners and educators to build more effective visual tools.
-(**Check this paragraph, took help from AI**)
-
-We will try to find out the importance of understanding the historical context of graphical methods, highlighting how their development reflects broader trends in statistical communication.
-We want to explore how modern visualization software (e.g., Tableau, Power BI) aligns with Cleveland and McGill's perceptual task hierarchy.
-Assess whether newer tools mitigate the limitations of suboptimal encodings like area and angle.\
-Another idea could be to investigate whether dynamic visualizations (e.g., interactive scatterplots, time-lapse animations) improve perceptual accuracy for traditionally low-ranking encodings such as area and slope.
-
-# Problem
-
-While charts like bar and pie graphs are staples of data communication, their utility varies widely depending on the context.
-Pie charts often appeal due to their simplicity and visual aesthetics but falter in scenarios requiring detailed quantitative comparisons due to their reliance on angular and area-based judgments, which humans inherently struggle to process accurately.
-Bar charts, on the other hand, excel in tasks requiring precision, as they leverage a common positional scale.
-However, they lack the visual immediacy that non-technical audiences may seek.
-Meanwhile, emerging formats like grables combine the precision of tables with graphical representation but are underexplored in practice.
-This research addresses these gaps, seeking to clarify how these visualizations perform across different cognitive and perceptual tasks.
-In this modern era where AI has become very available to make such kinds of charts within a minute, how do humans' perspectives change in different fields?
-[**Paraphrased this section**]{.mt}
-
-# Methods
-
-This study draws upon seminal experiments, including Cleveland and McGill's hierarchy of graphical encodings and Eells' foundational comparisons of pie and bar charts.
-A detailed review of tasks---value retrieval, trend analysis, and comparisons---provides insights into the cognitive load and accuracy associated with each visualization type.
-Historical analysis traces the evolution of statistical graphics, from William Playfair's pioneering work in bar and pie charts to modern hybrid visualizations.
-Experimental findings are organized into a taxonomy ranking encodings like position, length, area, and color in terms of effectiveness and alignment with specific tasks.
-
-# Conclusion
-
-This research highlights the importance of tailoring visualizations to specific tasks and user contexts.
-Bar charts, emphasizing positional accuracy, remain the gold standard for detailed analyses, while pie charts can serve as accessible tools for summarizing information.
-Grables represent a promising innovation, combining clarity with precision for complex data sets.
-By integrating historical perspectives with modern insights from graphical perception studies, this research provides a comprehensive roadmap for designing effective visual tools.
-Practitioners, educators, and researchers are encouraged to adopt these principles, ensuring their visualizations communicate data effectively and foster deeper engagement and understanding.
-
-This study emphasizes the relevance of designing visualizations for specific tasks and user settings.
-Bar charts, which emphasize positional accuracy, remain the gold standard for extensive analysis, although pie charts can be useful tools for summarizing information.
-Grables are a potential innovation that combines clarity and precision in big data sets.
-This study provides a complete roadmap for developing effective visual aids by combining historical viewpoints with contemporary insights from graphical perception studies.
-
-[This article also investigates the relevance of human visual perception in current tools such as Tableau and PowerBI Practitioners, educators, and researchers are advised to follow these guidelines to ensure that their visualizations effectively communicate data while fostering deeper engagement and understanding.]{.mt}

### 2024-12-05 index.qmd
update phrasing

@@
 
-# Historical Background
-
 William Playfair introduced the pie and bar charts in the early 1800s[, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.]{.svp}
@@
 [However, empirical tests are also extremely important: Tufte's data-ink ratio is oft repeated, but empirical studies show that the relationship between minimalism and chart perception is much more complicated [@gillanMinimalismSyntaxGraphs1994a;@tufteVisualDisplayQuantitative2001;@parsonsDataVisualizationPractitioners2020].]{.svp}
+
+# The Dawn of Empirical Graphics
 
@@
 [Subsequently, @croxtonBarChartsCircle1927 published additional data assessing accuracy of estimates from pie and bar charts, varying number of categories and orientation/alignment of sections.]{.svp}
+[@croxtonBarChartsCircle1927 found that pie charts matched or surpassed bar charts in most situations, across different scale alignment and in charts with up to 5 categories (though at 5 categories, differences were not statistically significant).]{.svp}
+The final paper in this early set of comparisons was @croxtonGraphicComparisonsBars1932, which examined a variety of representations of proportional data, but outside the context of a statistical chart; the experimental stimuli, while applicable to charts, show a focus on general perception, even though the paper was published in JASA.
 
-[@croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
-[Much later, @clevelandGraphicalPerceptionTheory1984 conducted experiments on simple graphical elements, stating that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
-[Their hierarchy of feature comprehension is often assumed to be experimentally derived, but the experiments presented in the papers [@clevelandGraphicalPerceptionTheory1984; @clevelandGraphicalPerceptionGraphical1985; @clevelandExperimentGraphicalPerception1986] only covered a few elements of this hierarchy.]{.svp}
+Experimental graphics studies seem to have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up in other disciplines, such as communications, during this time.
 
-[Here, we re-examine the historical literature surrounding the use of pie and bar charts, examining recommendations about the use of pie and bar charts, tracing these guidelines back to empirical studies, and evaluating whether the recommendations are well-supported by the studies.]{.svp}
-[We begin with an initial set of four studies from the late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.]{.svp}
-[We trace these studies forward in time, examining the claims made about the studies as well as the recommendations made in citing papers.]{.svp}
-[In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results.]{.svp}
+# Modern Experimental Graphics
+
+[Much later, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.]{.svp}
+[These experiments were presented as supporting a hierarchy of graphical perception which evolved over the course of the sequence of studies.]{.svp}
+[This hierarchy indicates that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
+[While the full hierarhcy of feature comprehension is often assumed to be experimentally derived, the experiments presented in the papers [@clevelandGraphicalPerceptionTheory1984; @clevelandGraphicalPerceptionGraphical1985; @clevelandExperimentGraphicalPerception1986] only inform a few elements of this hierarchy.]{.svp}
+
+Much as in the late 1920s, the combination of introspection based reasoning and a sequence of simple experiments did not go uncontested. 
+@spenceDisplayingProportionsPercentages1991 (both psychologists) used forced-choice questions of the form "which is bigger, A or B+C" to investigate the accuracy of pie and bar charts for making part-to-whole comparisons.
+The later studies show more of an interdisciplinary conversation around data displays: no longer are these charts the purview of statisticians alone. 
+As a result, the literature becomes fragmented; studies are conducted across different disciplines, with different methods.
+
+# Motivation
+
+[In this study, we re-examine the historical literature surrounding the use of pie and bar charts, including heuristics as well as empirical studies.
+We trace guidelines back to experimental studies, where possible, and examine whether justifications based on experimental studies are in fact directly supported by those studies.]{.svp}
+[We begin with an initial set of four JASA papers from the late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.]{.svp}
+[We trace these studies forward in time, examining papers which cite these studies, summaries of the original studies, and any additional information provided relevant to pie and bar charts.]{.svp}
+[In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how stimulus design, elicitation method, and other factors impact the century-old question of "which is better, pies or bars?".]{.svp}

### 2024-12-06 index.qmd
updates and draft

--- 2024-12-05 index.qmd
@@
     colorlinks: true
-author:
-  - name: Maksuda Atkar Toma
-    affiliation:
-      department: Statistics Department
-      name: University of Nebraska, Lincoln
-    email: mtoma2@huskers.unl.edu
-    corresponding: true
-  - name: Susan Vanderplas
-    affiliation:
-      department: Statistics Department
-      name: University of Nebraska, Lincoln
-    email: svanderplas2@unl.edu
-    orcid: 0000-0002-3803-0972
+# author:
+#   - name: Maksuda Atkar Toma
+#     affiliation:
+#       department: Statistics Department
+#       name: University of Nebraska, Lincoln
+#     email: mtoma2@huskers.unl.edu
+#     corresponding: true
+#   - name: Susan Vanderplas
+#     affiliation:
+#       department: Statistics Department
+#       name: University of Nebraska, Lincoln
+#     email: svanderplas2@unl.edu
+#     orcid: 0000-0002-3803-0972
 bibliographystyle: acl
@@
 
-# Historical Background
+William Playfair introduced the pie and bar charts in the early 1800s, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.
+For instance, @eellsRelativeMeritsCircles1926 quotes two contemporary resources which discuss pie charts as "not a desirable form of presentation"  [@brintonGraphicMethodsPresenting1914] and more strongly, "an insult to a man's intelligence" [@karsten1923charts].
+@brintonGraphicMethodsPresenting1914 even identifies that the "sector method" is common, but if horizontal bar methods were used more frequently, they would be preferred for both speed and accuracy.
+@eellsRelativeMeritsCircles1926 also quotes an excerpt from an introductory statistics textbook [@secristIntroductionStatisticalMethods1925], which argues that pie charts require more cognitive effort to use accurately than a comparable stacked bar chart, calling pie diagrams "clumsy and defective".
+These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use 
+<!-- [@ichartsmarketingdepartment4ProblemsPie2017;@holtzIssuePieChart2018] -->
+and a variety of empirical studies identify problems with circular data representations 
+<!-- [@hofmannGraphicalTestsPower2012] -->
+.
+@wickhamGraphicalCriticismHistorical2013 previously noted the prescience of design guidelines prior to empirical studies validating their recommendations across different types of visualizations.
 
-William Playfair introduced the pie and bar charts in the early 1800s[, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.]{.svp}
-[For instance, @eellsRelativeMeritsCircles1926 quotes two contemporary resources which discuss pie charts as "not a desirable form of presentation"  [@brintonGraphicMethodsPresenting1914] and more strongly, "an insult to a man's intelligence" [@karsten1923charts].]{.svp}
-[@brintonGraphicMethodsPresenting1914 even identifies that the "sector method" is common, but if horizontal bar methods were used more frequently, they would be preferred for both speed and accuracy.]{.svp}
-[@eellsRelativeMeritsCircles1926 also quotes an excerpt from @secristIntroductionStatisticalMethods1925, an introductory statistics textbook from 1925, which argues that pie charts require more cognitive effort to use accurately than a comparable stacked bar chart, calling pie diagrams "clumsy and defective" [p 184].]
-[These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use [@ichartsmarketingdepartment4ProblemsPie2017;@holtzIssuePieChart2018] and a variety of empirical studies identify problems with circular data representations [@hofmannGraphicalTestsPower2012].]{.svp}
-[@wickhamGraphicalCriticismHistorical2013 previously noted the prescience of design guidelines prior to empirical studies validating their recommendations across different types of visualizations.
-]{.svp}
+<!-- Introspection is an incredibly powerful tool when applied to the design of statistical charts, and can lead to design guidelines that precede empirical tests but match their results.   -->
+<!-- However, empirical tests are also extremely important: Tufte's data-ink ratio is oft repeated, but empirical studies show that the relationship between minimalism and chart perception is much more complicated [@gillanMinimalismSyntaxGraphs1994a;@tufteVisualDisplayQuantitative2001;@parsonsDataVisualizationPractitioners2020].
+-->
 
-[Introspection is an incredibly powerful tool when applied to the design of statistical charts, and can lead to design guidelines that precede empirical tests but match their results. ]{.svp} 
-[However, empirical tests are also extremely important: Tufte's data-ink ratio is oft repeated, but empirical studies show that the relationship between minimalism and chart perception is much more complicated [@gillanMinimalismSyntaxGraphs1994a;@tufteVisualDisplayQuantitative2001;@parsonsDataVisualizationPractitioners2020].]{.svp}
-
+## The Dawn of Empirical Graphics
 Interestingly, the question of pies and bars motivated some of the first empirical evaluations of charts, published in the Journal of the American Statistical Association: @eellsRelativeMeritsCircles1926, @vonhuhnFurtherStudiesGraphic1927, and @croxtonBarChartsCircle1927.
 Experimenters approached the problem head-on, directly asking participants to estimate quantities from charts, and evaluated participants on accuracy, and in some cases, speed. 
-While more sophisticated testing methodologies have been developed over the past century [see][@vanderplasTestingStatisticalCharts2020][for a review], studies are still conducted today using direct estimation approaches [@vanderplasFramedReproducingRevisiting2019].
+While more sophisticated testing methodologies have been developed over the past century [@vanderplasTestingStatisticalCharts2020], studies are still conducted today using direct estimation approaches [@vanderplasFramedReproducingRevisiting2019].
+Now as then, the phrasing of direct estimation questions is of primary importance: a simple mathematical transformation of the estimated quantity can have an outsized effect on the results.
 
-[The initial studies of the question of pies and bars failed to identify a clear "best chart" across all situations, or even to clearly identify in which situations pie charts or bar charts should be preferred.]{.svp}
-[@eellsRelativeMeritsCircles1926 found that pie charts were read as accurately as bar charts and with comparable speed.]{.svp}
-[The simplified nature of the experiment attracted critiques from @vonhuhnFurtherStudiesGraphic1927; attached to these criticisms is a short study previously run by Croxton which examined estimates of the ratio between smaller and larger sectors of two-category pie and bar charts; these contributions will be addressed separately even though they are often cited as a single paper, even within the JASA records.]{.svp}
+The initial studies of the question of pies and bars failed to identify a clear "best chart" across all situations, or even to clearly identify in which situations pie charts or bar charts should be preferred.
+@eellsRelativeMeritsCircles1926 found that pie charts were read as accurately as bar charts and with comparable speed.
+The simplified nature of the experiment attracted critiques from @vonhuhnFurtherStudiesGraphic1927; attached to these criticisms is a short study previously run by Croxton which examined estimates of the ratio between smaller and larger sectors of two-category pie and bar charts; these contributions will be addressed separately even though they are often cited as a single paper, even within the JASA records.
 @vonhuhnFurtherStudiesGraphic1927's paper provides several examples which highlight concerns with the Eells study, and argues that while pie charts may be better in certain simple situations, bar charts are much more extensible and facilitate more advanced comparisons, but does not include any experimental studies.
-(This is a common issue even today - experimental work is often criticized by those who provide toy examples but no empirical data, as if one person's experience is equivalent to another's experiment.)
+<!-- Not much has changed today - experimental work is often criticized by those who provide toy examples but no empirical data, due to compromises necessary to ensure experimental control.  -->
 Croxton's early experimental data (attached to @vonhuhnFurtherStudiesGraphic1927) had a larger number of participants than @eellsRelativeMeritsCircles1926, but only tested two-category charts with only two proportions: .25 and .4.
 In addition, Croxton's study elicited estimates from participants as $A/B$ comparisons, where $A + B = 1$, while Eells elicited estimates for $A$ and $B$ separately. 
-[It seems that prior to @eellsRelativeMeritsCircles1926, some researchers were empirically testing charts, but these experiments were not necessarily published for broader consumption; when the conversation arose in JASA, Croxton had data from years prior ready to provide additional insight.]{.svp}
+<!-- It seems that prior to @eellsRelativeMeritsCircles1926, some researchers were empirically testing charts, but these experiments were not necessarily published for broader consumption; when the conversation arose in JASA, Croxton had data from years prior ready to provide additional insight.
+-->
 This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.
 
-[Subsequently, @croxtonBarChartsCircle1927 published additional data assessing accuracy of estimates from pie and bar charts, varying number of categories and orientation/alignment of sections.]{.svp}
+Subsequently, @croxtonBarChartsCircle1927 published additional data assessing accuracy of estimates from pie and bar charts, varying number of categories and orientation/alignment of sections. 
+@croxtonBarChartsCircle1927 found that pie charts matched or surpassed bar charts in most situations, across different scale alignment and in charts with up to 5 categories (though at 5 categories, differences were not statistically significant).
+The final paper in this early set of comparisons was @croxtonGraphicComparisonsBars1932, which examined a variety of representations of proportional data, but outside the context of a statistical chart; the experimental stimuli, while applicable to charts, show a focus on general perception (though published in JASA).
 
-[@croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
-[Much later, @clevelandGraphicalPerceptionTheory1984 conducted experiments on simple graphical elements, stating that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
-[Their hierarchy of feature comprehension is often assumed to be experimentally derived, but the experiments presented in the papers [@clevelandGraphicalPerceptionTheory1984; @clevelandGraphicalPerceptionGraphical1985; @clevelandExperimentGraphicalPerception1986] only covered a few elements of this hierarchy.]{.svp}
+Experimental graphics studies seem to have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.
 
-[Here, we re-examine the historical literature surrounding the use of pie and bar charts, examining recommendations about the use of pie and bar charts, tracing these guidelines back to empirical studies, and evaluating whether the recommendations are well-supported by the studies.]{.svp}
-[We begin with an initial set of four studies from the late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.]{.svp}
-[We trace these studies forward in time, examining the claims made about the studies as well as the recommendations made in citing papers.]{.svp}
-[In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results.]{.svp}
+## Modern Experimental Graphics 
+Much later, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.
+These experiments were presented as supporting a hierarchy of graphical perception which evolved over the course of the sequence of studies.
+This ranking indicates that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.
+While the full hierarchy of feature comprehension is often assumed to be experimentally derived, the experiments presented in the series of Cleveland & McGill papers only inform a few elements of this hierarchy.
+As in the late 1920s, the combination of introspection based reasoning and a sequence of simple experiments did not go uncontested. 
+@spenceDisplayingProportionsPercentages1991 (both psychologists) used forced-choice questions of the form "which is bigger, A or B+C" to investigate the accuracy of pie and bar charts for making part-to-whole comparisons.
+The later studies show more of an interdisciplinary conversation around data displays, which have become a general tool not limited to statisticians. 
+As a result, the literature becomes fragmented; studies are conducted across different disciplines, with different methods, elicitation questions, protocols, and ultimately, different conclusions.
+
+![Different elicitation methods used to assess accuracy of pie chart estimates.](pie-type-example.png){#fig-elicitation}
+
+## Scope and Aims
+
+In this study, we re-examine the historical literature surrounding the use of pie and bar charts, including heuristics as well as empirical studies.
+We trace guidelines back to experimental studies, where possible, and examine whether justifications based on experimental studies are in fact directly supported by those studies.
+We begin with an initial set of four JASA papers from the late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.
+We trace these studies forward in time, examining papers which cite these studies, summaries of the original studies, and any additional information provided relevant to pie and bar charts.
+In addition, we specifically assess the different designs used in each experimental study, examining how the type of elicitation method impacts the results, as shown in @fig-elicitation.
+Synthesizing these results, we use the fundamental question of "which is better, pies or bars?" to examine the history of experimental graphics and graphical design guidelines.]{.svp}
+
+# References

### 2024-12-06 index.qmd
fixing issues with formatting

@@
 The final paper in this early set of comparisons was @croxtonGraphicComparisonsBars1932, which examined a variety of representations of proportional data, but outside the context of a statistical chart; the experimental stimuli, while applicable to charts, show a focus on general perception (though published in JASA).
-
 Experimental graphics studies seem to have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.
@@
 In addition, we specifically assess the different designs used in each experimental study, examining how the type of elicitation method impacts the results, as shown in @fig-elicitation.
-Synthesizing these results, we use the fundamental question of "which is better, pies or bars?" to examine the history of experimental graphics and graphical design guidelines.]{.svp}
+Synthesizing these results, we use the fundamental question of "which is better, pies or bars?" to examine the history of experimental graphics and graphical design guidelines.
 

### 2025-02-11 sdss-abstract/index.qmd
Restructure w/ folders

--- 2024-12-06 index.qmd
@@
 # author:
-#   - name: Maksuda Atkar Toma
+#   - name: Maksuda Aktar Toma
 #     affiliation:
@@
 William Playfair introduced the pie and bar charts in the early 1800s, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.
-For instance, @eellsRelativeMeritsCircles1926 quotes two contemporary resources which discuss pie charts as "not a desirable form of presentation"  [@brintonGraphicMethodsPresenting1914] and more strongly, "an insult to a man's intelligence" [@karsten1923charts].
+For instance, @eellsRelativeMeritsCircles1926 quotes two contemporary resources that discuss pie charts as "not a desirable form of presentation"  [@brintonGraphicMethodsPresenting1914] and, more strongly, "an insult to a man's intelligence" [@karsten1923charts].
 @brintonGraphicMethodsPresenting1914 even identifies that the "sector method" is common, but if horizontal bar methods were used more frequently, they would be preferred for both speed and accuracy.
@@
 
-<!-- Introspection is an incredibly powerful tool when applied to the design of statistical charts, and can lead to design guidelines that precede empirical tests but match their results.   -->
-<!-- However, empirical tests are also extremely important: Tufte's data-ink ratio is oft repeated, but empirical studies show that the relationship between minimalism and chart perception is much more complicated [@gillanMinimalismSyntaxGraphs1994a;@tufteVisualDisplayQuantitative2001;@parsonsDataVisualizationPractitioners2020].
+<!-- Introspection is an incredibly powerful tool when applied to the design of statistical charts and can lead to design guidelines that precede empirical tests but match their results.   -->
+<!-- However, empirical tests are also significant: Tufte's data-ink ratio is oft repeated, but empirical studies show that the relationship between minimalism and chart perception is much more complicated [@gillanMinimalismSyntaxGraphs1994a;@tufteVisualDisplayQuantitative2001;@parsonsDataVisualizationPractitioners2020].
 -->
@@
 Interestingly, the question of pies and bars motivated some of the first empirical evaluations of charts, published in the Journal of the American Statistical Association: @eellsRelativeMeritsCircles1926, @vonhuhnFurtherStudiesGraphic1927, and @croxtonBarChartsCircle1927.
-Experimenters approached the problem head-on, directly asking participants to estimate quantities from charts, and evaluated participants on accuracy, and in some cases, speed. 
+Experimenters approached the problem head-on, directly asking participants to estimate quantities from charts, and evaluated participants on accuracy and, in some cases, speed. 
 While more sophisticated testing methodologies have been developed over the past century [@vanderplasTestingStatisticalCharts2020], studies are still conducted today using direct estimation approaches [@vanderplasFramedReproducingRevisiting2019].
-Now as then, the phrasing of direct estimation questions is of primary importance: a simple mathematical transformation of the estimated quantity can have an outsized effect on the results.
+The phrasing of direct estimation questions is of primary importance: a simple mathematical transformation of the estimated quantity can have an outsized effect on the results.
 
-The initial studies of the question of pies and bars failed to identify a clear "best chart" across all situations, or even to clearly identify in which situations pie charts or bar charts should be preferred.
+The initial studies of the question of pies and bars failed to identify a clear "best chart" across all situations or even to clearly identify in which situations pie charts or bar charts should be preferred.
 @eellsRelativeMeritsCircles1926 found that pie charts were read as accurately as bar charts and with comparable speed.
-The simplified nature of the experiment attracted critiques from @vonhuhnFurtherStudiesGraphic1927; attached to these criticisms is a short study previously run by Croxton which examined estimates of the ratio between smaller and larger sectors of two-category pie and bar charts; these contributions will be addressed separately even though they are often cited as a single paper, even within the JASA records.
+The simplified nature of the experiment attracted critiques from @vonhuhnFurtherStudiesGraphic1927; attached to these criticisms is a short study previously run by Croxton, which examined estimates of the ratio between smaller and larger sectors of two-category pie and bar charts; these contributions will be addressed separately even though they are often cited as a single paper, even within the JASA records.
 @vonhuhnFurtherStudiesGraphic1927's paper provides several examples which highlight concerns with the Eells study, and argues that while pie charts may be better in certain simple situations, bar charts are much more extensible and facilitate more advanced comparisons, but does not include any experimental studies.
-<!-- Not much has changed today - experimental work is often criticized by those who provide toy examples but no empirical data, due to compromises necessary to ensure experimental control.  -->
+<!-- Not much has changed today - experimental work is often criticized by those who provide toy examples but no empirical data due to compromises necessary to ensure experimental control.  -->
 Croxton's early experimental data (attached to @vonhuhnFurtherStudiesGraphic1927) had a larger number of participants than @eellsRelativeMeritsCircles1926, but only tested two-category charts with only two proportions: .25 and .4.
@@
 
-Subsequently, @croxtonBarChartsCircle1927 published additional data assessing accuracy of estimates from pie and bar charts, varying number of categories and orientation/alignment of sections. 
-@croxtonBarChartsCircle1927 found that pie charts matched or surpassed bar charts in most situations, across different scale alignment and in charts with up to 5 categories (though at 5 categories, differences were not statistically significant).
+Subsequently, @croxtonBarChartsCircle1927 published additional data assessing the accuracy of estimates from pie and bar charts, varying number of categories, and orientation/alignment of sections. 
+@croxtonBarChartsCircle1927 found that pie charts matched or surpassed bar charts in most situations, across different scale alignments, and in charts with up to 5 categories (though differences were not statistically significant at 5 categories).
 The final paper in this early set of comparisons was @croxtonGraphicComparisonsBars1932, which examined a variety of representations of proportional data, but outside the context of a statistical chart; the experimental stimuli, while applicable to charts, show a focus on general perception (though published in JASA).
-Experimental graphics studies seem to have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.
+Experimental graphics studies have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.
 
@@
 Much later, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.
-These experiments were presented as supporting a hierarchy of graphical perception which evolved over the course of the sequence of studies.
-This ranking indicates that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.
+These experiments were presented as supporting a hierarchy of graphical perception that evolved over the course of the study sequence.
+This ranking indicates that aligned length judgments are more accurate than judgments based on area or angle, implying that bar charts are more accurately perceived than pie charts.
 While the full hierarchy of feature comprehension is often assumed to be experimentally derived, the experiments presented in the series of Cleveland & McGill papers only inform a few elements of this hierarchy.
@@
 The later studies show more of an interdisciplinary conversation around data displays, which have become a general tool not limited to statisticians. 
-As a result, the literature becomes fragmented; studies are conducted across different disciplines, with different methods, elicitation questions, protocols, and ultimately, different conclusions.
+As a result, the literature becomes fragmented; studies are conducted across different disciplines, with different methods, elicitation questions, protocols, and, ultimately, different conclusions.
 
@@
 
-In this study, we re-examine the historical literature surrounding the use of pie and bar charts, including heuristics as well as empirical studies.
-We trace guidelines back to experimental studies, where possible, and examine whether justifications based on experimental studies are in fact directly supported by those studies.
+In this study, we re-examine the historical literature surrounding the use of pie and bar charts, including heuristics and empirical studies.
+Where possible, we trace guidelines back to experimental studies and examine whether justifications based on experimental studies are, in fact, directly supported by those studies.
 We begin with an initial set of four JASA papers from the late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.
-We trace these studies forward in time, examining papers which cite these studies, summaries of the original studies, and any additional information provided relevant to pie and bar charts.
+We trace these studies forward in time, examining papers that cite them, summaries of the original studies, and any additional information provided relevant to pie and bar charts.
 In addition, we specifically assess the different designs used in each experimental study, examining how the type of elicitation method impacts the results, as shown in @fig-elicitation.
-Synthesizing these results, we use the fundamental question of "which is better, pies or bars?" to examine the history of experimental graphics and graphical design guidelines.
+Synthesizing these results, we use the fundamental question of "Which is better, pies or bars?" to examine the history of experimental graphics and graphical design guidelines.
 

### 2025-02-11 sdss-abstract/index.tex
Restructure w/ folders

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  10pt,
  letterpaper,
  twocolumn]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
\usepackage{url}
\usepackage{hyperref}
\usepackage{latexsym}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage{algorithm, algorithmic}
\usepackage[dvipsnames]{xcolor} % colors
\newcommand{\mt}[1]{{\textcolor{blue}{#1}}}
\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={100 years of Pies vs.~Bars},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{100 years of Pies vs.~Bars}
\vspace{-1in}
\date{}

\begin{document}
\maketitle

William Playfair introduced the pie and bar charts in the early 1800s,
and by the turn of the next century, strong opinions about the
effectiveness of alternative displays of the same data had been
publicized in textbooks and manuals. For instance, Eells (1926) quotes
two contemporary resources which discuss pie charts as ``not a desirable
form of presentation'' (Brinton 1914) and more strongly, ``an insult to
a man's intelligence'' (Karsten 1923). Brinton (1914) even identifies
that the ``sector method'' is common, but if horizontal bar methods were
used more frequently, they would be preferred for both speed and
accuracy. Eells (1926) also quotes an excerpt from an introductory
statistics textbook (Secrist 1925), which argues that pie charts require
more cognitive effort to use accurately than a comparable stacked bar
chart, calling pie diagrams ``clumsy and defective''. These insights
mirror today's status quo: pie charts are stubbornly common, even though
many design guidelines recommend against their use and a variety of
empirical studies identify problems with circular data representations .
Wickham (2013) previously noted the prescience of design guidelines
prior to empirical studies validating their recommendations across
different types of visualizations.

\subsection{The Dawn of Empirical
Graphics}\label{the-dawn-of-empirical-graphics}

Interestingly, the question of pies and bars motivated some of the first
empirical evaluations of charts, published in the Journal of the
American Statistical Association: Eells (1926), von Huhn (1927), and
Croxton and Stryker (1927). Experimenters approached the problem
head-on, directly asking participants to estimate quantities from
charts, and evaluated participants on accuracy, and in some cases,
speed. While more sophisticated testing methodologies have been
developed over the past century (Vanderplas, Cook, and Hofmann 2020),
studies are still conducted today using direct estimation approaches
(VanderPlas, Ryan, and Hofmann 2019). Now as then, the phrasing of
direct estimation questions is of primary importance: a simple
mathematical transformation of the estimated quantity can have an
outsized effect on the results.

The initial studies of the question of pies and bars failed to identify
a clear ``best chart'' across all situations, or even to clearly
identify in which situations pie charts or bar charts should be
preferred. Eells (1926) found that pie charts were read as accurately as
bar charts and with comparable speed. The simplified nature of the
experiment attracted critiques from von Huhn (1927); attached to these
criticisms is a short study previously run by Croxton which examined
estimates of the ratio between smaller and larger sectors of
two-category pie and bar charts; these contributions will be addressed
separately even though they are often cited as a single paper, even
within the JASA records. von Huhn (1927)'s paper provides several
examples which highlight concerns with the Eells study, and argues that
while pie charts may be better in certain simple situations, bar charts
are much more extensible and facilitate more advanced comparisons, but
does not include any experimental studies. Croxton's early experimental
data (attached to von Huhn (1927)) had a larger number of participants
than Eells (1926), but only tested two-category charts with only two
proportions: .25 and .4. In addition, Croxton's study elicited estimates
from participants as \(A/B\) comparisons, where \(A + B = 1\), while
Eells elicited estimates for \(A\) and \(B\) separately. This multi-year
back-and-forth conversation is interesting in part because it pinpoints
an early example of experimental statistical graphics as a discipline of
interest to statisticians for both personal investigation and as an
audience for the results.

Subsequently, Croxton and Stryker (1927) published additional data
assessing accuracy of estimates from pie and bar charts, varying number
of categories and orientation/alignment of sections. Croxton and Stryker
(1927) found that pie charts matched or surpassed bar charts in most
situations, across different scale alignment and in charts with up to 5
categories (though at 5 categories, differences were not statistically
significant). The final paper in this early set of comparisons was
Croxton and Stein (1932), which examined a variety of representations of
proportional data, but outside the context of a statistical chart; the
experimental stimuli, while applicable to charts, show a focus on
general perception (though published in JASA).

Experimental graphics studies seem to have largely gone unreported (if
they were conducted) between 1932 and the 1980s, particularly in
statistical publications, though the issue of pies and bars occasionally
pops up during this period in other disciplines.

\subsection{Modern Experimental
Graphics}\label{modern-experimental-graphics}

Much later, Cleveland and McGill (1984), a statistician and a
psychologist by training, respectively, conducted experiments on simple
graphical elements. These experiments were presented as supporting a
hierarchy of graphical perception which evolved over the course of the
sequence of studies. This ranking indicates that aligned length
judgments are more accurate than judgments based on area or angle, with
the implication that bar charts are more accurately perceived than pie
charts. While the full hierarchy of feature comprehension is often
assumed to be experimentally derived, the experiments presented in the
series of Cleveland \& McGill papers only inform a few elements of this
hierarchy. As in the late 1920s, the combination of introspection based
reasoning and a sequence of simple experiments did not go uncontested.
Spence and Lewandowsky (1991) (both psychologists) used forced-choice
questions of the form ``which is bigger, A or B+C'' to investigate the
accuracy of pie and bar charts for making part-to-whole comparisons. The
later studies show more of an interdisciplinary conversation around data
displays, which have become a general tool not limited to statisticians.
As a result, the literature becomes fragmented; studies are conducted
across different disciplines, with different methods, elicitation
questions, protocols, and ultimately, different conclusions.

\begin{figure}

\centering{

\includegraphics{pie-type-example.png}

}

\caption{\label{fig-elicitation}Different elicitation methods used to
assess accuracy of pie chart estimates.}

\end{figure}%

\subsection{Scope and Aims}\label{scope-and-aims}

In this study, we re-examine the historical literature surrounding the
use of pie and bar charts, including heuristics as well as empirical
studies. We trace guidelines back to experimental studies, where
possible, and examine whether justifications based on experimental
studies are in fact directly supported by those studies. We begin with
an initial set of four JASA papers from the late 1920s and early 1930s:
Eells (1926), Croxton and Stryker (1927), von Huhn (1927), and Croxton
and Stein (1932). We trace these studies forward in time, examining
papers which cite these studies, summaries of the original studies, and
any additional information provided relevant to pie and bar charts. In
addition, we specifically assess the different designs used in each
experimental study, examining how the type of elicitation method impacts
the results, as shown in Figure~\ref{fig-elicitation}. Synthesizing
these results, we use the fundamental question of ``which is better,
pies or bars?'' to examine the history of experimental graphics and
graphical design guidelines.{]}\{.svp\}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-brintonGraphicMethodsPresenting1914}
Brinton, Willard Cope. 1914. \emph{Graphic Methods for Presenting
Facts}. New York, The Engineering Magazine Company.
\url{http://archive.org/details/graphicmethodsfo00brinrich}.

\bibitem[\citeproctext]{ref-clevelandGraphicalPerceptionTheory1984}
Cleveland, William S., and Robert McGill. 1984. {``Graphical Perception:
Theory, Experimentation, and Application to the Development of Graphical
Methods.''} \emph{Journal of the American Statistical Association} 79
(387): 531--54. \url{https://doi.org/10.1080/01621459.1984.10478080}.

\bibitem[\citeproctext]{ref-croxtonGraphicComparisonsBars1932}
Croxton, Frederick E., and Harold Stein. 1932. {``Graphic Comparisons by
Bars, Squares, Circles, and Cubes.''} \emph{Journal of the American
Statistical Association} 27 (177): 54--60.

\bibitem[\citeproctext]{ref-croxtonBarChartsCircle1927}
Croxton, Frederick E., and Roy E. Stryker. 1927. {``Bar Charts Versus
Circle Diagrams.''} \emph{Journal of the American Statistical
Association} 22 (160): 473--82. \url{https://doi.org/10.2307/2276829}.

\bibitem[\citeproctext]{ref-eellsRelativeMeritsCircles1926}
Eells, Walter Crosby. 1926. {``The Relative Merits of Circles and Bars
for Representing Component Parts.''} \emph{Journal of the American
Statistical Association} 21 (154): 119--32.
\url{https://doi.org/10.2307/2277140}.

\bibitem[\citeproctext]{ref-karsten1923charts}
Karsten, K. G. 1923. \emph{Charts and {Graphs}: {An Introduction} to
{Graphic Methods} in the {Control} and {Analysis} of {Statistics}}.
Prentice-Hall, Incorporated.
\url{https://books.google.com/books?id=DNc1AAAAIAAJ}.

\bibitem[\citeproctext]{ref-secristIntroductionStatisticalMethods1925}
Secrist, Horace. 1925. \emph{An {Introduction} to {Statistical Methods}:
{A Textbook} for {College Students}, a {Manual} for {Statisticians} and
{Business Executives}}. Macmillan.
\url{https://www.google.com/books/edition/An_Introduction_to_Statistical_Methods/cuUBAAAAMAAJ?hl=en}.

\bibitem[\citeproctext]{ref-spenceDisplayingProportionsPercentages1991}
Spence, I., and S. Lewandowsky. 1991. {``Displaying Proportions and
Percentages.''} \emph{Applied Cognitive Psychology} 5 (1): 61--77.
\url{https://doi.org/10.1002/acp.2350050106}.

\bibitem[\citeproctext]{ref-vanderplasTestingStatisticalCharts2020}
Vanderplas, Susan, Dianne Cook, and Heike Hofmann. 2020. {``Testing
{Statistical Charts}: {What Makes} a {Good Graph}?''} \emph{Annual
Review of Statistics and Its Application} 7 (1): 61--88.
\url{https://doi.org/10.1146/annurev-statistics-031219-041252}.

\bibitem[\citeproctext]{ref-vanderplasFramedReproducingRevisiting2019}
VanderPlas, Susan, Goluch C Ryan, and Heike Hofmann. 2019. {``Framed!
{Reproducing} and {Revisiting} 150-{Year-Old Charts}.''} \emph{Journal
of Computational and Graphical Statistics} 28 (3): 620--34.
\url{https://doi.org/10.1080/10618600.2018.1562937}.

\bibitem[\citeproctext]{ref-vonhuhnFurtherStudiesGraphic1927}
von Huhn, R. 1927. {``Further Studies in the Graphic Use of Circles and
Bars: A Discussion of the Eell's Experiment.''} \emph{Journal of the
American Statistical Association} 22 (157): 31--39.
\url{https://doi.org/10.2307/2277346}.

\bibitem[\citeproctext]{ref-wickhamGraphicalCriticismHistorical2013}
Wickham, Hadley. 2013. {``Graphical {Criticism}: {Some Historical
Notes}.''} \emph{Journal of Computational and Graphical Statistics} 22
(1): 38--44. \url{https://doi.org/10.1080/10618600.2012.761140}.

\end{CSLReferences}

\bibliographystyle{sdss2020} % Please do not change the bibliography style

\end{document}

### 2025-02-11 sdss-abstract/toma.qmd
Restructure w/ folders

--- 2024-12-03 index.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-  10pt,
-  letterpaper,
-  twocolumn]{article}
+# Initial Studies
+<!-- XXX Who XXX?? -->
+<!-- [They demonstrated that the choice of visualization depends on the task and the data structure. -->
+<!-- Bars are preferable for comparisons and time-series data, while circles may be suitable for single distributions with grouped components.]{.mt} -->
 
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
-  \setmainfont[]{Times New Roman}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
+[Eells (1926, 1927) found that pie charts (circles) were more accurate than bar charts when presenting simple distributions, such as 50-50 or 25-75 relationships.
+However, bar charts were superior for more complex distributions and comparative tasks due to their linear scale, which facilitates comparisons across multiple diagrams.]{.mt}
+[Vohn (1927) argued that bar charts offer advantages for tasks requiring comparisons across time series or multiple distributions, owing to their standard scale and ease of labeling, while circles are more suited for showing numerous components grouped into categories.
+Frederic and Croxton (1932) compared bars, squares, circles, and cubes for simple comparisons.
+They confirmed that bar charts yielded the most accurate estimates, outperforming circles, squares, and cubes.
+Squares and circles were comparable in accuracy, while cubes consistently performed the worst due to the challenge of representing three-dimensional volume in two-dimensional drawings.
+These findings emphasize that linear encodings (e.g., bars) are more effective than area- or volume-based encodings for precise quantitative judgments.]{.mt} [Asking for absolute estimates of differences (A - B) likely gives the advantage to bar charts, while asking for part-to-whole estimates (A/(A+B)) would be expected to give the advantage to pie charts; asking for A/B ratio comparisons may produce yet a different result.]{.svp}
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+[In late 70's and 80's Studies on graphical perception reveal that position is the most accurate visual encoding method, followed by length and angle, with tasks involving area, volume, curvature, shading, and color saturation needing more reliability.
+Baird (1970) demonstrated that length is perceived more accurately than area and volume, which are often underestimated.
+Baird and Noma (1978) highlighted how visual cues, such as frames, improve accuracy by providing reference structures.
+Experiments comparing position and length judgments found that position tasks were significantly more accurate, with length judgments resulting in 40--250% higher errors.
+Similarly, comparisons of position and angle judgments revealed that pie charts, which rely on angle encodings, are far less effective than bar charts, with position judgments being nearly twice as accurate and accounting for significantly fewer errors.
+Biases and outliers were common, with position accuracy declining as distances increased.
+These findings emphasize the need to avoid traditional visualizations like pie and divided bar charts, recommending alternatives such as dot charts or framed rectangular charts, which leverage position-based encodings to improve accuracy and comprehension.
+These insights underscore the importance of redesigning data visualizations to optimize perceptual accuracy and usability.]{.mt}
 
-\usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
-\usepackage{url}
-\usepackage{hyperref}
-\usepackage{latexsym}
-\usepackage{amsmath, amsthm, amsfonts}
-\usepackage{algorithm, algorithmic}
-\usepackage[dvipsnames]{xcolor} % colors
-\newcommand{\mt}[1]{{\textcolor{blue}{#1}}}
-\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={100 years of Pies vs.~Bars},
-  pdfauthor={Maksuda Atkar Toma; Susan Vanderplas},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
+[While these studies are the earliest systematic studies of statistical graphics we could locate, modern statistical graphics research might be said to date from Cleveland & McGill's work on the perception of simple graphical elements; to include this work as well, we also look forward and backward from these papers, examining both cited papers and papers describing their work in subsequent research.]{.svp}
 
-\title{100 years of Pies vs.~Bars}
-\author{
-Maksuda Atkar Toma\\
-Statistics Department, \\University of Nebraska, Lincoln\\
-{\tt \href{mailto:mtoma2@huskers.unl.edu}{mtoma2@huskers.unl.edu}}\\
-\\\And
-Susan Vanderplas\\
-Statistics Department, \\University of Nebraska, Lincoln\\
-{\tt \href{mailto:svanderplas2@unl.edu}{svanderplas2@unl.edu}}\\
-}
-\date{}
+# Purpose and Motivation
 
-\begin{document}
-\maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
+Visualizing quantitative data effectively is essential in decision-making, scientific research, and education.
+Graphical tools must convey data precisely while engaging users to ensure optimal understanding.
+Yet, disagreements persist about which chart types best balance clarity, accessibility, and accuracy.
+Motivated by the need for improved visual communication, this research synthesizes findings from decades of studies in graphical perception, including experimental evaluations and historical insights.
+It aims to explore the relative merits of widely used visualizations, including bar charts, pie charts, and hybrid formats like grables, [@hinkGrablesEnableExtraction1998] which combine numerical tables with graphical clarity.
+The goal is to establish task-specific design principles and empower practitioners and educators to build more effective visual tools.
+(**Check this paragraph, took help from AI**)
 
-\hypertarget{historical-background}{%
-\section{Historical Background}\label{historical-background}}
+We will try to find out the importance of understanding the historical context of graphical methods, highlighting how their development reflects broader trends in statistical communication.
+We want to explore how modern visualization software (e.g., Tableau, Power BI) aligns with Cleveland and McGill's perceptual task hierarchy.
+Assess whether newer tools mitigate the limitations of suboptimal encodings like area and angle.\
+Another idea could be to investigate whether dynamic visualizations (e.g., interactive scatterplots, time-lapse animations) improve perceptual accuracy for traditionally low-ranking encodings such as area and slope.
 
-William Playfair introduced the pie and bar charts in the early
-1800s\svp{, and by the turn of the next century, strong opinions had
-been publicized in textbooks and manuals, and statisticians began
-conducting experimental studies comparing the two.} \svp{These studies
-have contradictory results:} Huhn (1927) concludes that bar charts are
-much more useful and efficient in the majority of situations, even
-though pie charts have specific applications\svp{, while Croxton and
-Stryker (1927) found that pie charts may match or surpass them in
-certain situations.} \svp{Much later, Cleveland and McGill (1984)
-conducted experiments on simple graphical elements, stating that aligned
-length judgments are more accurate than judgments based on area or
-angle, with the implication that bar charts are more accurately
-perceived than pie charts.} \svp{Their hierarchy of feature
-comprehension is often assumed to be experimentally derived, but the
-experiments presented in the papers (Cleveland and McGill 1984, 1985,
-1986) only covered a few elements of this hierarchy.}
+# Problem
 
-\svp{Here, we re-examine the historical literature surrounding the use
-of pie and bar charts, examining recommendations about the use of pie
-and bar charts, tracing these guidelines back to empirical studies, and
-evaluating whether the recommendations are well-supported by the
-studies.} \svp{We begin with an initial set of four studies from the
-late 1920s and early 1930s: Eells (1926), Croxton and Stryker (1927),
-Huhn (1927), and Croxton and Stein (1932).} \svp{We trace these studies
-forward in time, examining the claims made about the studies as well as
-the recommendations made in citing papers.} \svp{In addition, we lay out
-the different experimental designs used in each experimental study, with
-the goal of assessing how the design of each study might contribute to
-the ultimate results.} \mt{They demonstrated that the choice of
-visualization depends on the task and the data structure. Bars are
-preferable for comparisons and time-series data, while circles may be
-suitable for single distributions with grouped components.} \mt{Eells
-(1926, 1927) found that pie charts (circles) were more accurate than bar
-charts when presenting simple distributions, such as 50-50 or 25-75
-relationships. However, bar charts were superior for more complex
-distributions and comparative tasks due to their linear scale, which
-facilitates comparisons across multiple diagrams.} \mt{Vohn (1927)
-argued that bar charts offer advantages for tasks requiring comparisons
-across time series or multiple distributions, owing to their standard
-scale and ease of labeling, while circles are more suited for showing
-numerous components grouped into categories. Frederic and Croxton (1932)
-compared bars, squares, circles, and cubes for simple comparisons. They
-confirmed that bar charts yielded the most accurate estimates,
-outperforming circles, squares, and cubes. Squares and circles were
-comparable in accuracy, while cubes consistently performed the worst due
-to the challenge of representing three-dimensional volume in
-two-dimensional drawings. These findings emphasize that linear encodings
-(e.g., bars) are more effective than area- or volume-based encodings for
-precise quantitative judgments.} \svp{Asking for absolute estimates of
-differences (A - B) likely gives the advantage to bar charts, while
-asking for part-to-whole estimates (A/(A+B)) would be expected to give
-the advantage to pie charts; asking for A/B ratio comparisons may
-produce yet a different result.}
+While charts like bar and pie graphs are staples of data communication, their utility varies widely depending on the context.
+Pie charts often appeal due to their simplicity and visual aesthetics but falter in scenarios requiring detailed quantitative comparisons due to their reliance on angular and area-based judgments, which humans inherently struggle to process accurately.
+Bar charts, on the other hand, excel in tasks requiring precision, as they leverage a common positional scale.
+However, they lack the visual immediacy that non-technical audiences may seek.
+Meanwhile, emerging formats like grables combine the precision of tables with graphical representation but are underexplored in practice.
+This research addresses these gaps, seeking to clarify how these visualizations perform across different cognitive and perceptual tasks.
+In this modern era where AI has become very available to make such kinds of charts within a minute, how do humans' perspectives change in different fields?
+[**Paraphrased this section**]{.mt}
 
-\mt{In late 70's and 80's Studies on graphical perception reveal that
-position is the most accurate visual encoding method, followed by length
-and angle, with tasks involving area, volume, curvature, shading, and
-color saturation needing more reliability. Baird (1970) demonstrated
-that length is perceived more accurately than area and volume, which are
-often underestimated. Baird and Noma (1978) highlighted how visual cues,
-such as frames, improve accuracy by providing reference structures.
-Experiments comparing position and length judgments found that position
-tasks were significantly more accurate, with length judgments resulting
-in 40--250\% higher errors. Similarly, comparisons of position and angle
-judgments revealed that pie charts, which rely on angle encodings, are
-far less effective than bar charts, with position judgments being nearly
-twice as accurate and accounting for significantly fewer errors. Biases
-and outliers were common, with position accuracy declining as distances
-increased. These findings emphasize the need to avoid traditional
-visualizations like pie and divided bar charts, recommending
-alternatives such as dot charts or framed rectangular charts, which
-leverage position-based encodings to improve accuracy and comprehension.
-These insights underscore the importance of redesigning data
-visualizations to optimize perceptual accuracy and usability.}
+# Methods
 
-\svp{While these studies are the earliest systematic studies of
-statistical graphics we could locate, modern statistical graphics
-research might be said to date from Cleveland \& McGill's work on the
-perception of simple graphical elements; to include this work as well,
-we also look forward and backward from these papers, examining both
-cited papers and papers describing their work in subsequent research.}
+This study draws upon seminal experiments, including Cleveland and McGill's hierarchy of graphical encodings and Eells' foundational comparisons of pie and bar charts.
+A detailed review of tasks---value retrieval, trend analysis, and comparisons---provides insights into the cognitive load and accuracy associated with each visualization type.
+Historical analysis traces the evolution of statistical graphics, from William Playfair's pioneering work in bar and pie charts to modern hybrid visualizations.
+Experimental findings are organized into a taxonomy ranking encodings like position, length, area, and color in terms of effectiveness and alignment with specific tasks.
 
-\hypertarget{purpose-and-motivation}{%
-\section{Purpose and Motivation}\label{purpose-and-motivation}}
+# Conclusion
 
-Visualizing quantitative data effectively is essential in
-decision-making, scientific research, and education. Graphical tools
-must convey data precisely while engaging users to ensure optimal
-understanding. Yet, disagreements persist about which chart types best
-balance clarity, accessibility, and accuracy. Motivated by the need for
-improved visual communication, this research synthesizes findings from
-decades of studies in graphical perception, including experimental
-evaluations and historical insights. It aims to explore the relative
-merits of widely used visualizations, including bar charts, pie charts,
-and hybrid formats like grables, (Hink, Eustace, and S. Wogalter 1998)
-which combine numerical tables with graphical clarity. The goal is to
-establish task-specific design principles and empower practitioners and
-educators to build more effective visual tools. (\textbf{Check this
-paragraph, took help from AI})
+This research highlights the importance of tailoring visualizations to specific tasks and user contexts.
+Bar charts, emphasizing positional accuracy, remain the gold standard for detailed analyses, while pie charts can serve as accessible tools for summarizing information.
+Grables represent a promising innovation, combining clarity with precision for complex data sets.
+By integrating historical perspectives with modern insights from graphical perception studies, this research provides a comprehensive roadmap for designing effective visual tools.
+Practitioners, educators, and researchers are encouraged to adopt these principles, ensuring their visualizations communicate data effectively and foster deeper engagement and understanding.
 
-We will try to find out the importance of understanding the historical
-context of graphical methods, highlighting how their development
-reflects broader trends in statistical communication. We want to explore
-how modern visualization software (e.g., Tableau, Power BI) aligns with
-Cleveland and McGill's perceptual task hierarchy. Assess whether newer
-tools mitigate the limitations of suboptimal encodings like area and
-angle.\\
-Another idea could be to\,investigate whether dynamic visualizations
-(e.g., interactive scatterplots, time-lapse animations) improve
-perceptual accuracy for traditionally low-ranking encodings such as area
-and slope.
+This study emphasizes the relevance of designing visualizations for specific tasks and user settings.
+Bar charts, which emphasize positional accuracy, remain the gold standard for extensive analysis, although pie charts can be useful tools for summarizing information.
+Grables are a potential innovation that combines clarity and precision in big data sets.
+This study provides a complete roadmap for developing effective visual aids by combining historical viewpoints with contemporary insights from graphical perception studies.
 
-\hypertarget{problem}{%
-\section{Problem}\label{problem}}
-
-While charts like bar and pie graphs are staples of data communication,
-their utility varies widely depending on the context. Pie charts often
-appeal due to their simplicity and visual aesthetics but falter in
-scenarios requiring detailed quantitative comparisons due to their
-reliance on angular and area-based judgments, which humans inherently
-struggle to process accurately. Bar charts, on the other hand, excel in
-tasks requiring precision, as they leverage a common positional scale.
-However, they lack the visual immediacy that non-technical audiences may
-seek. Meanwhile, emerging formats like grables combine the precision of
-tables with graphical representation but are underexplored in practice.
-This research addresses these gaps, seeking to clarify how these
-visualizations perform across different cognitive and perceptual tasks.
-In this modern era where AI has become very available to make such kinds
-of charts within a minute, how do humans' perspectives change in
-different fields? \mt{\textbf{Paraphrased this section}}
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-This study draws upon seminal experiments, including Cleveland and
-McGill's hierarchy of graphical encodings and Eells' foundational
-comparisons of pie and bar charts. A detailed review of tasks---value
-retrieval, trend analysis, and comparisons---provides insights into the
-cognitive load and accuracy associated with each visualization type.
-Historical analysis traces the evolution of statistical graphics, from
-William Playfair's pioneering work in bar and pie charts to modern
-hybrid visualizations. Experimental findings are organized into a
-taxonomy ranking encodings like position, length, area, and color in
-terms of effectiveness and alignment with specific tasks.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-This research highlights the importance of tailoring visualizations to
-specific tasks and user contexts. Bar charts, emphasizing positional
-accuracy, remain the gold standard for detailed analyses, while pie
-charts can serve as accessible tools for summarizing information.
-Grables represent a promising innovation, combining clarity with
-precision for complex data sets. By integrating historical perspectives
-with modern insights from graphical perception studies, this research
-provides a comprehensive roadmap for designing effective visual tools.
-Practitioners, educators, and researchers are encouraged to adopt these
-principles, ensuring their visualizations communicate data effectively
-and foster deeper engagement and understanding.
-
-This study emphasizes the relevance of designing visualizations for
-specific tasks and user settings. Bar charts, which emphasize positional
-accuracy, remain the gold standard for extensive analysis, although pie
-charts can be useful tools for summarizing information. Grables are a
-potential innovation that combines clarity and precision in big data
-sets. This study provides a complete roadmap for developing effective
-visual aids by combining historical viewpoints with contemporary
-insights from graphical perception studies.
-
-\mt{This article also investigates the relevance of human visual
-perception in current tools such as Tableau and PowerBI Practitioners,
-educators, and researchers are advised to follow these guidelines to
-ensure that their visualizations effectively communicate data while
-fostering deeper engagement and understanding.}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-clevelandGraphicalPerceptionTheory1984}{}}%
-Cleveland, William S., and Robert McGill. 1984. {``Graphical Perception:
-Theory, Experimentation, and Application to the Development of Graphical
-Methods.''} \emph{Journal of the American Statistical Association} 79
-(387): 531--54. \url{https://doi.org/10.1080/01621459.1984.10478080}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-clevelandGraphicalPerceptionGraphical1985}{}}%
----------. 1985. {``Graphical Perception and Graphical Methods for
-Analyzing Scientific Data.''} \emph{Science} 229 (4716): 828--33.
-\url{https://doi.org/10.1126/science.229.4716.828}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-clevelandExperimentGraphicalPerception1986}{}}%
----------. 1986. {``An Experiment in Graphical Perception.''}
-\emph{International Journal of Man-Machine Studies} 25 (5): 491--500.
-\url{https://doi.org/10.1016/S0020-7373(86)80019-0}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-croxtonGraphicComparisonsBars1932}{}}%
-Croxton, Frederick E., and Harold Stein. 1932. {``Graphic Comparisons by
-Bars, Squares, Circles, and Cubes.''} \emph{Journal of the American
-Statistical Association} 27 (177): 54--60.
-
-\leavevmode\vadjust pre{\hypertarget{ref-croxtonBarChartsCircle1927}{}}%
-Croxton, Frederick E., and Roy E. Stryker. 1927. {``Bar Charts Versus
-Circle Diagrams.''} \emph{Journal of the American Statistical
-Association} 22 (160): 473--82. \url{https://doi.org/10.2307/2276829}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-eellsRelativeMeritsCircles1926}{}}%
-Eells, Walter Crosby. 1926. {``The Relative Merits of Circles and Bars
-for Representing Component Parts.''} \emph{Journal of the American
-Statistical Association} 21 (154): 119--32.
-\url{https://doi.org/10.2307/2277140}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hinkGrablesEnableExtraction1998}{}}%
-Hink, Jessica K, Jason K Eustace, and Michael S. Wogalter. 1998. {``Do
-Grables Enable the Extraction of Quantitative Information Better Than
-Pure Graphs or Tables?''} \emph{International Journal of Industrial
-Ergonomics} 22 (6): 439--47.
-\url{https://doi.org/10.1016/S0169-8141(97)00017-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vonhuhnFurtherStudiesGraphic1927}{}}%
-Huhn, R. von. 1927. {``Further Studies in the Graphic Use of Circles and
-Bars: A Discussion of the Eell's Experiment.''} \emph{Journal of the
-American Statistical Association} 22 (157): 31--39.
-\url{https://doi.org/10.2307/2277346}.
-
-\end{CSLReferences}
-
-\bibliographystyle{sdss2020} % Please do not change the bibliography style
-
-\end{document}
+[This article also investigates the relevance of human visual perception in current tools such as Tableau and PowerBI Practitioners, educators, and researchers are advised to follow these guidelines to ensure that their visualizations effectively communicate data while fostering deeper engagement and understanding.]{.mt}

### 2025-03-11 Draft.qmd
Draft

--- 2025-02-11 sdss-abstract/index.qmd
@@
     colorlinks: true
-# author:
-#   - name: Maksuda Aktar Toma
-#     affiliation:
-#       department: Statistics Department
-#       name: University of Nebraska, Lincoln
-#     email: mtoma2@huskers.unl.edu
-#     corresponding: true
-#   - name: Susan Vanderplas
-#     affiliation:
-#       department: Statistics Department
-#       name: University of Nebraska, Lincoln
-#     email: svanderplas2@unl.edu
-#     orcid: 0000-0002-3803-0972
+author:
+   - name: Maksuda Aktar Toma
+     affiliation:
+       department: Statistics Department
+       name: University of Nebraska, Lincoln
+     email: mtoma2@huskers.unl.edu
+     corresponding: true
+   - name: Susan Vanderplas
+     affiliation:
+       department: Statistics Department
+       name: University of Nebraska, Lincoln
+     email: svanderplas2@unl.edu
+     orcid: 0000-0002-3803-0972
 bibliographystyle: acl
@@
 
+# Abstract
+
 William Playfair introduced the pie and bar charts in the early 1800s, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.
@@
 These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use
-<!-- [@ichartsmarketingdepartment4ProblemsPie2017;@holtzIssuePieChart2018] -->
-and a variety of empirical studies identify problems with circular data representations 
-<!-- [@hofmannGraphicalTestsPower2012] -->
-.
+
 @wickhamGraphicalCriticismHistorical2013 previously noted the prescience of design guidelines prior to empirical studies validating their recommendations across different types of visualizations.
 
-<!-- Introspection is an incredibly powerful tool when applied to the design of statistical charts and can lead to design guidelines that precede empirical tests but match their results.   -->
-<!-- However, empirical tests are also significant: Tufte's data-ink ratio is oft repeated, but empirical studies show that the relationship between minimalism and chart perception is much more complicated [@gillanMinimalismSyntaxGraphs1994a;@tufteVisualDisplayQuantitative2001;@parsonsDataVisualizationPractitioners2020].
--->
+# 1. Introduction
 
-## The Dawn of Empirical Graphics
-Interestingly, the question of pies and bars motivated some of the first empirical evaluations of charts, published in the Journal of the American Statistical Association: @eellsRelativeMeritsCircles1926, @vonhuhnFurtherStudiesGraphic1927, and @croxtonBarChartsCircle1927.
-Experimenters approached the problem head-on, directly asking participants to estimate quantities from charts, and evaluated participants on accuracy and, in some cases, speed. 
-While more sophisticated testing methodologies have been developed over the past century [@vanderplasTestingStatisticalCharts2020], studies are still conducted today using direct estimation approaches [@vanderplasFramedReproducingRevisiting2019].
-The phrasing of direct estimation questions is of primary importance: a simple mathematical transformation of the estimated quantity can have an outsized effect on the results.
+# 2. The Dawn of Empirical Graphics
 
-The initial studies of the question of pies and bars failed to identify a clear "best chart" across all situations or even to clearly identify in which situations pie charts or bar charts should be preferred.
-@eellsRelativeMeritsCircles1926 found that pie charts were read as accurately as bar charts and with comparable speed.
-The simplified nature of the experiment attracted critiques from @vonhuhnFurtherStudiesGraphic1927; attached to these criticisms is a short study previously run by Croxton, which examined estimates of the ratio between smaller and larger sectors of two-category pie and bar charts; these contributions will be addressed separately even though they are often cited as a single paper, even within the JASA records.
-@vonhuhnFurtherStudiesGraphic1927's paper provides several examples which highlight concerns with the Eells study, and argues that while pie charts may be better in certain simple situations, bar charts are much more extensible and facilitate more advanced comparisons, but does not include any experimental studies.
-<!-- Not much has changed today - experimental work is often criticized by those who provide toy examples but no empirical data due to compromises necessary to ensure experimental control.  -->
-Croxton's early experimental data (attached to @vonhuhnFurtherStudiesGraphic1927) had a larger number of participants than @eellsRelativeMeritsCircles1926, but only tested two-category charts with only two proportions: .25 and .4.
-In addition, Croxton's study elicited estimates from participants as $A/B$ comparisons, where $A + B = 1$, while Eells elicited estimates for $A$ and $B$ separately. 
-<!-- It seems that prior to @eellsRelativeMeritsCircles1926, some researchers were empirically testing charts, but these experiments were not necessarily published for broader consumption; when the conversation arose in JASA, Croxton had data from years prior ready to provide additional insight.
--->
-This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.
+# 3.  Modern Experimental Graphics
 
-Subsequently, @croxtonBarChartsCircle1927 published additional data assessing the accuracy of estimates from pie and bar charts, varying number of categories, and orientation/alignment of sections. 
-@croxtonBarChartsCircle1927 found that pie charts matched or surpassed bar charts in most situations, across different scale alignments, and in charts with up to 5 categories (though differences were not statistically significant at 5 categories).
-The final paper in this early set of comparisons was @croxtonGraphicComparisonsBars1932, which examined a variety of representations of proportional data, but outside the context of a statistical chart; the experimental stimuli, while applicable to charts, show a focus on general perception (though published in JASA).
-Experimental graphics studies have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.
+# 4. Criticism of previous experimental design
 
-## Modern Experimental Graphics 
-Much later, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.
-These experiments were presented as supporting a hierarchy of graphical perception that evolved over the course of the study sequence.
-This ranking indicates that aligned length judgments are more accurate than judgments based on area or angle, implying that bar charts are more accurately perceived than pie charts.
-While the full hierarchy of feature comprehension is often assumed to be experimentally derived, the experiments presented in the series of Cleveland & McGill papers only inform a few elements of this hierarchy.
-As in the late 1920s, the combination of introspection based reasoning and a sequence of simple experiments did not go uncontested. 
-@spenceDisplayingProportionsPercentages1991 (both psychologists) used forced-choice questions of the form "which is bigger, A or B+C" to investigate the accuracy of pie and bar charts for making part-to-whole comparisons.
-The later studies show more of an interdisciplinary conversation around data displays, which have become a general tool not limited to statisticians. 
-As a result, the literature becomes fragmented; studies are conducted across different disciplines, with different methods, elicitation questions, protocols, and, ultimately, different conclusions.
-
-![Different elicitation methods used to assess accuracy of pie chart estimates.](pie-type-example.png){#fig-elicitation}
-
-## Scope and Aims
-
-In this study, we re-examine the historical literature surrounding the use of pie and bar charts, including heuristics and empirical studies.
-Where possible, we trace guidelines back to experimental studies and examine whether justifications based on experimental studies are, in fact, directly supported by those studies.
-We begin with an initial set of four JASA papers from the late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.
-We trace these studies forward in time, examining papers that cite them, summaries of the original studies, and any additional information provided relevant to pie and bar charts.
-In addition, we specifically assess the different designs used in each experimental study, examining how the type of elicitation method impacts the results, as shown in @fig-elicitation.
-Synthesizing these results, we use the fundamental question of "Which is better, pies or bars?" to examine the history of experimental graphics and graphical design guidelines.
+# 5. Field of comparison
 

### 2025-03-31 Draft.qmd
draft 1

@@
 
+The debate between pie charts and bar charts has been ongoing for over a century.
+Numerous studies with distinct methodologies and experimental designs have attempted to resolve this issue, but no single consensus has emerged.
+In this paper, we will provide a comprehensive summary of the existing literature, analyze the strengths and limitations of both chart types, and attempt to identify a unified solution based on empirical evidence and theoretical insights.
+
+It's really hard to reach in a conclusion that which one is better in just a one liner.
+We will review previous work on this topic, analyze their experimental designs, compare different scenarios, and attempt to generalize findings based on these studies.
+
+# 1. Introduction
 William Playfair introduced the pie and bar charts in the early 1800s, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.
@@
 
-# 1. Introduction
-
 # 2. The Dawn of Empirical Graphics
@@
 
+In 1991, Spence and Lewandowsky (@spenceDisplayingProportionsPercentages1991) demonstrated that pie charts can outperform bar charts under certain conditions when displaying proportions.
+They conducted a series of four experiments, each providing consistent insights into the comparative effectiveness of pie charts and bar charts for various tasks involving the interpretation of proportions.
+
+The first experiment investigated how people process pie charts, bar charts, and horizontal divided bar charts by measuring accuracy, processing time, and the effect of the number of components.
+Thirty participants were divided into three groups based on task type: simple comparison (A vs.
+B), comparison of a single value with a sum (A vs.
+B + C), and comparison of two sums (A + B vs.
+C + D).
+Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones.
+Bar charts performed slightly better than pie charts for simple comparisons under time pressure, but differences between chart types were negligible for complex tasks.
+
+The second experiment yielded similar results but employed a different approach.
+It introduced a table instead of a horizontal bar chart, fixed the processing time at 3.0 seconds, reduced the number of component levels, and adopted a within-subject design for tasks.
+Additionally, it tested the effect of ordering components by magnitude.
+The results confirmed the superiority of bar charts over pie charts for accuracy, particularly for simpler tasks and fewer components.
+However, the lack of an ordering effect challenges the common belief that ordered components improve perception.
+
+The third experiment examined how task complexity and display type (pie chart, bar chart, and table) affect performance using a within-subject design, where each subject performed all three task types.
+Unlike Experiment 1, which used a between-subjects design, and Experiment 2, which introduced tables and fixed processing time, Experiment 3 retained pie charts, bar charts, and tables while excluding the horizontal divided bar chart.
+Processing time was fixed at 3.0 seconds, and two levels of components (4 and 7) were tested.
+Accuracy declined with task complexity, and while pie and bar charts showed similar performance for simpler tasks, pie charts outperformed the others for complex tasks, likely due to easier component identification and combination.
+Adjacency influenced bar chart performance but not pie charts, suggesting that perceptual and cognitive factors favor pie charts for complex tasks.
+
+The fourth experiment tested whether color coding could enhance performance by simplifying component identification.
+The design was identical to Experiment 3, except components were color-coded (red or blue) instead of using alphabetical labels.
+Processing time was reduced to 1.0 second since the task was expected to be easier with color cues.
+Accuracy was higher for pie (86%) and bar (84%) charts than for tables (75%), with performance declining as task complexity increased (88%, 82%, and 76% for simple to complex tasks).
+Unlike previous experiments, the number of components had no effect on accuracy, suggesting that once components were easily located through color, cognitive comparison was unaffected by complexity.
+Similar to Experiment 3, pie charts outperformed bar charts for complex tasks, but bar charts were better for simpler tasks.
+This indicates that color coding can improve performance in complex charts by simplifying component identification.
+
+The experiment has several limitations.
+The small and homogeneous sample (mainly students) limits the generalizability of the findings.
+Short exposure times (1.0–6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.
+The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance.
+Reducing exposure time to 1.0 second in Experiment 4 may have biased results toward pie charts by encouraging intuitive rather than analytical processing
+
 # 5. Field of comparison
 
-# References
+# 6. Future Research
+
+# 7. References

### 2025-04-07 Draft.qmd
updated

@@
 
-# 4. Criticism of previous experimental design
+# 4. Criticism of previous Experimental design
+
+## 4.1 Experiment Description
+
+This poster synthesizes findings from a series of historical and modern experiments comparing pie and bar charts for interpreting proportional data.
+Early studies, such as Eells (1926) and Croxton & von Huhn (1927), focused on accuracy and speed in estimating single or dual category proportions (e.g., A vs.
+B or A / (A + B)).
+These experiments typically involved participants identifying which portion of a chart was larger or estimating ratios between segments.
+
+Modern experiments, such as Spence & Lewandowsky (1991), expanded the task complexity by asking participants to compare single components (A vs.
+B), a component against a pair (A vs.
+B + C), and two pairs (A + B vs.
+C + D).
+These tasks were tested across multiple chart formats (pie, bar, and table), with variations in exposure time, number of components, and use of color or ordering.
+Performance was evaluated in terms of speed and accuracy, allowing researchers to assess how perceptual and cognitive demands shift across visualization types and task difficulty.
+
+Peterson and Schramm (1954) evaluated the accuracy of proportion estimation across eight common chart types used to represent parts of a whole: circle, disc, single bar, multiple bar, multiple cylinder, multiple square column, multiple area column, and partial cosmograph.
+Each chart displayed five fixed proportions (43%, 23%, 18%, 10%, 6%) and was presented alongside a similar “decoy” graph to mask repetition.
+A total of 112 U.S.
+Air Force personnel viewed the charts in random order and estimated the percentage values for each component.
+After discarding incomplete responses, 86 participants contributed 3,440 data points, forming the basis for comparison across visual formats.
+
+Cleveland & McGill (1984) conducted a series of experiments to test a theory of graphical perception based on “elementary perceptual tasks” such as position, length, angle, area, and shading.
+They hypothesized a hierarchy of these tasks based on how accurately people extract quantitative information from them.
+In their experiments, participants viewed a range of common graph types—including bar charts, pie charts, and divided bar charts—and were asked to make visual judgments of values or differences.
+The results supported their theory: judgments involving position along a common scale were the most accurate, followed by length, with angle (used in pie charts) yielding the least accurate responses.
+These findings laid the foundation for many modern principles in data visualization design.
+
+Cleveland & McGill (1986) conducted a foundational experiment to evaluate the accuracy of seven basic graphical perception tasks: position (aligned and non-aligned), length, angle, slope, and area (circle and blob).
+Participants were shown visual stimuli in which one object served as a standard, and three comparison objects were judged as percentages relative to it.
+Each participant made 210 judgments across 70 displays.
+The study included 127 participants from diverse educational backgrounds.
+Results confirmed a perceptual hierarchy: position judgments were the most accurate, while angle and area were less reliable—insights that continue to influence data visualization design today.
+
+Simkin and Hastie (1987) investigated how different graphical encodings—position, length, and angle—affect performance on visual judgment tasks.
+200 Undergraduate participants were shown pie charts, bar charts, and divided bar charts and asked to make either comparison judgments (A/B) or proportion-of-the-whole judgments.
+Each graph type relied on a different visual code: bar charts used position, divided bars used length, and pie charts used angle.
+Participants completed 90 trials under timed conditions.
+Results showed that position and length led to higher accuracy and faster response times for comparison tasks, while pie charts supported better performance in proportion-of-the-whole judgments, demonstrating a clear interaction between graph type and task type.
+
+Very recently, more works have been discovered on this debate that catch researchers’ eyes on the topic.
+Kosara & Skau (2016) investigated how common distortions in pie charts affect judgment accuracy.
+Participants were shown a series of pie chart variations and asked to estimate the percentage represented by a highlighted slice.
+The variations included a standard pie chart, an "exploded" pie (with a slice offset), a "larger slice" chart (with an extended radius), a vertically compressed elliptical pie, and a square-shaped chart.
+All charts had two segments and were rotated randomly to eliminate positional bias.
+Participants were recruited via Mechanical Turk and asked to provide whole-number estimates for values ranging from 3% to 97%.
+Results showed that all distortions—except the exploded pie—led to significantly higher judgment errors compared to the baseline pie chart.
+
+In 2025, Hill did the study that involved three online experiments designed to evaluate how effectively bar, pie, and donut charts support part-whole estimation and ranking tasks.
+In Experiment 1, participants viewed randomized sets of charts (10 bar, 5 pie, 5 donut) created from multinomially generated data and answered ranking and proportion estimation questions.
+Experiment 2 focused on charts where elements A and B were similar in size, allowing for precise comparison of visual discrimination accuracy across chart types.
+Experiment 3 assessed relative magnitude estimation on pie and donut charts using a direct comparison task (“If A is 100, what is B?”).
+A total of 42 participants from the University of Surrey took part in the study via Qualtrics, and all data were anonymized
+
+## 4.2 Criticism of previous Experimental design
 

### 2025-04-08 Draft.qmd
updated-1

@@
 
-# 4. Criticism of previous Experimental design
+# 4.  Experimental design
 
@@
 
+# 5. Measurement and Preference
+
+Eells (1926) didn’t just look at how well people performed—he also explored how they felt about the charts and how they approached the task.
+When it came to accuracy, pie charts actually did better than bar charts, especially as the number of segments increased.
+People also tended to move a bit faster with pie charts, completing slightly more within five minutes.
+One of the most striking findings was that bar charts led to 70% more large errors, making them less reliable for precise judgment.
+After the task, students were asked which chart type they preferred, and the majority chose pie charts for showing part-whole data.
+When it came to strategy, most students said they judged the slices using arc length, with others using angles or area—and all three methods turned out to be about equally accurate.
+Overall, the results challenged the common criticism that pie charts are hard to read.
+
 # 5. Field of comparison
 
-# 6. Future Research
+# 6. Result
+Which chart works best really depends on the task.
+When people are asked to simply compare two values (A vs B), bar charts consistently come out on top—they’re faster and more accurate, and participants tend to prefer them for that reason (Spence & Lewandowsky, 1991; Simkin & Hastie, 1987; Hill, 2025).
+But when the task shifts to estimating what proportion one part is of the whole (A / A+B), pie charts do better, likely because they align with how we naturally think about part-whole relationships (Cleveland & McGill, 1984; Simkin & Hastie, 1987).
 
-# 7. References
+For more complex comparisons—like A vs B+C or A+B vs C+D—the results are more mixed.
+Bar charts have a slight edge when comparing a single value to a sum (A vs B+C), but when comparing two grouped values (A+B vs C+D), pie charts actually perform better (Spence & Lewandowsky, 1991).
+This might be because it's easier to visually group adjacent slices than non-aligned bars (Kosara & Skau, 2016).
+
+When it comes to ranking multiple parts, like identifying the top three largest values, bar charts win again, as their alignment allows for quicker visual scanning (Hill, 2025; Kosara & Skau, 2016).
+And in magnitude estimation tasks (e.g., “If A is 100, what is B?”), both pie and donut charts tend to outperform bars, though people are more likely to misjudge values slightly in those formats (Kosara & Skau, 2016; Hill, 2025).
+
+In the end, there’s no universal winner—the best chart depends on the specific task and what kind of judgment you're asking the reader to make.
+
+![*Grouped Bar Chart*](Grouped_Bar_Chart.png)
+
+# 7. Future Research
+
+# 8. References

### 2025-04-08 Draft.qmd
Start on modifications

@@
       - text: |
-          \usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
           \usepackage{url}
@@
           \newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
-    classoption: [twocolumn]
     mainfont: Times New Roman
@@
      orcid: 0000-0002-3803-0972
-bibliographystyle: acl
+
 bibliography: refs.bib
@@
 
-# 1. Introduction
+# Introduction
 William Playfair introduced the pie and bar charts in the early 1800s, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.
@@
 
-# 2. The Dawn of Empirical Graphics
+# The Dawn of Empirical Graphics
 
-# 3.  Modern Experimental Graphics
+#  Modern Experimental Graphics
 
-# 4.  Experimental design
+#  Experimental design
 
@@
 This poster synthesizes findings from a series of historical and modern experiments comparing pie and bar charts for interpreting proportional data.
-Early studies, such as Eells (1926) and Croxton & von Huhn (1927), focused on accuracy and speed in estimating single or dual category proportions (e.g., A vs.
-B or A / (A + B)).
+Early studies, such as @eellsRelativeMeritsCircles1926 and @croxtonBarChartsCircle1927, focused on accuracy and speed in estimating single or dual category proportions (e.g., $A/B$ or $A / (A + B)$).
 These experiments typically involved participants identifying which portion of a chart was larger or estimating ratios between segments. 
+\svp{In which experiment did they have to only identify which part of the chart was larger??}
 
-Modern experiments, such as Spence & Lewandowsky (1991), expanded the task complexity by asking participants to compare single components (A vs.
+Modern experiments, such as @spenceDisplayingProportionsPercentages1991, expanded the task complexity by asking participants to compare single components (A vs.
 B), a component against a pair (A vs.
@@
 C + D).
-These tasks were tested across multiple chart formats (pie, bar, and table), with variations in exposure time, number of components, and use of color or ordering.
+These tasks were tested across multiple formats (pie, bar, and table), with variations in exposure time, number of components, and use of color or ordering. 
 Performance was evaluated in terms of speed and accuracy, allowing researchers to assess how perceptual and cognitive demands shift across visualization types and task difficulty.
@@
 Peterson and Schramm (1954) evaluated the accuracy of proportion estimation across eight common chart types used to represent parts of a whole: circle, disc, single bar, multiple bar, multiple cylinder, multiple square column, multiple area column, and partial cosmograph.
-Each chart displayed five fixed proportions (43%, 23%, 18%, 10%, 6%) and was presented alongside a similar “decoy” graph to mask repetition.
+Each chart displayed five fixed proportions (43%, 23%, 18%, 10%, 6%) and was presented alongside a similar "decoy" graph to mask repetition.
 A total of 112 U.S.
@@
 
-Cleveland & McGill (1984) conducted a series of experiments to test a theory of graphical perception based on “elementary perceptual tasks” such as position, length, angle, area, and shading.
+Cleveland & McGill (1984) conducted a series of experiments to test a theory of graphical perception based on "elementary perceptual tasks" such as position, length, angle, area, and shading.
 They hypothesized a hierarchy of these tasks based on how accurately people extract quantitative information from them.
@@
 Experiment 2 focused on charts where elements A and B were similar in size, allowing for precise comparison of visual discrimination accuracy across chart types.
-Experiment 3 assessed relative magnitude estimation on pie and donut charts using a direct comparison task (“If A is 100, what is B?”).
+Experiment 3 assessed relative magnitude estimation on pie and donut charts using a direct comparison task ("If A is 100, what is B?").
 A total of 42 participants from the University of Surrey took part in the study via Qualtrics, and all data were anonymized
@@
 
-# 5. Measurement and Preference
+# Measurement and Preference
 
@@
 
-# 5. Field of comparison
+# Field of comparison
 
-# 6. Result
+# Result
 Which chart works best really depends on the task.
@@
 When it comes to ranking multiple parts, like identifying the top three largest values, bar charts win again, as their alignment allows for quicker visual scanning (Hill, 2025; Kosara & Skau, 2016).
-And in magnitude estimation tasks (e.g., “If A is 100, what is B?”), both pie and donut charts tend to outperform bars, though people are more likely to misjudge values slightly in those formats (Kosara & Skau, 2016; Hill, 2025).
+And in magnitude estimation tasks (e.g., "If A is 100, what is B?"), both pie and donut charts tend to outperform bars, though people are more likely to misjudge values slightly in those formats (Kosara & Skau, 2016; Hill, 2025).
 
@@
 
-![*Grouped Bar Chart*](Grouped_Bar_Chart.png)
+![Grouped Bar Chart](Grouped_Bar_Chart.png)
 
-# 7. Future Research
+# Future Research
 
-# 8. References
+# References

### 2025-04-08 Draft.qmd
Updates

@@
 
-## 4.1 Experiment Description
+## Experiment Description
 
@@
 
+\svp{Add study to bibliography}
 Peterson and Schramm (1954) evaluated the accuracy of proportion estimation across eight common chart types used to represent parts of a whole: circle, disc, single bar, multiple bar, multiple cylinder, multiple square column, multiple area column, and partial cosmograph.
@@
 
-Cleveland & McGill (1984) conducted a series of experiments to test a theory of graphical perception based on "elementary perceptual tasks" such as position, length, angle, area, and shading.
+@clevelandGraphicalPerceptionTheory1984 [-@clevelandGraphicalPerceptionGraphical1985] conducted a series of experiments to test a theory of graphical perception based on "elementary perceptual tasks" such as position, length, angle, area, and shading.
 They hypothesized a hierarchy of these tasks based on how accurately people extract quantitative information from them. 
@@
 
-Cleveland & McGill (1986) conducted a foundational experiment to evaluate the accuracy of seven basic graphical perception tasks: position (aligned and non-aligned), length, angle, slope, and area (circle and blob).
+@clevelandExperimentGraphicalPerception1986 conducted a \svp{follow-up} experiment to evaluate the accuracy of seven basic graphical perception tasks: position (aligned and non-aligned), length, angle, slope, and area (circle and blob).
 Participants were shown visual stimuli in which one object served as a standard, and three comparison objects were judged as percentages relative to it.
@@
 Results confirmed a perceptual hierarchy: position judgments were the most accurate, while angle and area were less reliable—insights that continue to influence data visualization design today. 
+\svp{Did the experiment actually cover all 7 tasks??? Please be careful with how you cite these, and provide more detail -- that's what we're specifically looking for here.}
 
 Simkin and Hastie (1987) investigated how different graphical encodings—position, length, and angle—affect performance on visual judgment tasks. 
-200 Undergraduate participants were shown pie charts, bar charts, and divided bar charts and asked to make either comparison judgments (A/B) or proportion-of-the-whole judgments.
+200 Undergraduate participants were shown pie charts, bar charts, and divided bar charts and asked to make either comparison judgments ($A/B$) or proportion-of-the-whole \svp{($A/(A+B)$)} judgements.
 Each graph type relied on a different visual code: bar charts used position, divided bars used length, and pie charts used angle. 
@@
 
+\svp{Add study to bibliography}
 Very recently, more works have been discovered on this debate that catch researchers’ eyes on the topic. 
@@
 
-## 4.2 Criticism of previous Experimental design
+## Criticism of previous experimental design
 
-In 1991, Spence and Lewandowsky (@spenceDisplayingProportionsPercentages1991) demonstrated that pie charts can outperform bar charts under certain conditions when displaying proportions.
+@spenceDisplayingProportionsPercentages1991 demonstrated that pie charts can outperform bar charts under certain conditions when displaying proportions.
 They conducted a series of four experiments, each providing consistent insights into the comparative effectiveness of pie charts and bar charts for various tasks involving the interpretation of proportions.
@@
 The first experiment investigated how people process pie charts, bar charts, and horizontal divided bar charts by measuring accuracy, processing time, and the effect of the number of components. 
-Thirty participants were divided into three groups based on task type: simple comparison (A vs.
-B), comparison of a single value with a sum (A vs.
-B + C), and comparison of two sums (A + B vs.
-C + D).
+Thirty participants were divided into three groups based on task type: simple comparison ($A$ vs.
+$B$), comparison of a single value with a sum ($A$ vs.
+$B + C$), and comparison of two sums ($A + B$ vs.
+$C + D$).
+\svp{Were participants estimating ratios, or just judging larger/smaller?}
 Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones. 
@@
 However, the lack of an ordering effect challenges the common belief that ordered components improve perception.
+\svp{How many categories were used?
+It's possible that this is true for small tables but not as true for larger tables.}
 
@@
 Unlike Experiment 1, which used a between-subjects design, and Experiment 2, which introduced tables and fixed processing time, Experiment 3 retained pie charts, bar charts, and tables while excluding the horizontal divided bar chart.
-Processing time was fixed at 3.0 seconds, and two levels of components (4 and 7) were tested.
+Processing time was fixed at 3.0 seconds \svp{You can fix stimulus time but not processing time -- processing takes as long as it takes}, and two levels of components (4 and 7) were tested. 
 Accuracy declined with task complexity, and while pie and bar charts showed similar performance for simpler tasks, pie charts outperformed the others for complex tasks, likely due to easier component identification and combination.
@@
 Short exposure times (1.0–6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.
+\svp{Why do psycologists use short exposure times?
+Look into this -- there's a reason}
 The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance. 
@@
 
-Eells (1926) didn’t just look at how well people performed—he also explored how they felt about the charts and how they approached the task.
+@eellsRelativeMeritsCircles1926 didn’t just look at how well people performed—he also explored how they felt about the charts and how they approached the task. 
 When it came to accuracy, pie charts actually did better than bar charts, especially as the number of segments increased. 
 People also tended to move a bit faster with pie charts, completing slightly more within five minutes.
+\svp{Eells didn't do the hypothesis test -- can you from provided information?
+Was it significant?
+Be careful about uncritically reporting results.}
 One of the most striking findings was that bar charts led to 70% more large errors, making them less reliable for precise judgment.
@@
 When people are asked to simply compare two values (A vs B), bar charts consistently come out on top—they’re faster and more accurate, and participants tend to prefer them for that reason (Spence & Lewandowsky, 1991; Simkin & Hastie, 1987; Hill, 2025).
-But when the task shifts to estimating what proportion one part is of the whole (A / A+B), pie charts do better, likely because they align with how we naturally think about part-whole relationships (Cleveland & McGill, 1984; Simkin & Hastie, 1987).
+When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (Cleveland & McGill, 1984; Simkin & Hastie, 1987).
+\svp{Is this still true when A and B aren't adjacent in e.g.
+a 3-4 component chart?}
 
-For more complex comparisons—like A vs B+C or A+B vs C+D—the results are more mixed.
-Bar charts have a slight edge when comparing a single value to a sum (A vs B+C), but when comparing two grouped values (A+B vs C+D), pie charts actually perform better (Spence & Lewandowsky, 1991).
+For more complex comparisons—like $A$ vs $B+C$ or $A+B$ vs $C+D$, the results are more mixed. 
+Bar charts have a slight edge when comparing a single value to a sum ($A$ vs $B+C$), but when comparing two grouped values ($A+B$ vs $C+D$), pie charts actually perform better (Spence & Lewandowsky, 1991).
 This might be because it's easier to visually group adjacent slices than non-aligned bars (Kosara & Skau, 2016).

### 2025-04-08 Draft.qmd
Remove template partials and other things from sdss template that aren't relevant

@@
     fontsize: 10pt
-    template-partials:
-      - title.tex
-      - before-bib.tex
     include-in-header:
@@
           \usepackage{amsmath, amsthm, amsfonts}
-          \usepackage{algorithm, algorithmic}
           \usepackage[dvipsnames]{xcolor} % colors
@@
        department: Statistics Department
-       name: University of Nebraska, Lincoln
+       name: University of Nebraska-Lincoln
      email: svanderplas2@unl.edu

### 2025-04-15 Draft.qmd
New draft

@@
 # Introduction
+
 William Playfair introduced the pie and bar charts in the early 1800s, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.
@@
 
+Interestingly, the question of pies and bars motivated some of the first empirical evaluations of charts, published in the Journal of the American Statistical Association: @eellsRelativeMeritsCircles1926, @vonhuhnFurtherStudiesGraphic1927, and @croxtonBarChartsCircle1927.
+Experimenters approached the problem head-on, directly asking participants to estimate quantities from charts, and evaluated participants on accuracy and, in some cases, speed.
+While more sophisticated testing methodologies have been developed over the past century [@vanderplasTestingStatisticalCharts2020], studies are still conducted today using direct estimation approaches [@vanderplasFramedReproducingRevisiting2019].
+The phrasing of direct estimation questions is of primary importance: a simple mathematical transformation of the estimated quantity can have an outsized effect on the results.
+
+The initial studies of the question of pies and bars failed to identify a clear "best chart" across all situations or even to clearly identify in which situations pie charts or bar charts should be preferred.
+@eellsRelativeMeritsCircles1926 found that pie charts were read as accurately as bar charts and with comparable speed.
+The simplified nature of the experiment attracted critiques from @vonhuhnFurtherStudiesGraphic1927; attached to these criticisms is a short study previously run by Croxton, which examined estimates of the ratio between smaller and larger sectors of two-category pie and bar charts; these contributions will be addressed separately even though they are often cited as a single paper, even within the JASA records.
+@vonhuhnFurtherStudiesGraphic1927's paper provides several examples which highlight concerns with the Eells study, and argues that while pie charts may be better in certain simple situations, bar charts are much more extensible and facilitate more advanced comparisons, but does not include any experimental studies.
+<!-- Not much has changed today - experimental work is often criticized by those who provide toy examples but no empirical data due to compromises necessary to ensure experimental control.
+--> Croxton's early experimental data (attached to @vonhuhnFurtherStudiesGraphic1927) had a larger number of participants than @eellsRelativeMeritsCircles1926, but only tested two-category charts with only two proportions: .25 and .4.
+In addition, Croxton's study elicited estimates from participants as $A/B$ comparisons, where $A + B = 1$, while Eells elicited estimates for $A$ and $B$ separately.
+<!-- It seems that prior to @eellsRelativeMeritsCircles1926, some researchers were empirically testing charts, but these experiments were not necessarily published for broader consumption; when the conversation arose in JASA, Croxton had data from years prior ready to provide additional insight.
+--> This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.
+
+Subsequently, @croxtonBarChartsCircle1927 published additional data assessing the accuracy of estimates from pie and bar charts, varying number of categories, and orientation/alignment of sections.
+@croxtonBarChartsCircle1927 found that pie charts matched or surpassed bar charts in most situations, across different scale alignments, and in charts with up to 5 categories (though differences were not statistically significant at 5 categories).
+The final paper in this early set of comparisons was @croxtonGraphicComparisonsBars1932, which examined a variety of representations of proportional data, but outside the context of a statistical chart; the experimental stimuli, while applicable to charts, show a focus on general perception (though published in JASA).
+Experimental graphics studies have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.
+
 # Modern Experimental Graphics
+
+Much later, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.
+These experiments were presented as supporting a hierarchy of graphical perception that evolved over the course of the study sequence.
+This ranking indicates that aligned length judgments are more accurate than judgments based on area or angle, implying that bar charts are more accurately perceived than pie charts.
+While the full hierarchy of feature comprehension is often assumed to be experimentally derived, the experiments presented in the series of Cleveland & McGill papers only inform a few elements of this hierarchy.
+As in the late 1920s, the combination of introspection based reasoning and a sequence of simple experiments did not go uncontested.
+@spenceDisplayingProportionsPercentages1991 (both psychologists) used forced-choice questions of the form "which is bigger, A or B+C" to investigate the accuracy of pie and bar charts for making part-to-whole comparisons.
+The later studies show more of an interdisciplinary conversation around data displays, which have become a general tool not limited to statisticians.
+As a result, the literature becomes fragmented; studies are conducted across different disciplines, with different methods, elicitation questions, protocols, and, ultimately, different conclusions.
+
+![Different elicitation methods used to assess accuracy of pie chart estimates.](pie-type-example.png)
 
@@
 
-This poster synthesizes findings from a series of historical and modern experiments comparing pie and bar charts for interpreting proportional data.
+This research synthesizes findings from a series of historical and modern experiments comparing pie and bar charts for interpreting proportional data.
 Early studies, such as @eellsRelativeMeritsCircles1926 and @croxtonBarChartsCircle1927, focused on accuracy and speed in estimating single or dual category proportions (e.g., $A/B$ or $A / (A + B)$).
 These experiments typically involved participants identifying which portion of a chart was larger or estimating ratios between segments.
+
+@eellsRelativeMeritsCircles1926 designed a study to objectively assess the accuracy of pie versus bar charts for representing proportions.
+Psychology students at Whitman College (n = 97) were first shown a page containing 15 subdivided pie charts, each representing 100%.
+Students were asked to write their best estimate of the percentage represented by each sector, using whole numbers only and avoiding any measuring tools.
+
+Three days later, the same students were given an equivalent set of bar charts showing the same proportions, and were again asked to estimate the percentages for each section.
+The design intentionally favored the bars (e.g., common alignment of bar bases) to test whether any advantage of the pie chart could still be observed.
+After completing the estimation tasks, participants reported the mental strategy they used (area, angle, arc length, or chord), and indicated which chart type they preferred.
+
+\mt{Eells (1926) did not ask participants to simply identify which segment was larger.
+Instead, it was a proportion estimation task—participants were asked to write down the percentage each segment represented of the whole ( A / (A + B + ...)), not to make a binary comparison (A vs B).
+The “which is larger” comparison task appears in later studies, such as @croxtonBarChartsCircle1927 and @spenceDisplayingProportionsPercentages1991}
+
+The large-scale study of @croxtonBarChartsCircle1927 aimed to determine which chart type---bar or circle---enabled more accurate estimation of component percentages.
+A total of 807 participants evaluated 27 charts, each depicting known part-to-whole proportions.
+For each percentage combination (ranging from 2 to 5 parts), one bar chart and one pie chart were created, yielding 13 matched pairs.
+Each chart was shown individually (without any numerical scale), and participants were instructed to numerically estimate the percentage value of each segment---not simply identify which was larger.
+This distinction is important that like @eellsRelativeMeritsCircles1926 , this was a proportion estimation task ( "What percent is this segment?"), not a comparison task ("Which is larger, A\>B?").
+Tasks requiring direct identification of the larger part appear in later studies, such as @spenceDisplayingProportionsPercentages1991 Each chart was printed on a large 22×28 inch card and shown one at a time in a randomized sequence to reduce bias.
+Some pie charts were repeated with slice positions altered (25% at 1:30 vs. 6:00 orientation) to test whether visual alignment influenced judgment accuracy.
+Unlike Eells, this study focused exclusively on accuracy, and results showed that while bar charts outperformed in a few specific conditions, pie charts generally led to more accurate estimates, especially with equal or near-equal slices and in multi-part configurations.
+
+In response to Eells' earlier work, @vonhuhnFurtherStudiesGraphic1927 ran his own large-scale experiment to see whether bar charts or pie charts made it easier for people to compare ratios between parts---like "A is to B as 1 is to what?" Instead of estimating percentages of a whole, participants were directly comparing how much bigger or smaller one part was than another.
+A total of 287 observers were shown two sets of matched charts: one set illustrated a 1:11 ratio, and another showed a 1:4 ratio, with each ratio presented once as a bar chart and once as a pie chart.
+No scales or labels were used, and the chart order was controlled to avoid bias.
+The key measure was accuracy---how close participants' estimates were to the actual ratios.
+Results showed that bar charts led to more correct judgments in both cases, especially with the more extreme 1:11 ratio.
+While pie charts performed slightly better in one metric, bar charts were clearly more consistent overall.
+Unlike Eells' percentage-based task, this was a direct comparison task (A vs. B), and the study focused purely on accuracy---not speed or preference.
+
 \svp{In which experiment did they have to only identify which part of the chart was larger??}
@@
 B + C), and two pairs (A + B vs.
-C + D).
-These tasks were tested across multiple formats (pie, bar, and table), with variations in exposure time, number of components, and use of color or ordering. 
+C + D).These tasks were tested across multiple formats (pie, bar, and table), with variations in exposure time, number of components, and use of color or ordering.
 Performance was evaluated in terms of speed and accuracy, allowing researchers to assess how perceptual and cognitive demands shift across visualization types and task difficulty.
 
-\svp{Add study to bibliography}
-Peterson and Schramm (1954) evaluated the accuracy of proportion estimation across eight common chart types used to represent parts of a whole: circle, disc, single bar, multiple bar, multiple cylinder, multiple square column, multiple area column, and partial cosmograph.
+@peterson1954accurately evaluated the accuracy of proportion estimation across eight common chart types used to represent parts of a whole: circle, disc, single bar, multiple bar, multiple cylinder, multiple square column, multiple area column, and partial cosmograph.
 Each chart displayed five fixed proportions (43%, 23%, 18%, 10%, 6%) and was presented alongside a similar "decoy" graph to mask repetition.
@@
 
-@clevelandGraphicalPerceptionTheory1984 [-@clevelandGraphicalPerceptionGraphical1985] conducted a series of experiments to test a theory of graphical perception based on "elementary perceptual tasks" such as position, length, angle, area, and shading.
+[@clevelandGraphicalPerceptionTheory1984; -@clevelandGraphicalPerceptionGraphical1985] conducted a series of experiments to test a theory of graphical perception based on "elementary perceptual tasks" such as position, length, angle, area, and shading.
 They hypothesized a hierarchy of these tasks based on how accurately people extract quantitative information from them.
-In their experiments, participants viewed a range of common graph types—including bar charts, pie charts, and divided bar charts—and were asked to make visual judgments of values or differences. 
+In their experiments, participants viewed a range of common graph types---including bar charts, pie charts, and divided bar charts---and were asked to make visual judgments of values or differences.
 The results supported their theory: judgments involving position along a common scale were the most accurate, followed by length, with angle (used in pie charts) yielding the least accurate responses.
@@
 
-@clevelandExperimentGraphicalPerception1986 conducted a \svp{follow-up} experiment to evaluate the accuracy of seven basic graphical perception tasks: position (aligned and non-aligned), length, angle, slope, and area (circle and blob).
+@clevelandExperimentGraphicalPerception1986 conducted a follow-up experiment to evaluate the accuracy of seven basic graphical perception tasks: position (aligned and non-aligned), length, angle, slope, and area (circle and blob).
 Participants were shown visual stimuli in which one object served as a standard, and three comparison objects were judged as percentages relative to it.
@@
 The study included 127 participants from diverse educational backgrounds.
-Results confirmed a perceptual hierarchy: position judgments were the most accurate, while angle and area were less reliable—insights that continue to influence data visualization design today. 
+Results confirmed a perceptual hierarchy: position judgments were the most accurate, while angle and area were less reliable---insights that continue to influence data visualization design today.
 \svp{Did the experiment actually cover all 7 tasks??? Please be careful with how you cite these, and provide more detail -- that's what we're specifically looking for here.}
 
-Simkin and Hastie (1987) investigated how different graphical encodings—position, length, and angle—affect performance on visual judgment tasks. 
-200 Undergraduate participants were shown pie charts, bar charts, and divided bar charts and asked to make either comparison judgments ($A/B$) or proportion-of-the-whole \svp{($A/(A+B)$)} judgements.
+```{=tex}
+\mt{They conducted a foundational study to assess how accurately people interpret different graphical encodings.
+They focused on six basic graphical perception tasks: position along a common scale, position along non-aligned scales, length, angle, slope, and area (represented by circles and irregular "blobs").
+Participants were shown 70 visual displays, each with one standard object and three comparisons, and asked to estimate each comparison as a percentage of the standard.
+The study involved 127 participants---high school students, college students, and professionals---who each made 210 judgments across all task types and standard sizes.
+Stimuli were randomized to avoid learning effects}
+```
+The main outcome was estimation accuracy---how close each response was to the true percentage.
+No scales or measurement tools were provided, so judgments relied purely on visual perception.
+Each of the six tasks was tested with both small and large reference standards, and values ranged from 17.5% to 87.5%.
+Results revealed a clear perceptual hierarchy: position along a common scale led to the most accurate estimates, followed by length.
+In contrast, angle and area judgments were significantly less accurate.
+These findings have had a lasting impact on visualization design, encouraging the use of perceptually stronger encodings for quantitative data.}
+
+@simkin1987information investigated how different graphical encodings---position, length, and angle---affect performance on visual judgment tasks.
+200 Undergraduate participants were shown pie charts, bar charts, and divided bar charts and asked to make either comparison judgments ($A/B$) or proportion-of-the-whole ($A/(A+B)$) judgements.
 Each graph type relied on a different visual code: bar charts used position, divided bars used length, and pie charts used angle.
@@
 
-\svp{Add study to bibliography}
-Very recently, more works have been discovered on this debate that catch researchers’ eyes on the topic. 
-Kosara & Skau (2016) investigated how common distortions in pie charts affect judgment accuracy.
+Very recently, more works have been discovered on this debate that catch researchers' eyes on the topic.
+@skau2016arcs investigated how common distortions in pie charts affect judgment accuracy.
 Participants were shown a series of pie chart variations and asked to estimate the percentage represented by a highlighted slice.
@@
 Participants were recruited via Mechanical Turk and asked to provide whole-number estimates for values ranging from 3% to 97%.
-Results showed that all distortions—except the exploded pie—led to significantly higher judgment errors compared to the baseline pie chart.
+Results showed that all distortions---except the exploded pie---led to significantly higher judgment errors compared to the baseline pie chart.
 
-In 2025, Hill did the study that involved three online experiments designed to evaluate how effectively bar, pie, and donut charts support part-whole estimation and ranking tasks. 
+```{=tex}
+\mt{Skau and Kosara (2016) conducted a focused experiment to isolate how individual visual encodings---arc length, angle, and area---contribute to interpretation accuracy in pie and donut charts.
+They created six chart types: baseline pie, baseline donut, arc-only, angle (pie-style), angle (donut-style), and area-only.
+Each chart showed just two segments---a blue "target" slice and a gray remainder---allowing participants to focus on estimating a single proportion without distractions.
+Across 48 randomized trials (eight per chart type), 100 Mechanical Turk participants were asked the same question for every chart: "What percentage of the whole is indicated below?" The charts were randomly rotated to avoid alignment biases, and a brief tutorial was provided to ensure participants understood each chart forma}
+```
+Measurement & Accuracy The study focused exclusively on perceptual accuracy.
+For each chart, researchers measured the absolute difference between the participant's estimate and the true value, as well as error patterns such as reversed judgments (e.g., answering 75% instead of 25% on angle charts).
+No time constraints were enforced, and response time or subjective preferences were not formally recorded.
+However, participants were asked before and after the main task which visual cue they believed they relied on---arc, angle, or area.
+Results revealed that baseline pie charts led to the most accurate judgments overall, followed by arc-only encodings.
+Surprisingly, many participants reported using angle judgments, even in chart types where angle cues were removed or degraded.}
+
+@hill2025bar did the study that involved three online experiments designed to evaluate how effectively bar, pie, and donut charts support part-whole estimation and ranking tasks.
 In Experiment 1, participants viewed randomized sets of charts (10 bar, 5 pie, 5 donut) created from multinomially generated data and answered ranking and proportion estimation questions.
@@
 $C + D$).
-\svp{Were participants estimating ratios, or just judging larger/smaller?}
-Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones. 
+\svp{Were participants estimating ratios, or just judging larger/smaller?} Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones.
 Bar charts performed slightly better than pie charts for simple comparisons under time pressure, but differences between chart types were negligible for complex tasks.
@@
 However, the lack of an ordering effect challenges the common belief that ordered components improve perception.
-\svp{How many categories were used?
-It's possible that this is true for small tables but not as true for larger tables.}
+\svp{How many categories were used? It's possible that this is true for small tables but not as true for larger tables.}
 
@@
 The small and homogeneous sample (mainly students) limits the generalizability of the findings.
-Short exposure times (1.0–6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.
+Short exposure times (1.0--6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.
 \svp{Why do psycologists use short exposure times?
-Look into this -- there's a reason}
-The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance. 
+Look into this -- there's a reason} The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance.
 Reducing exposure time to 1.0 second in Experiment 4 may have biased results toward pie charts by encouraging intuitive rather than analytical processing
@@
 
-@eellsRelativeMeritsCircles1926 didn’t just look at how well people performed—he also explored how they felt about the charts and how they approached the task. 
+@eellsRelativeMeritsCircles1926 didn't just look at how well people performed---he also explored how they felt about the charts and how they approached the task.
 When it came to accuracy, pie charts actually did better than bar charts, especially as the number of segments increased.
@@
 Was it significant?
-Be careful about uncritically reporting results.}
-One of the most striking findings was that bar charts led to 70% more large errors, making them less reliable for precise judgment.
+Be careful about uncritically reporting results.} One of the most striking findings was that bar charts led to 70% more large errors, making them less reliable for precise judgment.
 After the task, students were asked which chart type they preferred, and the majority chose pie charts for showing part-whole data.
-When it came to strategy, most students said they judged the slices using arc length, with others using angles or area—and all three methods turned out to be about equally accurate.
+When it came to strategy, most students said they judged the slices using arc length, with others using angles or area---and all three methods turned out to be about equally accurate.
 Overall, the results challenged the common criticism that pie charts are hard to read.
@@
 # Result
+
 Which chart works best really depends on the task.
-When people are asked to simply compare two values (A vs B), bar charts consistently come out on top—they’re faster and more accurate, and participants tend to prefer them for that reason (Spence & Lewandowsky, 1991; Simkin & Hastie, 1987; Hill, 2025).
+When people are asked to simply compare two values (A vs B), bar charts consistently come out on top---they're faster and more accurate, and participants tend to prefer them for that reason (@spenceDisplayingProportionsPercentages1991; @simkin1987information; @hill2025bar).
 When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (Cleveland & McGill, 1984; Simkin & Hastie, 1987).
 \svp{Is this still true when A and B aren't adjacent in e.g.
-a 3-4 component chart?}
+a 3-4 component chart?} \mt{While pie charts have been shown to support accurate part-to-whole estimations in two-part visualizations (e.g., A and B as adjacent slices), it is unclear whether this advantage persists in more complex charts.
+When pie charts include three or more segments, and the parts being compared (e.g., A and B) are not adjacent, the viewer must shift focus across non-contiguous regions of the circle.
+This added perceptual effort may diminish the intuitive advantage typically associated with pie charts in simpler tasks.
+Further studies are needed to assess whether the part-whole encoding remains effective when the target slices are visually separated—especially in 3- or 4-component pie charts where holistic comparison becomes less direct}
 
-For more complex comparisons—like $A$ vs $B+C$ or $A+B$ vs $C+D$, the results are more mixed. 
+For more complex comparisons---like $A$ vs $B+C$ or $A+B$ vs $C+D$, the results are more mixed.
 Bar charts have a slight edge when comparing a single value to a sum ($A$ vs $B+C$), but when comparing two grouped values ($A+B$ vs $C+D$), pie charts actually perform better (Spence & Lewandowsky, 1991).
@@
 
-In the end, there’s no universal winner—the best chart depends on the specific task and what kind of judgment you're asking the reader to make.
+In the end, there's no universal winner---the best chart depends on the specific task and what kind of judgment you're asking the reader to make.
 
-![Grouped Bar Chart](Grouped_Bar_Chart.png)
+```{r}
+#| echo: false
+
+library(ggplot2)
+
+# Data
+chart_data <- data.frame(
+  Chart_Type = rep(c("Bar", "Pie", "Donut"), each = 6),
+  Task_Type = rep(c("A vs B", "A / (A + B)", "A vs (B + C)", 
+                    "(A + B) vs (C + D)", "Ranking", "Magnitude Est."), 3),
+  Rank = c(
+    1, 3, 1, 2, 1, 3,  # Bar
+    2, 1, 2, 1, 2, 1,  # Pie
+    3, 2, 3, 3, 3, 1   # Donut
+  )
+)
+
+# Reorder x-axis: Pie, Bar, Donut
+chart_data$Chart_Type <- factor(chart_data$Chart_Type, levels = c("Pie", "Bar", "Donut"))
+
+# Set Rank so that 1 shows on top
+chart_data$Rank <- factor(chart_data$Rank, levels = c(3, 2, 1))
+
+# Plot
+ggplot(chart_data, aes(x = Chart_Type, y = Rank, fill = Task_Type)) +
+  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
+  labs(
+    title = "Chart Type Performance Across Tasks",
+    y = "Rank (1 = Best)",
+    x = NULL,
+    fill = "Task Type"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    legend.position = "right",
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 11),
+    plot.title = element_text(face = "bold", hjust = 0.5)
+  )
+
+```
+
+# Comparison table
+
+```{r}
+#| echo: false
+library(knitr)
+library(tibble)
+
+# Data with Study_Label and Preference column
+study_table <- tribble(
+  ~Study_Label, ~Study_Type,       ~Comparison, ~Question,                     ~Measure,                        ~Participants,          ~Trials_per_Participant,       ~Preference,
+  "A-exp1",     "Eells (1926)",    "A / (A+B+...)", "Estimate each segment",       "Accuracy, Preference",          "97 psychology students", "15 (Pie + Bar; no repeat)",   "Pie",
+  "A-exp2",     "Croxton (1927)",  "A / (A+B)", "Estimate percentage values",  "Accuracy",                      "807 students",           "27 (bar & pie mixed)",        "Pie",
+  "B-exp1",     "Von Huhn (1927)", "A / B",     "What is B if A is 1?",        "Accuracy",                      "287 participants",       "2 ratios × 2 chart types",    "Bar",
+  "B-exp2",     "Spence (1991)",   "A, A+B, etc.","Which part is largest?",   "Accuracy, Speed",               "Varied (online)",        "Not specified",               "Bar",
+  "C-exp1",     "Petersen (1954)", "A / (A+B)", "Estimate proportion",         "Accuracy",                      "86 air force personnel", "3440 responses total",        "Bar",
+  "D-exp1",     "Cleveland (1986)","A / (A+B)", "What % is this of standard?", "Accuracy",                      "127 (HS, college, tech)", "210 (6 task types)",          "Bar",
+  "E-exp1",     "Simkin & Hastie (1987)", "A / B & A / (A+B)", "Compare or estimate", "Accuracy, Time",       "200 undergraduates",      "90 (timed)",                  "Pie (for A/(A+B))",
+  "F-exp1",     "Skau & Kosara (2016)", "A / (A+B)", "What % is the blue part?", "Accuracy",                   "100 MTurk",              "48 (6 chart types × 8)",      "Pie",
+  "G-exp1",     "Hill (2025)",     "Ranking, A vs B, A+B", "Rank or estimate", "Accuracy",                     "42 Univ. of Surrey",     "3 experiments",               "Bar"
+)
+
+# Display table
+kable(
+  study_table[, c("Study_Label", "Study_Type", "Comparison", "Question", "Measure", "Participants", "Trials_per_Participant", "Preference")],
+  col.names = c("Label", "Citation", "Comparison", "Question", "Measure", "Participants", "Trials/Participant", "Preference"),
+  caption = "Summary of Experimental Studies on Chart Perception"
+)
+
+```
+
+## Pie Chart
+
+```{r}
+#| echo: false
+
+library(ggplot2)
+library(dplyr)
+
+# Data for 4 segments
+df1 <- data.frame(
+  group = c("A", "B", "C", "D"),
+  value = c(25, 25, 25, 25)
+)
+
+ggplot(df1, aes(x = "", y = value, fill = group)) +
+  geom_bar(width = 1, stat = "identity", color = "white") +
+  coord_polar("y") +
+  theme_void() +
+  ggtitle("Comparison: A, B vs C, D") +
+  theme(
+    legend.title = element_blank(),
+    plot.title = element_text(hjust = 0.5, size = 14)
+  )
+
+# Data for two pie charts
+df2 <- data.frame(
+  group = rep(c("A+B", "C+D"), each = 2),
+  sub_group = rep(c("A", "B", "C", "D"), 1),
+  value = c(20, 30, 25, 25)
+)
+
+df2$facet <- rep(c("Group AB", "Group CD"), each = 2)
+
+ggplot(df2, aes(x = "", y = value, fill = sub_group)) +
+  geom_bar(stat = "identity", width = 1, color = "white") +
+  coord_polar("y") +
+  facet_wrap(~ facet) +
+  theme_void() +
+  ggtitle("Comparison: AB vs CD") +
+  theme(
+    strip.text = element_text(size = 12),
+    legend.title = element_blank(),
+    plot.title = element_text(hjust = 0.5, size = 14)
+  )
+
+```
+
+```{r}
+#| echo: False
+library(ggplot2)
+library(dplyr)
+
+# Helper function to create pie chart data
+make_pie_data <- function(labels, values, group) {
+  df <- data.frame(
+    label = labels,
+    value = values
+  )
+  df$group <- group
+  df
+}
+
+# Create all pie segments
+pie_data <- bind_rows(
+  make_pie_data(c("A", "B"), c(0.4, 0.6), "A vs B"),
+  make_pie_data(c("A", "B", "C"), c(0.25, 0.25, 0.5), "A vs (B+C)"),
+  make_pie_data(c("A+B", "C"), c(0.5, 0.5), "(A+B) vs C")
+)
+
+# Prepare for pie chart layout
+pie_data <- pie_data %>%
+  group_by(group) %>%
+  mutate(
+    fraction = value / sum(value),
+    ymax = cumsum(fraction),
+    ymin = lag(ymax, default = 0),
+    label_pos = (ymin + ymax) / 2
+  )
+
+# Plot
+ggplot(pie_data, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = label)) +
+  geom_rect(color = "white") +
+  coord_polar(theta = "y") +
+  facet_wrap(~group, ncol = 1, strip.position = "left") +
+  geom_text(aes(x = 0.5, y = label_pos, label = label), size = 5) +
+  scale_fill_manual(values = c("A" = "#e0f3db", "B" = "#ccebc5", "C" = "#b2df8a", "A+B" = "#a6cee3")) +
+  theme_void() +
+  theme(
+    strip.text = element_text(size = 14, face = "bold"),
+    legend.position = "none",
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
+  ) +
+  ggtitle("Comparison Tasks")
+
+# ==== 1. Full Pie Chart: A, B, C, D ====
+df_full <- data.frame(
+  group = c("A", "B", "C", "D"),
+  value = c(25, 25, 25, 25)
+)
+
+df_full <- df_full %>%
+  mutate(
+    fraction = value / sum(value),
+    ymax = cumsum(fraction),
+    ymin = c(0, head(ymax, n = -1)),
+    label_pos = (ymin + ymax) / 2
+  )
+
+ggplot(df_full, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
+  geom_rect(color = "white") +
+  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
+  coord_polar(theta = "y") +
+  scale_fill_manual(values = gray.colors(4, start = 0.9, end = 0.4)) +
+  theme_void() +
+  ggtitle("Full Pie Chart: A, B, C, D")
+
+```
+
+```{r}
+#| echo: false
+library(ggplot2)
+library(dplyr)
+
+# Helper function to create pie chart data
+make_pie_data <- function(labels, values, group) {
+  df <- data.frame(
+    label = labels,
+    value = values
+  )
+  df$group <- group
+  df
+}
+
+# Combine all pie data
+pie_data <- bind_rows(
+  make_pie_data(c("A", "B"), c(0.4, 0.6), "A vs B"),
+  make_pie_data(c("A", "B", "C"), c(0.25, 0.25, 0.5), "A vs (B+C)"),
+  make_pie_data(c("A+B", "C"), c(0.5, 0.5), "(A+B) vs C")
+)
+
+# Compute positions
+pie_data <- pie_data %>%
+  group_by(group) %>%
+  mutate(
+    fraction = value / sum(value),
+    ymax = cumsum(fraction),
+    ymin = lag(ymax, default = 0),
+    label_pos = (ymin + ymax) / 2
+  )
+
+# Color palette (match original soft hues)
+color_palette <- c("A" = "#FBB4AE", "B" = "#B3CDE3", "C" = "#DECBE4", "A+B" = "#CCEBC5")
+
+# Plot
+ggplot(pie_data, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = label)) +
+  geom_rect(color = "white") +
+  geom_text(aes(x = 0.5, y = label_pos, label = label), size = 5) +  # Labels inside
+  coord_polar(theta = "y") +
+  facet_wrap(~group, nrow = 1) +  # Horizontal layout
+  scale_fill_manual(values = color_palette) +
+  theme_void() +
+  theme(
+    strip.text = element_text(size = 14, face = "bold"),
+    legend.position = "none"
+  )
+
+```
+
+```{r}
+#| echo: false
+# Load required libraries
+library(ggplot2)
+library(dplyr)
+
+# ==== 1. Full Pie Chart: A, B, C, D ====
+df_full <- data.frame(
+  group = c("A", "B", "C", "D"),
+  value = c(25, 25, 25, 25)
+)
+
+df_full <- df_full %>%
+  mutate(
+    fraction = value / sum(value),
+    ymax = cumsum(fraction),
+    ymin = c(0, head(ymax, n = -1)),
+    label_pos = (ymin + ymax) / 2
+  )
+
+ggplot(df_full, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
+  geom_rect(color = "white") +
+  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
+  coord_polar(theta = "y") +
+  scale_fill_manual(values = gray.colors(4, start = 0.9, end = 0.4)) +
+  theme_void() +
+  ggtitle("Full Pie Chart: A, B, C, D")
+
+# ==== 2. Half Pie Chart: Group AB ====
+df_ab <- data.frame(
+  group = c("A", "B"),
+  value = c(50, 50)
+)
+
+df_ab <- df_ab %>%
+  mutate(
+    fraction = value / sum(value),
+    ymax = cumsum(fraction),
+    ymin = c(0, head(ymax, n = -1)),
+    label_pos = (ymin + ymax) / 2
+  )
+
+p_ab <- ggplot(df_ab, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
+  geom_rect(color = "white") +
+  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
+  coord_polar(theta = "y", start = pi) +  # start at top to show half-pie
+  scale_fill_manual(values = gray.colors(2, start = 0.9, end = 0.7)) +
+  theme_void() +
+  ggtitle("Group AB")
+
+# ==== 3. Half Pie Chart: Group CD ====
+df_cd <- data.frame(
+  group = c("C", "D"),
+  value = c(50, 50)
+)
+
+df_cd <- df_cd %>%
+  mutate(
+    fraction = value / sum(value),
+    ymax = cumsum(fraction),
+    ymin = c(0, head(ymax, n = -1)),
+    label_pos = (ymin + ymax) / 2
+  )
+
+p_cd <- ggplot(df_cd, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
+  geom_rect(color = "white") +
+  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
+  coord_polar(theta = "y", start = pi) +
+  scale_fill_manual(values = gray.colors(2, start = 0.6, end = 0.4)) +
+  theme_void() +
+  ggtitle("Group CD")
+
+# ==== Combine AB and CD side by side ====
+library(patchwork)
+p_ab + p_cd
+```
+
+```{r}
+# Load libraries
+library(ggplot2)
+library(dplyr)
+library(patchwork)
+
+# Function to make pie chart with highlighted segments
+make_pie_plot <- function(data, title, highlight_group = NULL, highlight_color = "black") {
+  data <- data %>%
+    mutate(
+      fraction = value / sum(value),
+      ymax = cumsum(fraction),
+      ymin = lag(ymax, default = 0),
+      label_pos = (ymin + ymax) / 2,
+      label = label
+    )
+
+  # Add border flag
+  data$highlight <- ifelse(data$label %in% highlight_group, TRUE, FALSE)
+
+  ggplot(data) +
+    geom_rect(aes(
+      ymin = ymin, ymax = ymax, xmin = 0.5, xmax = 1.5,
+      fill = label, color = highlight
+    ), size = 1) +
+    coord_polar(theta = "y") +
+    geom_text(aes(
+      x = 1, y = label_pos, label = label
+    ), size = 5) +
+    scale_fill_manual(values = c(
+      "A" = "#f4a582", "B" = "#92c5de", "C" = "#d1bce3", "A+B" = "#b2df8a", "B+C" = "#b2abd2"
+    )) +
+    scale_color_manual(values = c("TRUE" = highlight_color, "FALSE" = NA), guide = "none") +
+    theme_void() +
+    ggtitle(title) +
+    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
+}
+
+# Define the three datasets
+data1 <- tibble(label = c("A+B", "C"), value = c(0.5, 0.5))  # (A+B) vs C
+data2 <- tibble(label = c("A", "B", "C"), value = c(0.25, 0.25, 0.5))  # A vs (B+C)
+data3 <- tibble(label = c("A", "B"), value = c(0.4, 0.6))  # A vs B
+
+# Make each plot with outlines
+p1 <- make_pie_plot(data1, "(A+B) vs C", highlight_group = "A+B")
+p2 <- make_pie_plot(data2, "A vs (B+C)", highlight_group = c("B", "C"))
+p3 <- make_pie_plot(data3, "A vs B", highlight_group = c("A", "B"))
+
+# Combine the three plots horizontally
+(p1 | p2 | p3)
+
+```
+
+## Summary: When to Use Each Chart Type
+
+| **Use**              | **When**                                                                      
+|----------------------|-------------------------------------------------------------------------------|
+| **Bar charts**       | Comparing individual parts (A vs B), ranking, or scanning many values         |
+| **Pie charts**       | Estimating proportions of a whole ( A / (A + B)) with few components          |
+| **Avoid pie charts** | When comparing non-adjacent slices or judging small differences               |
+
+# Conclusion
 
@@
 
-# References
+# Reference

### 2025-04-16 Draft.qmd
Added conclusion and correction

@@
 
-```{=tex}
 \mt{They conducted a foundational study to assess how accurately people interpret different graphical encodings.
@@
 Stimuli were randomized to avoid learning effects}
-```
+
 The main outcome was estimation accuracy---how close each response was to the true percentage.
@@
 
-```{=tex}
 \mt{Skau and Kosara (2016) conducted a focused experiment to isolate how individual visual encodings---arc length, angle, and area---contribute to interpretation accuracy in pie and donut charts.
@@
 Across 48 randomized trials (eight per chart type), 100 Mechanical Turk participants were asked the same question for every chart: "What percentage of the whole is indicated below?" The charts were randomly rotated to avoid alignment biases, and a brief tutorial was provided to ensure participants understood each chart forma}
-```
+
 Measurement & Accuracy The study focused exclusively on perceptual accuracy.
@@
 
-```{r}
-#| echo: false
-
-library(ggplot2)
-
-# Data
-chart_data <- data.frame(
-  Chart_Type = rep(c("Bar", "Pie", "Donut"), each = 6),
-  Task_Type = rep(c("A vs B", "A / (A + B)", "A vs (B + C)", 
-                    "(A + B) vs (C + D)", "Ranking", "Magnitude Est."), 3),
-  Rank = c(
-    1, 3, 1, 2, 1, 3,  # Bar
-    2, 1, 2, 1, 2, 1,  # Pie
-    3, 2, 3, 3, 3, 1   # Donut
-  )
-)
-
-# Reorder x-axis: Pie, Bar, Donut
-chart_data$Chart_Type <- factor(chart_data$Chart_Type, levels = c("Pie", "Bar", "Donut"))
-
-# Set Rank so that 1 shows on top
-chart_data$Rank <- factor(chart_data$Rank, levels = c(3, 2, 1))
-
-# Plot
-ggplot(chart_data, aes(x = Chart_Type, y = Rank, fill = Task_Type)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
-  labs(
-    title = "Chart Type Performance Across Tasks",
-    y = "Rank (1 = Best)",
-    x = NULL,
-    fill = "Task Type"
-  ) +
-  theme_minimal(base_size = 14) +
-  theme(
-    legend.position = "right",
-    legend.title = element_text(size = 12),
-    legend.text = element_text(size = 11),
-    plot.title = element_text(face = "bold", hjust = 0.5)
-  )
-
-```
-
 # Comparison table
@@
 ```{r}
-#| echo: false
-
-library(ggplot2)
-library(dplyr)
-
-# Data for 4 segments
-df1 <- data.frame(
-  group = c("A", "B", "C", "D"),
-  value = c(25, 25, 25, 25)
-)
-
-ggplot(df1, aes(x = "", y = value, fill = group)) +
-  geom_bar(width = 1, stat = "identity", color = "white") +
-  coord_polar("y") +
-  theme_void() +
-  ggtitle("Comparison: A, B vs C, D") +
-  theme(
-    legend.title = element_blank(),
-    plot.title = element_text(hjust = 0.5, size = 14)
-  )
-
-# Data for two pie charts
-df2 <- data.frame(
-  group = rep(c("A+B", "C+D"), each = 2),
-  sub_group = rep(c("A", "B", "C", "D"), 1),
-  value = c(20, 30, 25, 25)
-)
-
-df2$facet <- rep(c("Group AB", "Group CD"), each = 2)
-
-ggplot(df2, aes(x = "", y = value, fill = sub_group)) +
-  geom_bar(stat = "identity", width = 1, color = "white") +
-  coord_polar("y") +
-  facet_wrap(~ facet) +
-  theme_void() +
-  ggtitle("Comparison: AB vs CD") +
-  theme(
-    strip.text = element_text(size = 12),
-    legend.title = element_blank(),
-    plot.title = element_text(hjust = 0.5, size = 14)
-  )
-
-```
-
-```{r}
-#| echo: False
-library(ggplot2)
-library(dplyr)
-
-# Helper function to create pie chart data
-make_pie_data <- function(labels, values, group) {
-  df <- data.frame(
-    label = labels,
-    value = values
-  )
-  df$group <- group
-  df
-}
-
-# Create all pie segments
-pie_data <- bind_rows(
-  make_pie_data(c("A", "B"), c(0.4, 0.6), "A vs B"),
-  make_pie_data(c("A", "B", "C"), c(0.25, 0.25, 0.5), "A vs (B+C)"),
-  make_pie_data(c("A+B", "C"), c(0.5, 0.5), "(A+B) vs C")
-)
-
-# Prepare for pie chart layout
-pie_data <- pie_data %>%
-  group_by(group) %>%
-  mutate(
-    fraction = value / sum(value),
-    ymax = cumsum(fraction),
-    ymin = lag(ymax, default = 0),
-    label_pos = (ymin + ymax) / 2
-  )
-
-# Plot
-ggplot(pie_data, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = label)) +
-  geom_rect(color = "white") +
-  coord_polar(theta = "y") +
-  facet_wrap(~group, ncol = 1, strip.position = "left") +
-  geom_text(aes(x = 0.5, y = label_pos, label = label), size = 5) +
-  scale_fill_manual(values = c("A" = "#e0f3db", "B" = "#ccebc5", "C" = "#b2df8a", "A+B" = "#a6cee3")) +
-  theme_void() +
-  theme(
-    strip.text = element_text(size = 14, face = "bold"),
-    legend.position = "none",
-    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
-  ) +
-  ggtitle("Comparison Tasks")
-
-# ==== 1. Full Pie Chart: A, B, C, D ====
-df_full <- data.frame(
-  group = c("A", "B", "C", "D"),
-  value = c(25, 25, 25, 25)
-)
-
-df_full <- df_full %>%
-  mutate(
-    fraction = value / sum(value),
-    ymax = cumsum(fraction),
-    ymin = c(0, head(ymax, n = -1)),
-    label_pos = (ymin + ymax) / 2
-  )
-
-ggplot(df_full, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
-  geom_rect(color = "white") +
-  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
-  coord_polar(theta = "y") +
-  scale_fill_manual(values = gray.colors(4, start = 0.9, end = 0.4)) +
-  theme_void() +
-  ggtitle("Full Pie Chart: A, B, C, D")
-
-```
-
-```{r}
-#| echo: false
-library(ggplot2)
-library(dplyr)
-
-# Helper function to create pie chart data
-make_pie_data <- function(labels, values, group) {
-  df <- data.frame(
-    label = labels,
-    value = values
-  )
-  df$group <- group
-  df
-}
-
-# Combine all pie data
-pie_data <- bind_rows(
-  make_pie_data(c("A", "B"), c(0.4, 0.6), "A vs B"),
-  make_pie_data(c("A", "B", "C"), c(0.25, 0.25, 0.5), "A vs (B+C)"),
-  make_pie_data(c("A+B", "C"), c(0.5, 0.5), "(A+B) vs C")
-)
-
-# Compute positions
-pie_data <- pie_data %>%
-  group_by(group) %>%
-  mutate(
-    fraction = value / sum(value),
-    ymax = cumsum(fraction),
-    ymin = lag(ymax, default = 0),
-    label_pos = (ymin + ymax) / 2
-  )
-
-# Color palette (match original soft hues)
-color_palette <- c("A" = "#FBB4AE", "B" = "#B3CDE3", "C" = "#DECBE4", "A+B" = "#CCEBC5")
-
-# Plot
-ggplot(pie_data, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = label)) +
-  geom_rect(color = "white") +
-  geom_text(aes(x = 0.5, y = label_pos, label = label), size = 5) +  # Labels inside
-  coord_polar(theta = "y") +
-  facet_wrap(~group, nrow = 1) +  # Horizontal layout
-  scale_fill_manual(values = color_palette) +
-  theme_void() +
-  theme(
-    strip.text = element_text(size = 14, face = "bold"),
-    legend.position = "none"
-  )
-
-```
-
-```{r}
-#| echo: false
-# Load required libraries
-library(ggplot2)
-library(dplyr)
-
-# ==== 1. Full Pie Chart: A, B, C, D ====
-df_full <- data.frame(
-  group = c("A", "B", "C", "D"),
-  value = c(25, 25, 25, 25)
-)
-
-df_full <- df_full %>%
-  mutate(
-    fraction = value / sum(value),
-    ymax = cumsum(fraction),
-    ymin = c(0, head(ymax, n = -1)),
-    label_pos = (ymin + ymax) / 2
-  )
-
-ggplot(df_full, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
-  geom_rect(color = "white") +
-  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
-  coord_polar(theta = "y") +
-  scale_fill_manual(values = gray.colors(4, start = 0.9, end = 0.4)) +
-  theme_void() +
-  ggtitle("Full Pie Chart: A, B, C, D")
-
-# ==== 2. Half Pie Chart: Group AB ====
-df_ab <- data.frame(
-  group = c("A", "B"),
-  value = c(50, 50)
-)
-
-df_ab <- df_ab %>%
-  mutate(
-    fraction = value / sum(value),
-    ymax = cumsum(fraction),
-    ymin = c(0, head(ymax, n = -1)),
-    label_pos = (ymin + ymax) / 2
-  )
-
-p_ab <- ggplot(df_ab, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
-  geom_rect(color = "white") +
-  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
-  coord_polar(theta = "y", start = pi) +  # start at top to show half-pie
-  scale_fill_manual(values = gray.colors(2, start = 0.9, end = 0.7)) +
-  theme_void() +
-  ggtitle("Group AB")
-
-# ==== 3. Half Pie Chart: Group CD ====
-df_cd <- data.frame(
-  group = c("C", "D"),
-  value = c(50, 50)
-)
-
-df_cd <- df_cd %>%
-  mutate(
-    fraction = value / sum(value),
-    ymax = cumsum(fraction),
-    ymin = c(0, head(ymax, n = -1)),
-    label_pos = (ymin + ymax) / 2
-  )
-
-p_cd <- ggplot(df_cd, aes(ymax = ymax, ymin = ymin, xmax = 1, xmin = 0, fill = group)) +
-  geom_rect(color = "white") +
-  geom_text(aes(x = 0.5, y = label_pos, label = group), size = 6, fontface = "bold") +
-  coord_polar(theta = "y", start = pi) +
-  scale_fill_manual(values = gray.colors(2, start = 0.6, end = 0.4)) +
-  theme_void() +
-  ggtitle("Group CD")
-
-# ==== Combine AB and CD side by side ====
-library(patchwork)
-p_ab + p_cd
-```
-
-```{r}
-# Load libraries
+# | echo: false
 library(ggplot2)
@@
 
-# Function to make pie chart with highlighted segments
-make_pie_plot <- function(data, title, highlight_group = NULL, highlight_color = "black") {
+# Helper function for pie chart
+create_pie <- function(data, fill_colors, title, bold_groups = NULL) {
   data <- data %>%
@@
       label_pos = (ymin + ymax) / 2,
-      label = label
+      border_size = ifelse(label %in% bold_groups, 3, 0.8)
     )
 
-  # Add border flag
-  data$highlight <- ifelse(data$label %in% highlight_group, TRUE, FALSE)
-
   ggplot(data) +
-    geom_rect(aes(
-      ymin = ymin, ymax = ymax, xmin = 0.5, xmax = 1.5,
-      fill = label, color = highlight
-    ), size = 1) +
+    geom_rect(aes(ymin = ymin, ymax = ymax, xmin = 0.5, xmax = 1.5,
+                  fill = label, size = border_size),
+              color = "black", show.legend = FALSE) +
+    geom_text(aes(x = 1, y = label_pos, label = label), size = 6) +
+    scale_fill_manual(values = fill_colors) +
+    scale_size_identity() +
     coord_polar(theta = "y") +
-    geom_text(aes(
-      x = 1, y = label_pos, label = label
-    ), size = 5) +
-    scale_fill_manual(values = c(
-      "A" = "#f4a582", "B" = "#92c5de", "C" = "#d1bce3", "A+B" = "#b2df8a", "B+C" = "#b2abd2"
-    )) +
-    scale_color_manual(values = c("TRUE" = highlight_color, "FALSE" = NA), guide = "none") +
+    labs(title = title) +
     theme_void() +
-    ggtitle(title) +
-    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
+    theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))
 }
 
-# Define the three datasets
-data1 <- tibble(label = c("A+B", "C"), value = c(0.5, 0.5))  # (A+B) vs C
-data2 <- tibble(label = c("A", "B", "C"), value = c(0.25, 0.25, 0.5))  # A vs (B+C)
-data3 <- tibble(label = c("A", "B"), value = c(0.4, 0.6))  # A vs B
+# Pie 1: A vs B
+data1 <- tibble(label = c("B", "A"), value = c(0.35, 0.65))
+fill1 <- c("A" = "#f4a582", "B" = "#92c5de")
+p1 <- create_pie(data1, fill1, "A vs B")
 
-# Make each plot with outlines
-p1 <- make_pie_plot(data1, "(A+B) vs C", highlight_group = "A+B")
-p2 <- make_pie_plot(data2, "A vs (B+C)", highlight_group = c("B", "C"))
-p3 <- make_pie_plot(data3, "A vs B", highlight_group = c("A", "B"))
+# Pie 2: (A+B) vs C
+data2 <- tibble(label = c("C", "B", "A"), value = c(0.25, 0.25, 0.5))
+fill2 <- c("A" = "white", "B" = "#ff00ff", "C" = "#00ffff")
+p2 <- create_pie(data2, fill2, "(A+B) vs C", bold_groups = c("A", "B"))
 
-# Combine the three plots horizontally
+# Pie 3: (A+B) vs (C+D)
+data3 <- tibble(label = c("C", "B", "A", "D"), value = c(0.25, 0.20, 0.35, 0.20))
+fill3 <- c("A" = "white", "B" = "white", "C" = "#00ffff", "D" = "pink")
+p3 <- create_pie(data3, fill3, "(A+B) vs (C+D)", bold_groups = c("A", "B"))
+
+# Combine with patchwork
 (p1 | p2 | p3)
@@
 # Conclusion
+As of now, no one has determined a definitive winner between pie charts and bar charts.
+These two charts remain essential in every part of science, business, sociology, and related fields.
+The efficacy of various charts is contingent upon ratio, measurement, and perception aspects of the representation.
+In research, a chart should not be dismissed solely because it is deemed inadequate.
+Pie charts make it easier to assess proportions relative to the whole, but bar charts excel in comparing magnitudes and ranking activities.
+Performance fluctuates depending on whether the task entails value estimation, segment comparison, or identification of the major component.
+Generally, charts that clearly delineate comparative parts (e.g., adjacent slices or bold outlines) enhance interpretability.
+
+We have also discovered an intriguing fact that, despite the evolution of formats and digital interfaces, fundamental perceptual concepts (@eellsRelativeMeritsCircles1926, @clevelandGraphicalPerceptionTheory1984 and so on) continue to have validity.
+Nonetheless, this finding does not signify the conclusion of the pursuit for the optimal chart.
+Future research can clarify the dispute through improved experimental design and more compelling methodologies.
 
@@
 
+Numerous constraints are present in both early and contemporary investigations.
+Initial trials (from the 1920s to 1980s) frequently exhibited insufficient statistical rigor, characterized by delicate measurement frameworks and rudimentary designs.
+Contemporary research (@spenceDisplayingProportionsPercentages1991; @hill2025bar) has implemented enhanced experimental designs, however it continues to encounter difficulties, especially with participant diversity.
+
+Contemporary research predominantly utilizes student samples, so constraining generalizability and potentially failing to represent wider perceptual disparities across age, education, or cultural contexts.
+Subsequent investigations ought to tackle this issue by enhancing sample variety and refining experimental methodologies.
+This may reveal fresh insights and promote more inclusive, contextually suitable chart type.
+
+ Individuals' perceptions of charts may alter between devices—mobile phones, tablets, or desktop monitors—due to variations in screen size, resolution, and interaction methods.
+Furthermore, enhancing the reliability of perceptual measurements is crucial.
+Future research may include technologies such as eye-tracking, response time analysis, or cognitive load evaluations to provide more nuanced findings.
+The objective is to convert experimental results into practical design advice that assist practitioners in choosing the most suitable chart kinds for certain audiences and jobs.
+
 # Reference

### 2025-04-22 Draft.qmd
A few comments

@@
 @eellsRelativeMeritsCircles1926 also quotes an excerpt from an introductory statistics textbook [@secristIntroductionStatisticalMethods1925], which argues that pie charts require more cognitive effort to use accurately than a comparable stacked bar chart, calling pie diagrams "clumsy and defective".
-These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use
+These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use.
 
 @wickhamGraphicalCriticismHistorical2013 previously noted the prescience of design guidelines prior to empirical studies validating their recommendations across different types of visualizations.
+[Ok, can you connect this better to the previous paragraph somehow?]{.svp}
 
@@
 The phrasing of direct estimation questions is of primary importance: a simple mathematical transformation of the estimated quantity can have an outsized effect on the results. 
+[You might want to demonstrate this with an example, perhaps showing the phrasing of Croxton's study vs. Eells? You could even include screenshots from the papers.]{.svp}
 
@@
 Experimental graphics studies have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.
+[However, guidance to favor bar charts over pie charts still shows up in statistical literature -- cite @fienbergGraphicalMethodsStatistics1979, @haemerPseudoThirdDimension1951, @haemerPresentationProblemsSuiting1949, @nisselson1978graphic.
+It almost seems like approximately once every 20 years people think about graphics, decry the state of the field, and call for a better set of empirical guidelines!]{.svp}
 
@@
 
-![Different elicitation methods used to assess accuracy of pie chart estimates.](pie-type-example.png)
+<!-- ![Different elicitation methods used to assess accuracy of pie chart estimates.](pie-type-example.png) -->
 
@@
 
-\mt{Eells (1926) did not ask participants to simply identify which segment was larger.
+[Eells (1926) did not ask participants to simply identify which segment was larger.
 Instead, it was a proportion estimation task—participants were asked to write down the percentage each segment represented of the whole ( A / (A + B + ...)), not to make a binary comparison (A vs B).
-The “which is larger” comparison task appears in later studies, such as @croxtonBarChartsCircle1927 and @spenceDisplayingProportionsPercentages1991}
+The “which is larger” comparison task appears in later studies, such as @croxtonBarChartsCircle1927 and @spenceDisplayingProportionsPercentages1991]{.mt}
 
@@
 
-\svp{In which experiment did they have to only identify which part of the chart was larger??}
+[In which experiment did they have to only identify which part of the chart was larger??]{.svp}
 
@@
 Results confirmed a perceptual hierarchy: position judgments were the most accurate, while angle and area were less reliable---insights that continue to influence data visualization design today.
-\svp{Did the experiment actually cover all 7 tasks??? Please be careful with how you cite these, and provide more detail -- that's what we're specifically looking for here.}
+[Did the experiment actually cover all 7 tasks??? Please be careful with how you cite these, and provide more detail -- that's what we're specifically looking for here.]{.svp}
 
@@
 $C + D$).
-\svp{Were participants estimating ratios, or just judging larger/smaller?} Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones.
+[Were participants estimating ratios, or just judging larger/smaller?]{.svp} Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones.
 Bar charts performed slightly better than pie charts for simple comparisons under time pressure, but differences between chart types were negligible for complex tasks.
@@
 However, the lack of an ordering effect challenges the common belief that ordered components improve perception.
-\svp{How many categories were used? It's possible that this is true for small tables but not as true for larger tables.}
+[How many categories were used? It's possible that this is true for small tables but not as true for larger tables.]{.svp}
 
@@
 Unlike Experiment 1, which used a between-subjects design, and Experiment 2, which introduced tables and fixed processing time, Experiment 3 retained pie charts, bar charts, and tables while excluding the horizontal divided bar chart.
-Processing time was fixed at 3.0 seconds \svp{You can fix stimulus time but not processing time -- processing takes as long as it takes}, and two levels of components (4 and 7) were tested.
+Processing time was fixed at 3.0 seconds [You can fix stimulus time but not processing time -- processing takes as long as it takes]{.svp}, and two levels of components (4 and 7) were tested.
 Accuracy declined with task complexity, and while pie and bar charts showed similar performance for simpler tasks, pie charts outperformed the others for complex tasks, likely due to easier component identification and combination.
@@
 Short exposure times (1.0--6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.
-\svp{Why do psycologists use short exposure times?
-Look into this -- there's a reason} The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance.
+[Why do psycologists use short exposure times?
+Look into this -- there's a reason]{.svp} The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance.
 Reducing exposure time to 1.0 second in Experiment 4 may have biased results toward pie charts by encouraging intuitive rather than analytical processing
@@
 People also tended to move a bit faster with pie charts, completing slightly more within five minutes.
-\svp{Eells didn't do the hypothesis test -- can you from provided information?
+[Eells didn't do the hypothesis test -- can you from provided information?
 Was it significant?
-Be careful about uncritically reporting results.} One of the most striking findings was that bar charts led to 70% more large errors, making them less reliable for precise judgment.
+Be careful about uncritically reporting results.]{.svp} One of the most striking findings was that bar charts led to 70% more large errors, making them less reliable for precise judgment.
 After the task, students were asked which chart type they preferred, and the majority chose pie charts for showing part-whole data.
@@
 When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (Cleveland & McGill, 1984; Simkin & Hastie, 1987).
-\svp{Is this still true when A and B aren't adjacent in e.g.
-a 3-4 component chart?} \mt{While pie charts have been shown to support accurate part-to-whole estimations in two-part visualizations (e.g., A and B as adjacent slices), it is unclear whether this advantage persists in more complex charts.
+[Is this still true when A and B aren't adjacent in e.g.
+a 3-4 component chart?]{.svp} [While pie charts have been shown to support accurate part-to-whole estimations in two-part visualizations (e.g., A and B as adjacent slices), it is unclear whether this advantage persists in more complex charts.
 When pie charts include three or more segments, and the parts being compared (e.g., A and B) are not adjacent, the viewer must shift focus across non-contiguous regions of the circle.
 This added perceptual effort may diminish the intuitive advantage typically associated with pie charts in simpler tasks.
-Further studies are needed to assess whether the part-whole encoding remains effective when the target slices are visually separated—especially in 3- or 4-component pie charts where holistic comparison becomes less direct}
+Further studies are needed to assess whether the part-whole encoding remains effective when the target slices are visually separated—especially in 3- or 4-component pie charts where holistic comparison becomes less direct]{.mt}
 

### 2025-06-10 Draft.qmd
New (Added Motivation part)

--- 2025-04-16 Draft.qmd
@@
 
+It's really hard to reach in a conclusion that which one is better in just a one liner.
+We will review previous work on this topic, analyze their experimental designs, compare different scenarios, and attempt to generalize findings based on these studies.
+
+# Motivation
+
+Choosing the most effective chart type is crucial for accurately conveying quantitative information.
+Decades of research in graphical perception have shown that bar charts outperform pie charts in terms of accuracy, efficiency, and ease of interpretation @clevelandGraphicalPerceptionTheory1984; @spenceDisplayingProportionsPercentages1991.
+Despite this, pie charts remain prevalent in media, business, and education.
+Understanding why this mismatch persists — and how preferences or traditions shape visualization choices — is important for improving data literacy and promoting evidence-based design @skau2016arcs; @few2007pies.
+This study revisits 100 years of pie and bar chart usage to uncover trends in effectiveness, preference, and pedagogical shifts.
+
 # The Dawn of Empirical Graphics
@@
 Three days later, the same students were given an equivalent set of bar charts showing the same proportions, and were again asked to estimate the percentages for each section.
-The design intentionally favored the bars (e.g., common alignment of bar bases) to test whether any advantage of the pie chart could still be observed.
+The design intentionally favored the bars (common alignment of bar bases) to test whether any advantage of the pie chart could still be observed.
 After completing the estimation tasks, participants reported the mental strategy they used (area, angle, arc length, or chord), and indicated which chart type they preferred.
@@
 Each chart was shown individually (without any numerical scale), and participants were instructed to numerically estimate the percentage value of each segment---not simply identify which was larger.
-This distinction is important that like @eellsRelativeMeritsCircles1926 , this was a proportion estimation task ( "What percent is this segment?"), not a comparison task ("Which is larger, A\>B?").
+This distinction is important that like @eellsRelativeMeritsCircles1926 , this was a proportion estimation task ( "What percent is this segment?"), not a comparison task ("Which is larger, A>B?").
 Tasks requiring direct identification of the larger part appear in later studies, such as @spenceDisplayingProportionsPercentages1991 Each chart was printed on a large 22×28 inch card and shown one at a time in a randomized sequence to reduce bias.
@@
 
-Very recently, more works have been discovered on this debate that catch researchers' eyes on the topic.
+Very recently, more works have been discovered on this debate that caught researchers' eyes on the topic.
 @skau2016arcs investigated how common distortions in pie charts affect judgment accuracy.
@@
 
+@eellsRelativeMeritsCircles1926 study offered significant preliminary insights on the perception of proportions through pie and bar charts; yet, it possesses multiple drawbacks.
+While Von Hun has already identified several limits in his paper, further ones warrant discussion.
+The sample comprised exclusively psychology students from one institution, hence introducing potential sampling bias and constraining generalisability.
+Secondly, the design failed to mitigate the sequence of chart types – participants were exposed to pie charts initially, followed by bar charts three days later, potentially introducing learning or memory effects that could skew the comparison in favour of bar charts.
+Third, although participants reported their estimation tactics post hoc (e.g., angle, area), no measures were taken during the task (such as eye tracking or time consumed per chart), which reduces the capacity to analyse the underlying cognitive mechanisms.
+Ultimately, the study exclusively assessed whole-number estimates, perhaps obscuring nuanced variations in accuracy or bias among chart styles.
+
 @spenceDisplayingProportionsPercentages1991 demonstrated that pie charts can outperform bar charts under certain conditions when displaying proportions.
@@
 $C + D$).
-\svp{Were participants estimating ratios, or just judging larger/smaller?} Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones.
+\svp{Were participants estimating ratios, or just judging larger/smaller?}.
+\mt{Participants were not assessing ratios; they were rendering binary judgements regarding which component (or combination of components) was greater.
+The experiment employed a two-alternative forced-choice (2AFC) design, in which participants identified the greater of two possibilities (e.g., $A$ vs.
+$B$, or $A + B$ vs.
+$C + D$).
+Instead of conducting numerical estimation or ratio calculation, they compare relative magnitude.}
+Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones.
 Bar charts performed slightly better than pie charts for simple comparisons under time pressure, but differences between chart types were negligible for complex tasks.
@@
 \svp{How many categories were used? It's possible that this is true for small tables but not as true for larger tables.}
+\mt{The experiment included two categorical levels: four and seven components.
+The limited quantity likely facilitated the lack of an ordering effect, since participants could readily compare values without reliance on positional signals.
+In larger tables with more components, ordering may significantly enhance the accuracy and efficiency of interpretation.}
 
@@
 The small and homogeneous sample (mainly students) limits the generalizability of the findings.
-Short exposure times (1.0--6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.
+Short exposure times (1.0--6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.\mt{Might delete this paragraph}
+
 \svp{Why do psycologists use short exposure times?
-Look into this -- there's a reason} The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance.
-Reducing exposure time to 1.0 second in Experiment 4 may have biased results toward pie charts by encouraging intuitive rather than analytical processing
+Look into this -- there's a reason} The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance.\mt{Psychologists employ brief exposure durations (often 1–6 seconds) to concentrate on rapid perceptual reactions rather than advanced cognitive processing.
+This ensures that participants depend on visual perception rather than memory or intentional comparison.
+Foundational research, shown by @clevelandGraphicalPerceptionTheory1984, employed brief display durations to elicit direct assessments of visual attributes.}
 
@@
 
-# Field of comparison
-
 # Result
@@
 When people are asked to simply compare two values (A vs B), bar charts consistently come out on top---they're faster and more accurate, and participants tend to prefer them for that reason (@spenceDisplayingProportionsPercentages1991; @simkin1987information; @hill2025bar).
-When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (Cleveland & McGill, 1984; Simkin & Hastie, 1987).
+When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (@clevelandGraphicalPerceptionTheory1984; @simkin1987information).
 \svp{Is this still true when A and B aren't adjacent in e.g.
@@
 For more complex comparisons---like $A$ vs $B+C$ or $A+B$ vs $C+D$, the results are more mixed.
-Bar charts have a slight edge when comparing a single value to a sum ($A$ vs $B+C$), but when comparing two grouped values ($A+B$ vs $C+D$), pie charts actually perform better (Spence & Lewandowsky, 1991).
-This might be because it's easier to visually group adjacent slices than non-aligned bars (Kosara & Skau, 2016).
+Bar charts have a slight edge when comparing a single value to a sum ($A$ vs $B+C$), but when comparing two grouped values ($A+B$ vs $C+D$), pie charts actually perform better @spenceDisplayingProportionsPercentages1991.
+This might be because it's easier to visually group adjacent slices than non-aligned bars (@skau2016arcs).
 
-When it comes to ranking multiple parts, like identifying the top three largest values, bar charts win again, as their alignment allows for quicker visual scanning (Hill, 2025; Kosara & Skau, 2016).
-And in magnitude estimation tasks (e.g., "If A is 100, what is B?"), both pie and donut charts tend to outperform bars, though people are more likely to misjudge values slightly in those formats (Kosara & Skau, 2016; Hill, 2025).
+When it comes to ranking multiple parts, like identifying the top three largest values, bar charts win again, as their alignment allows for quicker visual scanning (@hill2025bar; @skau2016arcs).
+And in magnitude estimation tasks (e.g., "If A is 100, what is B?"), pie charts tend to outperform bars, though people are more likely to misjudge values slightly in those formats (@skau2016arcs; @hill2025bar).
 
@@
 
-# Comparison table
+## Comparison table
 
@@
 
-## Pie Chart
-
-```{r}
-# | echo: false
-library(ggplot2)
-library(dplyr)
-library(patchwork)
-
-# Helper function for pie chart
-create_pie <- function(data, fill_colors, title, bold_groups = NULL) {
-  data <- data %>%
-    mutate(
-      fraction = value / sum(value),
-      ymax = cumsum(fraction),
-      ymin = lag(ymax, default = 0),
-      label_pos = (ymin + ymax) / 2,
-      border_size = ifelse(label %in% bold_groups, 3, 0.8)
-    )
-
-  ggplot(data) +
-    geom_rect(aes(ymin = ymin, ymax = ymax, xmin = 0.5, xmax = 1.5,
-                  fill = label, size = border_size),
-              color = "black", show.legend = FALSE) +
-    geom_text(aes(x = 1, y = label_pos, label = label), size = 6) +
-    scale_fill_manual(values = fill_colors) +
-    scale_size_identity() +
-    coord_polar(theta = "y") +
-    labs(title = title) +
-    theme_void() +
-    theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))
-}
-
-# Pie 1: A vs B
-data1 <- tibble(label = c("B", "A"), value = c(0.35, 0.65))
-fill1 <- c("A" = "#f4a582", "B" = "#92c5de")
-p1 <- create_pie(data1, fill1, "A vs B")
-
-# Pie 2: (A+B) vs C
-data2 <- tibble(label = c("C", "B", "A"), value = c(0.25, 0.25, 0.5))
-fill2 <- c("A" = "white", "B" = "#ff00ff", "C" = "#00ffff")
-p2 <- create_pie(data2, fill2, "(A+B) vs C", bold_groups = c("A", "B"))
-
-# Pie 3: (A+B) vs (C+D)
-data3 <- tibble(label = c("C", "B", "A", "D"), value = c(0.25, 0.20, 0.35, 0.20))
-fill3 <- c("A" = "white", "B" = "white", "C" = "#00ffff", "D" = "pink")
-p3 <- create_pie(data3, fill3, "(A+B) vs (C+D)", bold_groups = c("A", "B"))
-
-# Combine with patchwork
-(p1 | p2 | p3)
-
-```
-
 ## Summary: When to Use Each Chart Type
 
-| **Use**              | **When**                                                                      
-|----------------------|-------------------------------------------------------------------------------|
+| **Use**              | **When**                                                              |
+|----------------------|-----------------------------------------------------------------------|
 | **Bar charts**       | Comparing individual parts (A vs B), ranking, or scanning many values |
@@
 # Conclusion
+
 As of now, no one has determined a definitive winner between pie charts and bar charts.
@@
 
- Individuals' perceptions of charts may alter between devices—mobile phones, tablets, or desktop monitors—due to variations in screen size, resolution, and interaction methods.
+Individuals' perceptions of charts may alter between devices---mobile phones, tablets, or desktop monitors---due to variations in screen size, resolution, and interaction methods.
 Furthermore, enhancing the reliability of perceptual measurements is crucial.

### 2025-06-10 sdss-abstract/index.qmd
WIP: save local changes before rebase

@@
 @eellsRelativeMeritsCircles1926 also quotes an excerpt from an introductory statistics textbook [@secristIntroductionStatisticalMethods1925], which argues that pie charts require more cognitive effort to use accurately than a comparable stacked bar chart, calling pie diagrams "clumsy and defective".
-These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use 
-<!-- [@ichartsmarketingdepartment4ProblemsPie2017;@holtzIssuePieChart2018] -->
-and a variety of empirical studies identify problems with circular data representations 
-<!-- [@hofmannGraphicalTestsPower2012] -->
-.
+These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use <!-- [@ichartsmarketingdepartment4ProblemsPie2017;@holtzIssuePieChart2018] --> and a variety of empirical studies identify problems with circular data representations <!-- [@hofmannGraphicalTestsPower2012] --> .
 @wickhamGraphicalCriticismHistorical2013 previously noted the prescience of design guidelines prior to empirical studies validating their recommendations across different types of visualizations.
@@
 <!-- Introspection is an incredibly powerful tool when applied to the design of statistical charts and can lead to design guidelines that precede empirical tests but match their results.   -->
+
 <!-- However, empirical tests are also significant: Tufte's data-ink ratio is oft repeated, but empirical studies show that the relationship between minimalism and chart perception is much more complicated [@gillanMinimalismSyntaxGraphs1994a;@tufteVisualDisplayQuantitative2001;@parsonsDataVisualizationPractitioners2020].
@@
 ## The Dawn of Empirical Graphics
+
 Interestingly, the question of pies and bars motivated some of the first empirical evaluations of charts, published in the Journal of the American Statistical Association: @eellsRelativeMeritsCircles1926, @vonhuhnFurtherStudiesGraphic1927, and @croxtonBarChartsCircle1927.
@@
 @vonhuhnFurtherStudiesGraphic1927's paper provides several examples which highlight concerns with the Eells study, and argues that while pie charts may be better in certain simple situations, bar charts are much more extensible and facilitate more advanced comparisons, but does not include any experimental studies.
-<!-- Not much has changed today - experimental work is often criticized by those who provide toy examples but no empirical data due to compromises necessary to ensure experimental control.  -->
-Croxton's early experimental data (attached to @vonhuhnFurtherStudiesGraphic1927) had a larger number of participants than @eellsRelativeMeritsCircles1926, but only tested two-category charts with only two proportions: .25 and .4.
+<!-- Not much has changed today - experimental work is often criticized by those who provide toy examples but no empirical data due to compromises necessary to ensure experimental control.
+--> Croxton's early experimental data (attached to @vonhuhnFurtherStudiesGraphic1927) had a larger number of participants than @eellsRelativeMeritsCircles1926, but only tested two-category charts with only two proportions: .25 and .4.
 In addition, Croxton's study elicited estimates from participants as $A/B$ comparisons, where $A + B = 1$, while Eells elicited estimates for $A$ and $B$ separately.
 <!-- It seems that prior to @eellsRelativeMeritsCircles1926, some researchers were empirically testing charts, but these experiments were not necessarily published for broader consumption; when the conversation arose in JASA, Croxton had data from years prior ready to provide additional insight.
--->
-This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.
+--> This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.
 
@@
 ## Modern Experimental Graphics
+
 Much later, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.

### 2025-06-17 presentation.qmd
Added readme and presentation files

---
title: "100 years of Pies vs. Bars"
author: " Maksuda Aktar Toma"
institute: "University of Nebraska Lincoln"
date: today
format:
  beamer:
    theme: metropolis
    colortheme: beaver
    title-slide-attributes: "align=center"
    pdf-engine: xelatex
    mainfont: Times New Roman
---

# Content

- Introduction
- Motivation
- The Dawn of Empirical Graphics
- Modern Experimental Graphics
- Historical Experimental design
- Result
- Conclusion

# Introduction: Pie vs. Bar — A Century-Long Debate

- Pie and bar charts date back to Playfair (early 1800s).
- Pie charts often criticized as less accurate, yet remain widely used.
- This review explores:
  - Historical experiments and design
  - Task performance comparisons
  - Practical usefulness of each chart type

# Motivation: Why Revisit Pie vs. Bar?

- Effective charts are essential for accurate communication of quantitative data.
- Bar charts consistently outperform pie charts in accuracy, speed, and clarity (based on early research).
- Yet, pie charts remain popular in media, business, and education.
  - This raises key questions:
  - Why do pie charts persist?
  - How do habits, aesthetics, or tradition influence chart selection?

![Pie vs. Bar](000023.png){width=80% height=160px}

# The Dawn of Empirical Graphics

- Early JASA studies (1926–1930s) by Eells, von Huhn, and Croxton tested pie vs. bar charts for accuracy and speed.
- Tasks varied: Eells used part estimation; Croxton used ratio comparisons like $A/B$, $A+B=1$
- Findings: No clear winner—results depended on task design; von Huhn and Croxton critiqued and refined earlier work.

![*Examples of comparison tasks used in graphical perception studies*](pie-slices.png)

# Modern Experimental Graphics

![*Visual encodings ordered by perceptual accuracy, as proposed by Cleveland and McGill*](cleveland-mcgill.jpg){width=80% height=200px}

\scriptsize
- The first six activities are listed according to Cleveland and McGill's accuracy tests. Assumed to be far less accurate than the others.

# Modern Experimental Graphics (Cont.)

- Spence & Lewandowsky (1991): Compared A vs. B+C using bar, pie, and table; measured accuracy and speed.
- Peterson (1954): Evaluated 8 chart types for part-whole accuracy; found format matters.
- Simkin & Hastie (1987): Bars better for comparisons; pies better for proportions.
- Skau & Kosara (2016): Pie distortions reduce accuracy; arc cues work best.
- Hill (2025): Bars best for ranking; pies/donuts strong in simple proportion tasks.

# Experimental Description

![*Summary of Experimental Studies on Chart Perception*](picture3.png){height=280px}

# Accuracy of Charts

![*Accuracy of Bar Chart By Task Type*](000013.png){width=80% height=200px}

\scriptsize
- Filled dots indicate studies that reported standard errors; hollow dots did not
- Hill (2024) showed the lowest estimation error, indicating the highest accuracy among all studies.
- Studies with Close bar alignment performed better than those with Far bar placement

# Accuracy of Charts (Cont.)

- Heer & Bostock (2010) showed the lowest error, indicating better accuracy with segmented bars.
- Older studies had higher error rates and often lacked standard error reporting (hollow points).

:::{.columns}
::: {.column width="50%"}

![*Accuracy of Bar Chart By Task Type*](000013.png)
:::
::: {.column width="50%"}

![*Accuracy of Bar Chart By Task Type*](00009.png)
:::
:::

# Accuracy of Charts (Cont.)

- Pie charts perform well for part-to-whole tasks, with low error (1–3%).
- Segment comparisons are harder, showing higher errors (4–6%).
- Recent studies show improved accuracy with better design.

:::{.columns}
::: {.column width="50%"}

![*Accuracy of Pie Chart By Task Type*](00008.png)
:::
::: {.column width="50%"}
![*Accuracy of Pie Chart By Task Type*](00007.png)
:::
:::

---

# Result

- Bar charts excel in comparison and ranking tasks — faster, more accurate, and user-preferred.
- Pie charts perform well in part-to-whole and grouped estimation tasks, aligning with intuitive reasoning.
- Task matters: Bar > Pie for $A$ vs $B$; Pie > Bar for $A / (A+B)$; mixed results for $A$ vs $B+C$, $A+B$ vs $C+D$

# Conclusion

:::{.callout-tip title="Conclusion"}
There is no universal winner.
The most effective chart type depends on the specific judgment or task.
:::

- Both chart types remain widely used across science, business, and social sciences.
- The effectiveness of a chart depends on:
  - the task (comparison, estimation, ranking),
  - perceptual features (angle, length, position),
  - and visual design elements ( adjacency, clarity).
- Pie charts are generally better for conveying part-to-whole relationships.
- Bar charts are superior for comparing magnitudes and identifying rankings.

# References

\scriptsize

Funkhouser, H. G. (1937). Historical development of the graphical representation of statistical data. Osiris, 3, 269–404. https://doi.org/10.1080/03697827.1937.10559475

Croxton, F. E., & Stein, H. (1932). Graphic comparisons by bars, squares, circles, and cubes. Journal of the American Statistical Association, 27(177), 54–60.

von Huhn, R. (1927). Further studies in the graphic use of circles and bars: A discussion of the Eells experiment. Journal of the American Statistical Association, 22(157), 31–39. https://doi.org/10.2307/2277346

Croxton, F. E., & Stryker, R. E. (1927). Bar charts versus circle diagrams. Journal of the American Statistical Association, 22(160), 473–482. https://doi.org/10.2307/2276829

Eells, W. C. (1926). The relative merits of circles and bars for representing component parts. Journal of the American Statistical Association, 21(154), 119–132. https://doi.org/10.2307/2277140

Cleveland, W. S., & McGill, R. (1984). Graphical perception: Theory, experimentation, and application to the development of graphical methods. Journal of the American Statistical Association, 79(387), 531–554. https://doi.org/10.1080/01621459.1984.10478080

Cleveland, W. S., & McGill, R. (1985). Graphical perception and graphical methods for analyzing scientific data. Science, 229(4716), 828–833. https://doi.org/10.1126/science.229.4716.828

# References

\scriptsize

Cleveland, W. S., & McGill, R. (1986). An experiment in graphical perception. International Journal of Man-Machine Studies, 25(5), 491–500. https://doi.org/10.1016/S0020-7373(86)80019-0

Simkin, D., & Hastie, R. (1987). An information-processing analysis of graphs and tables. Cognitive Psychology, 19(4), 530–544. https://doi.org/10.2307/2289447

Holtz, Y. (2018, August). The issue with pie chart. Data To Viz. https://www.data-to-viz.com/caveat/www.data-to-viz.com/caveat/pie.html

iCharts Marketing Department. (2017, November). 4 problems with pie charts (and what to do instead). iCharts. https://web.archive.org/web/20180223231734/https://icharts.net/blog/data-visualization-guides/4-problems-pie-charts-instead

Skau, D., & Kosara, R. (2016). Arcs, angles, or areas: Individual data encodings in pie charts. Computer Graphics Forum, 35(3), 121–130. https://doi.org/10.1111/cgf.12888

Heer, J., & Bostock, M. (2010). Crowdsourcing graphical perception: Using Mechanical Turk to assess visualization design. In Proceedings of the ACM CHI Conference on Human Factors in Computing Systems (pp.203–212). https://doi.org/10.1145/1753326.1753357

Hill, N. D. (2025). Bar, pie or donut? Exploring chart effectiveness for part-to-whole communication. Information Visualization, 24(1), 3–20. https://doi.org/10.1177/14738716241259432

---

# Thank You

::: {style="text-align: center;"}

\begin{center}
\LARGE \textbf{Questions?}
\end{center}

:::

### 2025-06-28 Draft.qmd
added SS and edited until modern graphics

@@
 The debate between pie charts and bar charts has been ongoing for over a century.
-Numerous studies with distinct methodologies and experimental designs have attempted to resolve this issue, but no single consensus has emerged.
-In this paper, we will provide a comprehensive summary of the existing literature, analyze the strengths and limitations of both chart types, and attempt to identify a unified solution based on empirical evidence and theoretical insights.
+<!-- Ok, the debate is not between the charts -- it's between either proponents of either chart (e.g. the people promoting them) or about the use of each chart.  -->
+Numerous studies with distinct methodologies and experimental designs have attempted to 
+determine which chart is more effective, but no single consensus has emerged, even after 100 years of experimentation.
+In this paper, we assess the existing literature and attempt to reconcile experiments with diverging results.
 
@@
 
-William Playfair introduced the pie and bar charts in the early 1800s, and by the turn of the next century, strong opinions about the effectiveness of alternative displays of the same data had been publicized in textbooks and manuals.
+Although William Playfair is frequently acknowledged with popularizing pie and bar charts in the early 1800s, the earliest documented use of bar charts can be traced to the 14th century, when Nicole Oresme employed bar-like representations in The Latitude of Forms.
+Consequently, Playfair's contribution is mostly in the formalization and distribution of these charts rather than their creation.
+By the turn of the 20th century, strong opinions about the effectiveness of different chart types—especially pies versus bars—were already appearing in textbooks and manuals.
+
 For instance, @eellsRelativeMeritsCircles1926 quotes two contemporary resources that discuss pie charts as "not a desirable form of presentation" [@brintonGraphicMethodsPresenting1914] and, more strongly, "an insult to a man's intelligence" [@karsten1923charts].
@@
 @eellsRelativeMeritsCircles1926 also quotes an excerpt from an introductory statistics textbook [@secristIntroductionStatisticalMethods1925], which argues that pie charts require more cognitive effort to use accurately than a comparable stacked bar chart, calling pie diagrams "clumsy and defective".
-These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use
+These insights mirror today's status quo: pie charts are stubbornly common, even though many design guidelines recommend against their use.
+Building on these early critiques, @wickhamGraphicalCriticismHistorical2013 observed that design guidelines anticipated later empirical findings across many visualization types.
 
 @wickhamGraphicalCriticismHistorical2013 previously noted the prescience of design guidelines prior to empirical studies validating their recommendations across different types of visualizations.
+[Ok, can you connect this better to the previous paragraph somehow?]{.svp}
 
@@
 The phrasing of direct estimation questions is of primary importance: a simple mathematical transformation of the estimated quantity can have an outsized effect on the results.
+[One important insight from both historical and modern work is that the phrasing of the estimation task itself can strongly influence performance.
+For instance, @croxtonBarChartsCircle1927 asked participants to "estimate what percentage each sector represented," whereas @eellsRelativeMeritsCircles1926 phrased his question more generally, asking participants to “judge the relative size of the sectors.” (See Figure 1 for a visual comparison.) These subtle differences in wording can lead to significantly different error rates—underscoring how critical question design is in graphical perception research.]{.mt}
+
+::: {.center}
+
+\begin{figure}[ht]
+\centering
+\begin{minipage}{0.32\textwidth}
+  \includegraphics[width=\linewidth]{eells-1.png}
+\end{minipage}
+\begin{minipage}{0.32\textwidth}
+  \includegraphics[width=\linewidth]{eells-2.png}
+\end{minipage}
+\begin{minipage}{0.32\textwidth}
+  \includegraphics[width=\linewidth]{crox.png}
+\end{minipage}
+\caption{Sample stimuli from Eells (1926), where participants were asked to estimate proportions directly from unlabeled pie charts.Diagrams from Croxton (1927), using labeled segments to compare accuracy in circle and bar comparisons*}
+\end{figure}
+
+:::
+
+[You might want to demonstrate this with an example, perhaps showing the phrasing of Croxton's study vs. Eells? You could even include screenshots from the papers.]{.svp}
 
@@
 <!-- It seems that prior to @eellsRelativeMeritsCircles1926, some researchers were empirically testing charts, but these experiments were not necessarily published for broader consumption; when the conversation arose in JASA, Croxton had data from years prior ready to provide additional insight.
---> This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.
+-->
+This multi-year back-and-forth conversation is interesting in part because it pinpoints an early example of experimental statistical graphics as a discipline of interest to statisticians for both personal investigation and as an audience for the results.
+[It is both an example of early graphical testing and a cautionary tale, as proponents of different viewpoints were primarily talking past each other.
+While the riposte in these early papers and the strong opinions in contemporary design guidelines are entertaining to read, they do not generally support informed choices about what chart to use in practice.]{.svp}
 
@@
 Experimental graphics studies have largely gone unreported (if they were conducted) between 1932 and the 1980s, particularly in statistical publications, though the issue of pies and bars occasionally pops up during this period in other disciplines.
+[However, guidance to favor bar charts over pie charts still shows up in statistical literature [@haemerPresentationProblemsSuiting1949;@haemerPseudoThirdDimension1951;@fienbergGraphicalMethodsStatistics1979;@nisselson1978graphic;@wickhamGraphicalCriticismHistorical2013].
+It almost seems like approximately once a generation, statisticians think about graphics, decry the state of the field, and call for a better set of empirical guidelines.]{.svp}
 
 # Modern Experimental Graphics
-
-Much later, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.
+Approximately 55 years after the original sequence of experiments, @clevelandGraphicalPerceptionTheory1984, a statistician and a psychologist by training, respectively, conducted experiments on simple graphical elements.
 These experiments were presented as supporting a hierarchy of graphical perception that evolved over the course of the study sequence.
@@
 @spenceDisplayingProportionsPercentages1991 (both psychologists) used forced-choice questions of the form "which is bigger, A or B+C" to investigate the accuracy of pie and bar charts for making part-to-whole comparisons.
-The later studies show more of an interdisciplinary conversation around data displays, which have become a general tool not limited to statisticians.
+The later studies show that the 1980-1990s experiments were more of an interdisciplinary conversation around data displays, [reflecting the fact that statistical graphics had]{.svp} become a [more]{.svp} general [scientific]{.svp} tool [and were no longer]{.svp} limited to statisticians.
 As a result, the literature becomes fragmented; studies are conducted across different disciplines, with different methods, elicitation questions, protocols, and, ultimately, different conclusions.
 
-![Different elicitation methods used to assess accuracy of pie chart estimates.](pie-type-example.png)
+![Different methods used to assess accuracy of pie chart estimates.
+In addition to the different ways to divide up a pie chart's comparisons, there are also different numerical quantities which can be requested: the ratio of C to A + B and the ratio of C to the whole (A + B + C).
+These different elicitation methods impact the accuracy of participant responses for each type of chart.](pie-slices.png){fig-alt="Three pie charts, divided up in A/B, (A/B)/C, and A/(B/C)/D slices.
+The first compares A to B, the second compares A+B to C, and the third compares A+D to B+C.
+Slices that are grouped (A/B) and (A/D), (B/C) are colored similarly."}
 
@@
 The sample comprised exclusively psychology students from one institution, hence introducing potential sampling bias and constraining generalisability.
-Secondly, the design failed to mitigate the sequence of chart types – participants were exposed to pie charts initially, followed by bar charts three days later, potentially introducing learning or memory effects that could skew the comparison in favour of bar charts.
+Secondly, the design failed to mitigate the sequence of chart types – participants were exposed to pie charts initially, followed by bar charts three days later, potentially introducing learning or memory effects that could skew the comparison in favor of bar charts.
 Third, although participants reported their estimation tactics post hoc (e.g., angle, area), no measures were taken during the task (such as eye tracking or time consumed per chart), which reduces the capacity to analyse the underlying cognitive mechanisms.
@@
 $C + D$).
+
 \svp{Were participants estimating ratios, or just judging larger/smaller?}.
@@
 
+[Were participants estimating ratios, or just judging larger/smaller?]{.svp} Accuracy improved with longer display times, reaching near 100% for simple tasks and around 90% for complex ones.
+Bar charts performed slightly better than pie charts for simple comparisons under time pressure, but differences between chart types were negligible for complex tasks.
+
 The second experiment yielded similar results but employed a different approach.
@@
 However, the lack of an ordering effect challenges the common belief that ordered components improve perception.
+
 \svp{How many categories were used? It's possible that this is true for small tables but not as true for larger tables.}
@@
 
+[How many categories were used? It's possible that this is true for small tables but not as true for larger tables.]{.svp}
+
 The third experiment examined how task complexity and display type (pie chart, bar chart, and table) affect performance using a within-subject design, where each subject performed all three task types.
 Unlike Experiment 1, which used a between-subjects design, and Experiment 2, which introduced tables and fixed processing time, Experiment 3 retained pie charts, bar charts, and tables while excluding the horizontal divided bar chart.
-Processing time was fixed at 3.0 seconds \svp{You can fix stimulus time but not processing time -- processing takes as long as it takes}, and two levels of components (4 and 7) were tested.
+Processing time was fixed at 3.0 seconds [You can fix stimulus time but not processing time -- processing takes as long as it takes]{.svp}, and two levels of components (4 and 7) were tested.
 Accuracy declined with task complexity, and while pie and bar charts showed similar performance for simpler tasks, pie charts outperformed the others for complex tasks, likely due to easier component identification and combination.
@@
 The small and homogeneous sample (mainly students) limits the generalizability of the findings.
+
 Short exposure times (1.0--6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.\mt{Might delete this paragraph}
@@
 
+Short exposure times (1.0--6.0 seconds) may not reflect real-world scenarios where viewers have more time to process information.
+[Why do psycologists use short exposure times?
+Look into this -- there's a reason]{.svp} The artificial setting (controlled computer-based tests) fails to capture real-world variability, such as distractions and viewing distance.
+Reducing exposure time to 1.0 second in Experiment 4 may have biased results toward pie charts by encouraging intuitive rather than analytical processing
+
 # Measurement and Preference
@@
 When people are asked to simply compare two values (A vs B), bar charts consistently come out on top---they're faster and more accurate, and participants tend to prefer them for that reason (@spenceDisplayingProportionsPercentages1991; @simkin1987information; @hill2025bar).
+
 When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (@clevelandGraphicalPerceptionTheory1984; @simkin1987information).
@@
 Further studies are needed to assess whether the part-whole encoding remains effective when the target slices are visually separated—especially in 3- or 4-component pie charts where holistic comparison becomes less direct}
+
+When the task shifts to estimating what proportion one part is of the whole ($A / (A+B)$), pie charts do better, likely because they align with how we naturally think about part-whole relationships (Cleveland & McGill, 1984; Simkin & Hastie, 1987).
+[Is this still true when A and B aren't adjacent in e.g.
+a 3-4 component chart?]{.svp} [While pie charts have been shown to support accurate part-to-whole estimations in two-part visualizations (e.g., A and B as adjacent slices), it is unclear whether this advantage persists in more complex charts.
+When pie charts include three or more segments, and the parts being compared (e.g., A and B) are not adjacent, the viewer must shift focus across non-contiguous regions of the circle.
+This added perceptual effort may diminish the intuitive advantage typically associated with pie charts in simpler tasks.
+Further studies are needed to assess whether the part-whole encoding remains effective when the target slices are visually separated—especially in 3- or 4-component pie charts where holistic comparison becomes less direct]{.mt}
 
